### 2020-08-10 text/ecography/round_1/review/reply_to_reviewers.Rmd
migrate gitlab repo to github

---
output:
  bookdown::pdf_document2:
    keep_tex: no
    latex_engine: xelatex
    # float is needed for floating figures
    extra_dependencies: ["fontspec", "sectsty", "float"]
    toc: false
    fig_caption: yes
    number_sections: true
header-includes:
  # adjust figure captions
  - \usepackage{caption}
  - \captionsetup{font={footnotesize,bf,singlespacing}}

title: "Point-by-point reply to reviewers' comments"
bibliography: ../../round_2/transect_refs.bib
csl: ../../round_2/ecography.csl
# biblio-style: "natbib"
link-citations: true

---

```{r setup, include=FALSE}
library("knitr")
library("kableExtra")
library("tidyverse")
library("here")
# make sure that figures are allowed to fload (hold the assigned position and are not always put on top of a page)
knitr::opts_chunk$set(fig.pos = 'H')
```

# Reply to subject editor Bethany Bradley

Dear Dr Bradley,

thank you very much for handling our manuscript.
We are pleased to resubmit a revised version of our manuscript now entitled  *Monitoring and predictive mapping of floristic biodiversity along a climatic gradient in ENSO's terrestrial core region, NW Peru*.
We are also very thankful for the constructive comments of both reviewers which helped to improve substantially our manuscript.
Please find below a point-by-point reply to all concerns raised by the reviewers.
In accordance with your and the reviewers' suggestions, we have:

- formulated clear hypotheses and adjusted accordingly the framing of the manuscript.
- explained our methods in much more detail.
To make our study truly reproducible, we will publish our study upon acceptance as a research compendium.
This means that we will make our code and text publicly available via [github](https://github.com).
The data will be published on DRYAD or Zenodo.
And a docker container will make available the software used during the analyses thereby ensuring that our code will run on any operating system and regardless of software updates.
- embedded our study even better in the ongoing scientific debate while focussing on the key points of our study.
- highlighted more prominently the novelty of our study in the introduction and discussion section.
- written out El Niño/La Niña in full rather than abbreviating.

Therefore, we hope that our manuscript is now suitable for publication.

Yours sincerely,

Dr Jannes Muenchow

# Reply to reviewer #1

1. **The authors assessed the effect of water availability on species diversity across a gradient in the Northwest Peru in different years in which the effect of climatic events (ENSO) have influence on plant diversity. They used 50 (0.09ha) plots distributed in different water availability conditions along a transect and measured plant identity and plant coverage of vascular plants in each one. In addition, they used an irrigation-experiment to assess under controlled conditions the effects of water availability and their interaction with nutrient enrichment on plant diversity accumulation. This integrative approach which combines macroecological analyses and experimental tests of the potential drivers is very interesting and rare in ecological studies. However, the methods and results are not clear. Particularly the description of the plot data, the modeling approaches, and the display of results. Moreover, there is a mismatch in the response variable between the observational analyses and the experiments. Finally, in the introduction, the hypotheses do not have clear predictions, and the discussion focusses on topics that are not presented as the focus of the manuscript in the introduction. In addition, the text should be reviewed to improve the quality of the writing, there are some sentences that are not clear. In my opinion, this study has a lot of potential, but these issues need to be addressed.**

First of all, thank you very much for your detailed feedback which helped to improve substantially the manuscript.
In accordance with your suggestions, we have rewritten the methods and results sections.
Additionally, we have described in much more detail the plot data and the modeling approaches.
We also show that in fact there is no mismatch between the response variables (please refer to the reply to comment #12).
We are sorry that this was not clear from the beginning.
We also have rewritten the research questions as hypotheses with clear predictions.
Additionally, we absolutely agree with you that parts of the discussion in the first draft of the manuscript were out of scope.
Thanks for pointing this out, we have deleted or largely rewritten the corresponding sections and the Conclusions.
Finally, we have improved the writing.

2. **Line 48: The introduction should include a bit about the expectation of species diversity response to climate conditions. Also, the term “ecosystem diversity” should be clarified.**

We have rewritten our research questions as follows:

>>> Specifically, we investigated the:
>>>
(1) change in the different levels of biodiversity, namely alpha, beta, gamma and ecosystem diversity. 
The latter was represented by ordination scores and additionally visualized by means of predictive mapping.
Overall, we expected a boost in biodiversity as soon as water was no longer a limiting factor.
Regarding the spatial distribution of the floristic composition we expected a more pronounced development of vegetation assemblages along the climatic gradient with wetter conditions.
>>>
(2) effects of soil characteristics and topography in shaping floristic composition and vegetation cover (biomass production). 
We expected that the influence of topographic and edaphic variables might gain in influence in wetter years.
>>>
(3) We complemented the plot survey at the landscape scale with an irrigation-fertilization experiment, thereby considering the socio-economic perspective of the ENSO variability. 
Specifically, we quantified the effect of the water-nutrient interaction on biomass development under different rainfall scenarios.
We expected that the water input of the rainfall scenario representing a Super-Niño event in combination with the fertilizer would be most conducive to biomass and biodiversity development.

Additionally, we now specifically say that we represented ecosystem diversity by ordination scores.

3. **Line 50: The introduction is focused on floristic composition and the second question also includes information on biomass production. Here, there are many studies that showed the effect of ENSO and extreme climatic events in aboveground biomass production (e.g. Phillips et al. 2009). Based on these previous studies, the authors could include a bit more of information on the response of biomass production and their expectations for the study system.**

Regarding our expectations, please refer to the reply to the previous comment.
Thanks for pointing us to relevant literature.
We have added following sentence to the introduction:

>>>In any case, ENSO certainly drives a large amount of the interannual variability of woody plant growth across the tropics [@rifai_enso_2018] though the specific response to El Niño may vary from region to region [@phillips_changes_2009].

<!--
- [@phillips_changes_2009] -> mixed response of Amazon forest to EN induced droughts.
However, climate change affects Amazon moist forests, there is a risk that they might be converted from carbon sinks to sources.
- cite also Salazar et al. (accepted) since they also show that there is a risk that climate change might convert TDF from sink to sources.
- [@rifai_enso_2018]: ENSO Drives interannual variation of forest woody growth across the tropics 
Overall, there is high correlation (r = -0.75) between the annual anomaly of tropical forest woody growth and the annual mean of the El Niño 3.4 index, driven mainly by strong correlations with anomalies of soil water deficit, vapour pressure deficit and shortwave radiation.

Climate change has the potential to convert both tropical moist as well as dry forests from carbon sinks to sources.
-->

4. **Line 86: The description of vegetation sampling should be improved. It is not clear how the proportion of area occupied by each species is measured. Is it based on basal area o crown dimensions?**

We have clarified as follows:

>>>Plant cover refers to the relative projected area covered by a species [@damgaard_estimating_2014].

5. **Line 100: Are there more than one plot within the same NDVI pixel?**

There is not one pair of plots which are located in the same pixel (see Table \@ref(tab:ndvi)).

```{r ndvi, echo=FALSE, eval=TRUE, message=FALSE}
# attach packages
library("knitr")
library("dplyr")
library("raster")
library("reshape2")
# attach raster stack
ndvi = readRDS(here("images/00_ndvi_rasters.rds"))
# attach env_data
env_data = readRDS(here("images/11_env_data.rds"))
env_data$year = gsub(".*_", "", env_data$pnr)

# just keep unique plots (could have also used the id-column)
data = env_data[!duplicated(paste(env_data$lon, env_data$lat)),
                  c("lon", "lat")]
# extract values from NDVI rasters to plot locations
vals = raster::extract(ndvi[[c("ndvi_11", "ndvi_12", "ndvi_16", "ndvi_17")]], 
                       data[, c("lon", "lat")])
# count different NDVI values for each sampled year
tab = melt(vals) %>%
  group_by(Var2) %>%
  summarize(no_plots = n(), distinct_ndvi_values = n_distinct(value))
tab = rename(tab, year = Var2) %>%
  mutate(year = gsub("ndvi_", 20, year))
kable(tab, "latex", booktabs = TRUE, escape = TRUE,
      caption = "Showing that each plot has a distinct NDVI value.")
```

6. **Line 101: Worldclim data are static and describe the general pattern of variation, but I´m not sure if this database is appropriate for this study. Can you explore the inclusion on TRMM data? These data have a coarse spatial resolution but a fine temporal resolution and could improve your analysis.**

```{r gam-fig, echo=FALSE, message=FALSE, results = "asis", }
library("mgcv")
require("itsadug")

# attach response (DCA scores)
data = readRDS(here("images/15_model_input_data.rds"))
data$id = as.numeric(data$id)
data$year = as.factor(data$year)
fo = as.formula(DCA1 ~ s(ndvi, by = year, k = 3) + year)
gam_1 = gam(fo, data = data)
# gamtabs(summary(gam_1))
# plot_smooth(gam_1, view = "ndvi", plot_all = "year", ylim = c(-2, 3))
```

We appreciate the recommendation.
However, the spatial resolution of TRMM data is too low for our data, i.e., one pixel would cover several plots.
Please note also that we already have used the MODIS data for the vegetation period of each year, i.e., we have used MODIS data from 2011, 2012, 2016 and 2017. 
This also means that the NDVI for each year was computed as the mean for all NDVI MODIS images available for the vegetation period (March to May). 
We say so now more clearly in the text as follows:

>>> We downloaded the normalized difference vegetation index (NDVI) raster imagery from the Moderate Resolution Imaging Spectrometer (MODIS) available for the vegetation period (March, April) for each studied year (2011, 2012, 2016, 2017) at a 250 × 250 m^2^ resolution for our study area.
Since three NDVI scenes were available for each vegetation period, we computed the mean NDVI for each vegetation period.

Please note also that our GAM model clearly reveals a relationship between the interaction of the MODIS NDVI and year and the response variable (adjusted R squared: `r round(summary(gam_1)$r.sq, 2)`) proving that the MODIS data is appropriate for the chosen modeling approach.

7. **Line 105: This is an interesting integrative framework, but it is not clear to me what the source of species in each one plot is after the experiment was establishment. Where they planted as seeds? Is the result of immigration from nearby forest remnants or are artificially included in each plot?**

Thank you very much for the compliment!
The experiment was established on the campus of the University of Piura.
The campus comprises large undisturbed areas mirroring the typical vegetation formation, open xeric shrublands, typical of the region around Piura.
Hence, we only observed how the natural seed bank developed when exposed to artificial water/nutrient input.
Please refer also to the Supplementary Information for a detailed description of the experimental setup.

8. **Line 132: What is Londo-scale?**

Actually, we were referring to the decimal scale proposed by Londo. 
The decimal scale has finer intervals than the Braun-Blanquet scale (Table \@ref(tab:decimal-scale)).
In addition, the decimal scale does not use symbols like `r` and `+` which are hard to handle for software expecting numeric values such as ordination techniques.
We have added following sentence to the description of the vegetation sampling:

>>>Prior to the analysis, we transformed the plant cover values in accordance with the decimal scale by Londo [@londo_decimal_1976].

```{r decimal-scale, echo=FALSE}
tab = tibble(
  "coverage (in %)" = c("<1", "1-3", "3-5", "5-15", "15-25", "25-35", "35-45", "45-55", "55-65", "65-75", "75-85", "85-95", "95-100"),
  "londo_values" = c(0.1, 0.2, 0.4, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10))
cap = "Decimal scale as proposed by Londo (1976)."
kable(tab, "latex", booktabs = TRUE, escape = TRUE, 
      # align = "c",
      # col.names = linebreak(nms, align = "c"),
      caption = cap)
```

9. **Line 138: The modeling approach maybe inappropriate. The authors used a detrended canonical analysis to reduce dimensionality of the community matrix and then used a generalized additive model to compare the DCA axis with environmental data. There are other methods more appropriate to evaluate the effect of climate and other factors on species composition such as canonical correspondence analysis or redundancy analysis. Also, the generalized dissimilarity modeling allows to account for non-linear relationships among predictors and species composition variation.**

Apparently, we needed to rewrite our methods section to communicate more clearly how we have addressed methodologically our aims.
We have now made crystal-clear in the text that we were interested in spatially predicting the floristic gradient represented by the scores of the first DCA axis.
Kindly refer to the completely rewritten section "Modeling and predictive mapping" which now reads as follows:

>>>One aim of this study was to spatio-temporally predict ecosystem diversity (floristic composition) represented by ordination scores along the climatic gradient in the study area.
@muenchow_coupling_2013 have already shown that NDVI is an excellent predictor for vegetation assemblages in the study area.
Therefore, we modeled the scores of the first DCA axis representing the main observed floristic gradient in the study area with the help of a Generalized Additive Model (GAM) as a function of the interaction between NDVI and year:

>>>$$
\begin{aligned}
\text{Scores}_{i}\ \sim\ \mathrm{N}\left( \mu_{i},\ \sigma^{2} \right) \\
\mu_{i} = \alpha + f_{\text{year}_{i}}\left( \text{NDVI}_{i} \right) + \beta_{1}\text{year}_{i}
\end{aligned}
$$

>>>(Formula 1)
>>>
>>>where *i* refers to the *i*th observation, *$\alpha$* is the intercept, *$\beta$*~1~ is the estimated coefficient for year and *f* is a spline smoother for the NDVI:year interaction.
>>>
>>>The sole purpose of the model is the spatial prediction of the ordination scores, not statistical inference or the interpretation of the models' coefficients.
We will use variation partitioning later on to assess how the influence of environmental predictors has changed over the studied years (see also next section). 
Consequently, here we are only interested in a good predictive performance of the model to make sure that it is able to generalize, i.e., that it is able to predict reasonably well unseen data [@james_introduction_2013; @kuhn_applied_2013]. 
Only this can be the basis for a reliable map of the predicted ordination scores.
To make sure this is the case, we used spatial cross-validation.
Overall, cross-validation accounts for over-optimistic predictions on the training set by randomly splitting a data set into partitions used for training and testing [@brenning_spatial_2012;@lovelace_geocomputation_2019a]. 
Spatial cross-validation is an extension that measures a model's ability to spatially predict the response variable in the presence of spatial autocorrelation [@schratz_hyperparameter_2019]. 
We ran a 100-repeated 5-fold spatial cross-validation in which the partitioning is based on *k*-means clustering (*k* = 5; @russ_data_2010).
We also made sure that all annual observations of one plot were jointly placed in either the test or the training data set [@meyer_improving_2018]. 
We used the normalized root mean-square error (NRMSE) in percent as performance measure for the spatial cross-validation.
>>>
>>>$$
NRMSE = \frac{\sqrt{\frac{\sum_{i = 1}^{n}{(e_{i} - y_{i})²}}{n}}}{max(y) - \ min(y)}*100
>>>$$ 
>>>(Formula 2)

>>>where *n* is the number of observations, *e* corresponds to the residuals, and *y* refers to the observed values.

Since spatial prediction was the aim, we were not very much interested in the effect of the predictors (climate, NDVI) but only in a good predictive performance of the model to ensure that it is able to generalize [@james_introduction_2013].
But please note that the variation partioning uses the RDA approach, as suggested, to assess the impact of environmental predictor groups on the floristic composition across time (see also reply to your next comment).

Of course, we are also aware of canonical (constrained) ordination techniques [@legendre_numerical_2012]. 
Constrained ordination techniques model every single species as a function of environmental predictors, and then use the fitted values as input for an ordination method - either a PCA (then called RDA) or a CA (then called CCA).
One shortcoming of these techniques is that they usually explain less of the observed variance.
Another reason why we have not used constrained techniques for the predictive modeling was that we would have to use the same predictors (NDVI) twice - once as input for the CCA/RDA and then again for modeling the scores of the CCA/RDA.

<!--
One shortcoming of constrained techniques is that they exclusively use linear models, which might not be the most appropriate choice for each species encountered along a pronounced environmental gradient [@borcard_numerical_2018].
Secondly, in our case, species were recorded using the Londo scale which is a categorical variable. 
However, a linear model expects a response variable that follows a Gaussian distribution.
That alone would make the usage of a constrained ordination technique at least questionable.
Still, we could use one, and though we would not very much trust the modeled values of a canonical ordination technique for the reasons listed above, we could still use its scores as a response variable instead of the scores of a DCA.
But again, there are at least two reasons why we would refrain from such an approach:

1. Constrained techniques usually explain less of the observed variance of the original input matrix compared to their unconstrained counterparts since they only include in the ordination what the model was able to explain, the rest, i.e., the residuals, are simply ignored, which of course is not the case in the case of unconstrained techniques.
1. We would have to use the same predictors (NDVI and time) for the canonical ordination and the subsequent score modeling, i.e., we would e.g., model the CCA scores which themselves are the fitted values based on the same predictors. 
This smells very much like circular reasoning.
--> 
Finally, the generalized dissimilarity modeling approach is interesting, as it allows to model species turnover (beta diversity), and in fact we will give it a try in a future study.
However, in this study, we were interested in modeling the spatio-temporal change of the vegetation assemblages along a pronounced climatic gradient during four different ENSO years.

10. **Line 156: The variance partitioning approach used in not clear. If it is based on the GAM models employed previously, it is incorrect because they are based on the likelihood of the model (explained deviance) that is similar to R2 in an ordinary least square model, but not the same. In addition, the redundancy analysis approach (RDA) is more frequently used to assess the effect of environment on species composition. Also, the use of similar modeling methods previously employed allow comparability with other studies.**

We are sorry again that obviously we have not presented our methodological approach in sufficient detail.
First, the variation partitioning is not based on the GAM.
Instead, it is based, as suggested in your comment, on an RDA approach.
The variation partitioning computes with the help of RDA how the influence of predictor groups - in our case edaphic and topographic variables - has changed with different levels of water availability [@borcard_numerical_2018].
<!--
Second, you are right that RDA is used to assess the effect of environmental predictors on the species composition.
However, a RDA is not able to quantifiy the effects of different predictor groups - in our case edaphic and topographic variables - on species composition over time.
-->
We now made our reasoning now much clearer in the methods sections.
The updated text in the manuscript reads as follows:

>>>Variation partitioning quantifies the variation explained of the response, in our case the first two DCA axes, by one predictor group, in our case edaphic variables, while controlling for the effect of another predictor group [@borcard_numerical_2018; @peres-neto_variation_2006], in our case topographic variables.
Internally, variation partitioning computes a Redundancy Analysis (RDA) separately for each predictor group and for all predictor groups together, and subesequently quantifies how much each predictor group explains alone and jointly with the other predictor group of the variation of the response.
This is frequently displayed with the help of a Venn diagramm.
We ran the variation partitioning for each studied year (2011, 2012, 2016, 2017) to assess how edaphic and topographic variables influence vegetation composition under increasingly humid conditions.

Please note that upon acceptance we will make our analysis, data and software environment publicly available to make our research reproducible.
The variation paritioning will be available in the code repository under `code/analysis/17_varpart.R`.

11. **Line 173: The results of local species richness in Figure 2 compare among years and along the spatial gradient. But the authors do not describe in the Methods section the approach used to assess these differences? Are they using an ANOVA? Also, the modeling of local species richness maybe done in a similar way than the models employed here with DCA axes. Also, these models need to include spatial autocorrelation in their structure.**

Figure 2 is just an exploratory figure.
The line smoother was only added to aid visual interpretation.
Hence, we are not using an ANOVA here.
We are sorry for the confusion.
The caption now contains a sentence making this clear:

>>>To aid visual interpretation, loess smoothers were added.

And of course, we could have also modeled species richness, however, the aim of the study was to model the floristic gradient, i.e., the change of the floristic composition along the gradient.
Please see also our reply to comment #9 where we describe our modeling approach in more detail including how we have addressed spatial autocorrelation.

12. **Line 217: The results based on the irrigation experiment are focused on plant cover rather than plant species diversity that is the main goal of the study. These results are incongruent with the previous approach that is assessing the effect of a spatiotemporal water availability gradient on species diversity. The integrated approach is very interesting, but the response variable should be the same in the observational and experimental analyses to allow comparability. The authors can focus on plant coverage change across the plots is response to water availability. Also, if the goal is diversity, they should show results of irrigation and nutrient enrichment on the plant diversity in their experimental plots.**

In fact, there was no mismatch regarding the response variables for the predictive modeling and the irrigation-nutrient experiment. 
Sorry for not making this clear from the beginning.
We recorded the plant coverage of each species in the observational analyses and the experiments.
The decimal scale transformed plant coverage values (see also reply to comment #8) were the input for the Detrended Correspondence Analysis whose results were subsequently used as response variable for the predictive modeling approach.
In the case of the experiment, plant coverage values were simply plotted against time.
We hope we have made this clearer in the rewritten version.
Additionally, we have plotted alpha diversity against time for the experiment data (see Figure \@ref(fig:alpha-exp)).
<!-- JM: add figure to Appendix and reference it in the text --->
We have added the corresponding figure to the Supplementary Information.
Please note that alpha diversity was not very high since mostly a few dominant species (grasses) have covered the experimental plots.
This is also in correspondence with our observations from the field (see discussion section).

```{r alpha-exp, echo=FALSE, fig.cap="Number of species plotted against time during the duration of the experiment.", message=FALSE}
p_1 = readRDS(file = here("images/18_irrigation_spri.rds"))
print(p_1)
```

13. **Finally, some of these papers could be included in the introduction and discussion sections**

Thank you very much for taking the time to recommend relevant literature. 
We now cite @engelbrecht_drought_2007, @condit_species_2013, @alvarez-davila_forest_2017, @esquivel-muelbert_biogeographic_2017, @esquivel-muelbert_seasonal_2017 in our rewritten introduction and discussion sections.

# Reply to reviewer #2

1. **The manuscript addresses a very relevant question regarding our understanding of the dynamics of tropical seasonal ecosystems, especially those under strong influence of events like ENSO. The results are interesting, but I have not been able to convince myself that the approach taken by the authors is adequate. First, my main concern is that the four years of data, are not enough to reveal the relationship of ENSO with these coastal ecosystems. I agree that ENSO is a very complex phenomenon (as shown by the very heterogeneous response of precipitation in the four years of data collection, irrespective of what type of ENSO event happened) and as such it needs more intensive and long-term data. As shown by the authors, rainfall seems to be a good variable to study the response of plant communities in the area and for the time span used, but this general conclusion has been shown also by previous studies. In this regard, I would have also liked to see and discussed references addressing similar questions (e.g. the study by Elsa Cleland et al. Ecology 2013, in temperate grasslands).**

First of all, thank you very much for taking the time to improve our manuscript with your very helpful comments.
We agree that much more research is needed to deepen our understanding of the relationship between ENSO and the dry Peruvian coastal ecosystems.
Still, monitoring vegetation development across four different ENSO years along a pronounced climatic gradient is a rather unique research opportunity, which probably has not been done before.
Additionally, we would like to point out that one of our co-authors has revisited the study area each year since 1997.
Moreover, two of our Peruvian co-authors are living in Piura, one of them has even been born there.
This is to say that we are quite familiar with the dynamics of this ecosystem in response to ENSO events.
In our manuscript we have backed up this expert knowlegde by a data-driven approach.
Still, we plan to continue the revisiting of our plots within a new proposed research project, in which we will focus on how climate change affects tropical dry forests on various spatial scales.
This is of utmost importance especially when considering that climate change will affect the intensity and frequency of ENSO events.

Thank you for suggesting relevant literature.
The updated discussion pararaph now reads as follows:

>>>Beta diversity (expressed as the range of the first DCA axis), increased with wetter conditions (Fig. 2 & Table 1).
Naturally, higher species turnover was mainly driven by short-lived, less abundant species.
Concordantly, @cleland_sensitivity_2013 reports similar findings from temperate grasslands.

Please see also our reply to your next comment.

2. **Second, I would like to know what the authors think about a lagged response of the studied ecosystem to the ENSO conditions of the previous season (or seasons). Again, there are some interesting studies looking into this in temperate grasslands (e.g. Joan Dudney et al. J. Ecol. 2016, Brigitte Tenhumberg et al. Ecology 2018), and it is not improbable that similar mechanisms operate in the study area. I acknowledge that the data in the paper are not enough to incorporate this into the manuscript, but I would have welcomed its discussion.**

Thanks for pointing us to the time-lagged response of ecosystems.
Accordingly, we have added to the discussion following paragraph:

>>>Though we focussed on how current-year precipitation affects the vegetation composition, previous year rainfall might also have an effect [@tenhumberg_timelagged_2018].
For instance, @dudney_lagging_2017 observed an increase in grass and a decrease in forb abundance when rainfall was high one year earlier.
Similarly, @walter_plants_2011 suggest that grasses might have a 'drought-memory'.
Time-lagged responses can be also observed in woody species.
For example, tree ring growth of *Prosopis pallida* is highly correlated with rainfall of the previous year [@salazar_effect_2018].

3. **Third, I would have liked to see the authors use and discuss their results in the light of the ENSO index developed specifically for the southern Ecuadorean and northern Peruvian coasts to characterize conditions during each season (the Índice Costero El Niño – ICEN, see http://www.met.igp.gob.pe/publicaciones/2014/ElNino_v2_27_03_2014_7.pdf). This index seems to reflect much better the oceanic conditions in the area than the ONI index, as the latter covers a huge, nearly Pacific-wide region.**

Thanks for the recommendation.
In fact, we would love to use a local index.
However, the proposed index, does not classify any year between 1999 to 2017 as La Niña (when considering the three-month mean of December/January/February and January/Feburary/March) which strikes me as a bit odd (see Figure \@ref(fig:bp)).
For example, 2011 and 2012 have been La Niña years. 
And La Niña had a strong influence on the coastal vegetation of Peru as we clearly observed on Lomas mountains near Casma and Trujillo [@muenchow_predictive_2013a;@muenchow_soil_2013].
It is also interesting that 2005 was not classified as La Niña despite much lower rainfall than average across the complete study area.

```{r bp, echo=FALSE, message=FALSE, fig.width=12, fig.height=4, fig.cap="Ratio between annual precipitation and median precipitation (median Paita: 25.9mm, median Piura: 71.4mm, median Chulucanas: 272.4mm) measured at the three automated climatic stations (see section 2.2.2). In case the precipitation was smaller than the median, we computed the ratio as: 1 / (precipitation / median(precipitation)) * -1. The years in which vegetation sampling took place are printed in bold (x-axis). The ENSO phases are in correspondence with the ICEN index found at: http://www.met.igp.gob.pe/datos/icen.txt."}
# the result is the barplot figure (bp)
source(here("code/misc/explore_icen_index.R"))
print(bp)
```

The index seems to be interesting but still is a bit experimental as also pointed out in the publication you have indicated.
Therefore, we have refrained from using the ICEN index but we have prominently mentioned it in a rewritten discussion section now named *ENSO, climate change and ENSO indices*:

>>>Naturally, local anomalies possibly overlap, dilute and modify the ENSO signal stemming from the central Pacific. 
Studies have shown that indices computed from the El Niño region 1+2, which corresponds to an area in the tropical Pacific close to the South American coast, already provide better correlations with climate anomalies in northern Peru and southern Ecuador [@kao_contrasting_2009;@salazar_effect_2018]. 
At least two further local indexes seem to be also promising for predicting climate anomalies along the Peruvian coast.
The first is the Índice Costero El Niño (ICEN; see http://www.met.igp.gob.pe/publicaciones/2014/ElNino_v2_27_03_2014_7.pdf), and the second is an ENSO index for the insurance of agricultural products [@mortensen_enso_2018].

4. **Fourth, I do not agree with several of the statements made in the Conclusions. To begin with, I do not think that there is a current simplistic understanding of ENSO. For several years now, government and academic sectors working in southern Ecuador and northern Peru are officially using locally developed indexes with a complex array of ENSO definitions and characterizations. And I expect this to be reflected with increased presence in the scientific literature in coming years.**

We absolutely agree.
Hence, we have completely rewritten our Conclusions:

>>Constant monitoring is a prerequisite for a better understanding of ENSO-affected ecosystems. 
Since ENSO exerts its most direct influence on tropical regions, and all corresponding ecosystems are endangered without exception, the correct prediction of the expected ENSO effects are a cornerstone of informed conservation and reforestation management. 
Conserving and reversing the state of degraded ecosystems also helps to mitigate the effects of floodings and landslides due to excessive rainfalls, and thereby contributes to protect the population and infrastructure from unnecessary harm [@fraser_peru_2017; @ramirez_understanding_2017].
Currently, the regional government has set up eleven restoration projects in northern Peru which are trying to protect the endemic vegetation and increase vegetation cover through reforestation [@cordova_piura_2019].
Additionally, NGOs and mining companies are also working on reforestation projects [@gutierrez_anuncian_2015].
Most of these projects rely on extreme events (like the 2017 El Niño) to boost plant growth to overcome the otherwise very limited access to water.
Our experimental results indicated that water availability showed a non-linear relationship with biodiversity and plant cover.
Therefore, restoration projects might consider water regulation in combination with a sensible amount of fertilization [@dahlsjo_local_2019].
The same applies to a sustainable agrarian management.
Indeed, agricultural development has been very intense over the last 20 years in the study area.
Personal communications with the main local companies (e.g., https://www.ecosac.com.pe/) indicated that extreme events are an important issue.
On the one hand, El Niño has the potential to boost crop productivity.
On the other hand, especially short and very intense rainfall might create havoc during the crop season as was the case during the 2017 El Niño.
Additionally, our results clearly reveal patterns in the spatial variability of biomass production in response to water availability along the climatic gradient, which might be also an important information for agrarian and restoration activities.

5. **Furthermore, after reading this section several times, it is not clear to me how the present study can achieve the statements presented in the Abstract and barely touched upon in the rest of the manuscript and in this final section in particular. The abstract mentions that the results from this study can help “protect local human populations from devastating ENSO effects through informed conservation management and restoration activities, and that the study can help equip farmers with information on sustainable agrarian management”. I agree that both are valid and important points to consider, but I would have preferred the authors to expand on this and provide logical connections between their results and restoration ecology approaches in seasonal tropical environment and TDFs. Likewise, I would think that the high-tech agricultural development that has occurred in the study area in the past 20 years has produced knowledge in terms of dealing with variable climatic conditions and I would have expected some literature review on this as well.**

Again, we absolutely agree and we appreciate the reviewer's concerns about the expected impact of our results.
Therefore, we have completely rewritten the conclusion section which now, as suggested, focusses much more on restoration ecology approaches (see also reply to your previous comment).
Unfortunately, we could not find any peer-reviewed scientific literature on high-tech agriculture from North Peru, and therefore refer to personal communications we had with local business and governmental stakeholders.

6. **29: why is the area one of the most climatically vulnerable regions on earth?**

Thanks for noting, we toned down as follows while also offering a reference:

>>> one of the most variable regions on Earth in terms of climate [@wehrden_global_2012].

7. **30: I have never seen the term “endemic” applied to an ecosystem. How do you define an endemic ecosystem?**

We replaced *endemic ecosystem* by *ecosystem featuring a high rate of endemism*.

8. **31: I beg to disagree, but TDF almost as high in biodiversity as humid tropical forests? In what terms?**

We agree this is indeed surprising but it is what the literature says:

> "Tropical dry forests have smaller stature and lower basal area than tropical rain forests (Murphy & Lugo 1986), they are diverse both locally and regionally (Givnish 1999) and although less known, they are almost as diverse as tropical rain forests."

@espinosa_what_2011

> "There is also a general assumption that rain forest is the biome richest
in species in the tropics. However, a recent evaluation of plant species diversity in dry forests across Latin America and the Caribbean suggests that they have more than 7500 woody species, almost certainly an underestimate because it is based on only 1600 inventories covering 125,000 square km across this vast region (for example, the dry forests of NE Brazil cover 750,000km^2 alone; Figure 1). This suggests that in their entirety, the dry forests of the New World may
contain as many or possibly more species than the Amazon rain forest, where the most recent estimate of tree species numbers is ~6200. Similarly, considering all plant growth forms, the savannas of Central Brazil contain 11.384 species, 35 more than the Brazilian Amazon rain forest."

@pennington_tropical_2018

9. **34: TDFs are understudied, and a specific reference for this statement would also be Escribano-Avila et al. 2017**

Thanks! 
We have added @escribano-avila_biodiversity_2017 to the text.

10. **37-39: the group of Carlos Espinoza from UTPL in Ecuador has conducted several studies looking into these questions. There are also quite a few studies in Mexico and Colombia, specifically in TDFs. I would consider some of those, key references for the manuscript.**

Thank you very much for the literature recommendation.
We now additionally cite @ulrich_climate_2014 in the introduction.
Additionally, we have added following sentences to the discussion section:

>>>However, the most important topographic variable is altitude since it is highly correlated with mean annual precipitation in the study area. 
This is also in correspondence with results from Mexico where topography played also a part in explaining the composition of dry forests [@mendez-toribio_effects_2016].

and ...

>>>Other studies have also confirmed the importance of soil for the composition of the vegetation in drylands.
For example, @ulrich_climate_2014 found that soil fertility and variability in rainfall and temperature are the most important predictors for beta diversity.
Similarly, soil plays an important role in Colombian dry forests [@gonzalez-m_disentangling_2018] 
Finally, @espinosa_what_2011 found out that soil moisture, soil temperature and nitrogen explained in large parts the floristic composition of Tumbesian dry forests, which is quite close to our study area.

11. **80: “uniform during normal years”. Based on the previous text, what is considered normal? No rains or even drought?**

We are sorry, of course, we were referring to neutral years, i.e., neither La Niña nor El Niño years.
We have updated the text accordingly.

12. **82: Were plants collected? How was the ID process? In my experience, even after years of monitoring permanent plots in these seasonal environments, there is a percentage of unidentifiable material, especially if it is collected during the vegetative growth phase.how did you ensure across-season uniformity in species ID**

We collected all plants which were unidentified in the field.
Our co-author, Eric Rodríguez, who is a botanist and leading the herbarium at the University of Trujillo, determined the unidentified species.
Still, a few species remained for which either the genus nor the species could not be determined.
These were only named after their family, e.g., unkn_Sola for unknown Solanaceae.
If more than one species from the same family could not be identified, we simply numbered the names, e.g., unkn_Sola1 and unkn_Sola2.
If even the family could not be determined, the plant was identified as unknown.
To be able to recognize the same species, especially the unidentified species, during each sample  we have taken various fotos of each single species.
The data sampled in the field was stored in an Access database using specifically designed forms to ensure that the same attributes for each plot were recorded and that the same species names were used for each plot.

```{r unknown-table, echo=FALSE}
pa = readRDS(here("images/11_ct_pa.rds"))
d = data.frame(unidentified_species = sort(grep("unkn", names(pa), 
                                                value = TRUE)))
cap = paste("Unidentified species. Please note that unkn\\_Faba1 was", 
            "determined later on which is why it disappeared from the table.")
kable(d, "latex", booktabs = TRUE, escape = TRUE, 
      # align = "c",
      # col.names = linebreak(nms, align = "c"),
      caption = cap)
```

13. **83: What do you mean by “randomly sampled”? You randomly sampled from 50 plots? Or do you randomly established 50 plots along the gradient?**

We have clarified as follows:

>>> We randomly sampled 50 vegetation plots of 30 × 30 m^2^ along the climatic gradient between Paita and Chulucanas (Fig. 1).
To ensure that all parts of the study area were equally likely to be sampled, we stratified the sampling by distance to the Pacific Ocean in ten sampling zones.

14. **84: what are those 10 classes? They are not mentioned anywhere in the text or figures.**

As mentioned in the previous reply, the ten classes are just zones stratified by distance to the sea as shown in Figure \@ref(fig:sampling-zones).
We have added the figure to the Supplementary Information and referenced it in the main document.

```{r sampling-zones, echo=FALSE, fig.cap="Study area stratified by distance to sea in 10 classes."}
p_1 = readRDS(here("images/18_sample_stratification.rds"))
print(p_1)
```

15. **84: that whole study area is a mosaic of different land uses and scarce (if any) areas devoid of human use. Humans and especially free-ranging cattle could turn into major disrupting factors of vegetation dynamics. Was this considered when selecting location of plots, and more importantly, when evaluating results?**

You are absolutely right.
The humidity gradient is at the same time a land use gradient.
Please refer also to the methods section in which we describe in detail how plot locations were sampled.
Prior to the sampling, we have excluded explicitly urban, riparian and agricultural areas. 
The latter included mostly farmed fields but also fenced areas to enclose goats, and rarely also cows and horses.
Still, some of these animals are allowed to roam freely the properties of their owners.
Therefore, we added a sentence to the methods...

>>>Agricultural activity includes field farming and extensive grazing caused by free-ranging livestock.

... and to the discussion section:

>>>Finally, free-ranging livestock might also have an impact on vegetation dynamics in the study area, especially in its eastern parts with more than 100mm of annual precipitation.
Livestock might alter floristic composition and diversity while supporting especially plants adapted to grazing.

16. **90: Did you sample soils only once or for every field season? If once, any idea if soil composition or chemistry could vary between years?**

We collected soil samples only once, and therefore have updated the text as follows:

>>>We randomly took and mixed three soil samples from 15cm and 30cm depth in each plot in 2011. 
Soil composition should not vary too much between the years in which vegetation sampling took place since the study area largely features arenosols, i.e., a very weekly developed soil type.
On the other hand, soil chemistry can vary largely between two years.
For instance, El Niño events can increase soil respiration, i.e., soil CO\textsubscript{2} emissions, by a factor of 100, especially in the vicinity of trees [@salazar_tree_2019].

17. **99: the vegetative period is from-to?**

The vegetation period usually lasts from March to May, we now also say so in the text.

18. **132: what is the Londo-scale?**

Please refer to our reply to comment #8 of reviewer #1 who also raised the same question.
We are sorry that we have not made clear from the beginning that we are referring to the decimal scale created by Londo [@londo_decimal_1976].

19. **174: results from Table 1, I would have liked to see some max-min ranges of each variable.**

Thanks for the advice. 
We have added the requested columns to the table.

20. **256-284: Very interesting discussion, and I wonder what differences might be revealed if the data is stratified by growth form (annual herbs vs perennial woody species for instance).**

```{r lf-dca, eval=TRUE, echo=FALSE, fig.cap="DCA species plot with the species colored by lifeform.", out.width="75%", fig.align="center", message=FALSE}
# attach packages
library("vegan")
library("dplyr")
library("lattice")
library("magrittr")
library("ggplot2")
library("RSQLite")
# source own functions
source(here("code/funs/ordi_helper_funs.R"))
source(here("code/funs/bestisomap.R"))
# attach data
londo = readRDS(here("images/11_ct_londo.rds"))
cover = readRDS(here("images/11_ct_cover.rds"))
env_data = readRDS(here("images/11_env_data.rds"))
lf = readRDS(here("images/11_condensed_lifeforms.rds"))

# dca = ord_fun(londo, env_data, method = "dca")
dca = decorana(dplyr::select(londo, -pnr), iweigh = 1)

sc = scores(dca, display = "species")
# just checking
# setdiff(rownames(sc), lf$species)  # 0, perfect
# setdiff(lf$species, rownames(sc))  # 0, perfect
sc = as.data.frame(sc)
sc$species = rownames(sc)
sc = inner_join(sc, lf, by = "species")
cols = c("darkgreen", "yellow3", "brown")
p_1 = xyplot(DCA2 ~ DCA1, data = sc, 
             groups = sc$type, col = cols, pch = 19, 
             auto.key = list(corner = c(0, 0)),
             par.settings = 
               list(superpose.symbol = list(col = cols, pch = 19)),
             scales = list(alternating = c(1, 0),
                           tck = c(1, 0)))
print(p_1)
# # another idea would be to use the percentage of each lifeform for each plot
# # as predictors in envfit
# # run through plots, merge lifeform and compute percentage of each lifeform
# lf_agg = plyr::ldply(seq_len(nrow(cover)), function(i) {
#   tmp = cover[i, ] %>%
#     reshape2::melt(.)
#   id = tmp$pnr[1]
#   tmp = inner_join(tmp, lf, by = c("variable" = "species"))
#   tmp_2 = group_by(tmp, type) %>% summarize(portion = sum(value)) %>%
#     mutate(percentage = portion / sum(portion) * 100) %>%
#     select(-portion) 
#   tmp_2 = melt(tmp_2) %>% dcast(., variable ~ type)
#   tmp_2 = rename(tmp_2, id = variable)
#   tmp_2$id = id
#   # return your result
#   tmp_2
#   })
# fit = envfit(dca, select(lf_agg, -id))
# 
# # retrieve arrow heads manually, see here:
# # browseURL("https://stackoverflow.com/questions/14711470/plotting-envfit-vectors-vegan-package-in-ggplot2")
# 
# coords = scores(fit, "vectors") * vegan:::ordiArrowMul(fit)
# plot(dca, display = "sites")
# plot(fit)  # coords correspond to arrowheads 
# # even better would be to compute the envift for each year but then you would
# # have to run a DCA for each year individually

```

Thank you for the compliment!
And yes, you are right, it would be interesting to see if there are differences along the gradient between lifeforms.
Fig. \@ref(fig:lf-dca) is a first attempt to delve deeper into this question.
However, for the moment we chose not to discuss the implications further in the manuscript since the editor asked us to stick to the most important results.

21. **368: Mention of Panel b is actually c?**

Thanks for spotting.
We changed the caption accordingly.

# Bibliography

### 2020-08-10 text/ecography/round_3/reply_to_reviewers_2.Rmd
migrate gitlab repo to github

---
output:
  bookdown::pdf_document2:
    keep_tex: no
    latex_engine: xelatex
    # float is needed for floating figures
    extra_dependencies: ["fontspec", "sectsty", "float"]
    toc: false
    fig_caption: yes
    number_sections: true
header-includes:
  # adjust figure captions
  - \usepackage{caption}
  - \captionsetup{font={footnotesize,bf,singlespacing}}

title: "Point-by-point reply to reviewers' comments"
bibliography: ../transect_refs.bib
csl: ../ecography.csl
# biblio-style: "natbib"
link-citations: true
---

```{r setup, include=FALSE}
library("knitr")
library("kableExtra")
library("tidyverse")
library("here")
library("RSQLite")
library("sf")
# make sure that figures are allowed to fload (hold the assigned position and are not always put on top of a page)
knitr::opts_chunk$set(fig.pos = 'H')
# needed for dates in experiment figure species richness, otw months are displayed in German
Sys.setlocale("LC_TIME", "C")
```

# Reply to subject editor Bethany Bradley

Dear Dr Bradley,

thank you very much for handling our manuscript.
We are pleased to resubmit a revised version of our manuscript entitled  *Monitoring and predictive mapping of floristic biodiversity along a climatic gradient in ENSO's terrestrial core region, NW Peru*.
We are also very thankful for the constructive comments of both reviewers which helped to improve substantially our manuscript.
Please find below a point-by-point reply to all concerns raised by the reviewers.
In accordance with your and the reviewers' suggestions, we have:

- integrated into a coherent story the vegetation transect analysis and the irrigation-fertiization experiment.
- deleted parts of the discussion to discuss above all findings supported by our data.
- completely rewritten the Abstract, the Conclusions and in great parts the Introduction to make our aims and the corresponding results clearer.
The Conclusions now exclusively feature the 'take-home' points providing an answer to the hypotheses posed in the introduction. 
- added a new figure showing how species richness developed in the experimental study.

After having addressed thoroughly and carefully the reviewers' comments for the third time, we now hope that you deem our manuscript suitable for publication in *Ecography*.
In any case, we would like to express our gratitude again towards you and the reviewers for taking the time to consider and revise our manuscript.

Yours sincerely,

The authors

# Reply to reviewer #1

1. **The different studies are not well combined and appear too much in parallel, especially in the discussion section. Also it is unclear why the spatial prediction was performed because it does not appear to be needed to discuss the findings. In general, the manuscript needs a lot more clarification in the presentation of the aim, hypotheses, and discussion of the results.**

Thanks for taking the time to revise and improve our manuscript.
We have integrated into a single story our transect analysis and the water-fertilizer experiment.
Please refer to the replies to comment #7, 11, 23.

The spatial prediction is crucial for this study.
We have added a new paragraph to the introduction, described the reasoning in more detail and highlighted the resulting findings more prominently throughout the whole manuscript. 
Please refer also the replies to comments #16-19.

Furthermore, we have rewritten our hypotheses, clarified the general aim of our study, have deleted parts of the discussion and have completely rewritten our Conclusions mirroring now the posed hypotheses, and hence the structure of the manuscript.
Please refer to the replies to comments #7-14, 23, 26.

2. **The abstract is not very clearly written and is not presenting a good summary of the paper. E.g. the first two sentences are a poor motivation for this study (the motivation is to study TDF only because they are understudied?). It would be better to start with a motivation that highlight the relevance, e.g. that these areas feature unique biodiversity but that the impact of ENSO on biodiversity patterns is still unknown.**

Thanks for the advice.
We have rewritten as follows:

>Tropical dry forests (TDF) are heavily shaped by the El Niño Southern Oscillation (ENSO) in NW Peru, where especially El Niño episodes bring rain to arid to semi-arid areas.
However, the resulting effects on biodiversity patterns remain largely unknown as well as the effect of environmental variables on the floristic composition under varying rainfall patterns.
Therefore, we studied the spatio-temporal effects of different ENSO episodes on floristic biodiversity along a climatic gradient ranging from the coastal desert to the Andean foothills.
We sampled 50 vegetation plots in four years representing different ENSO episodes.
To highlight the spatio-temporal changes in floristic composition and beta diversity across ENSO episodes, we predicted ordination scores with a Generalized Additive Model.
We applied variation partitioning to test if topographic or edaphic variables gained in importance during more humid ENSO episodes.
Additionally, we executed an irrigation-fertilization experiment to quantify the beneficial effects of the water-nutrient interaction under different simulated ENSO rainfall scenarios. 
Plant species richness increased under humid conditions during the humid La Niña (2012) and the moderate El Niño episodes (2016), and slightly decreased under the very humid conditions during the coastal El Niño episode (2017).
The spatial prediction revealed that specific vegetation formations became more pronounced with increasing water input, but that a large water surplus led to the disruption of the strict order along the climatic gradient.
Edaphic and topographic variables gained in importance with increased water availability (2012 & 2016), however, this effect was not further amplified under very wet conditions (2017).
The experiment showed that plant cover under Super Niño conditions was three times higher when fertilized.
Overall, our spatial predictions concede detailed insights into spatio-temporal ecosystem dynamics of TDF in response to varying rainfall caused by different ENSO episodes while the results of the experiment can support farmers regarding a sustainable agrarian management.

3. **Carefully check for logical reasoning in the abstract. E.g. you write that you wanted to test the effect of other variables when water is no longer the limiting factor. But it was not mentioned before that water is a limiting factor.**

Thanks, please refer to the reply to the previous comment.

4. **Your aims and hypotheses other than just studying biodiversity in the TDF (which is quite a broad aim) is not getting clear. Be more specific here on what you want to find out with your study and what you expect.**

Please refer to the reply to comment #2.

5. **It appears unclear at that point how the results of this study “can help the local population from devastating ENSO effects”.**

We have toned down by rewriting the text as follows:

>Overall, our spatial predictions concede detailed insights into spatio-temporal ecosystem dynamics of TDF in response to varying rainfall caused by different ENSO episodes while the results of the experiment can support farmers regarding a sustainable agrarian management.

6. **Line 11-15: As your study area is in Peru you should mention what effect the ENSO phases have for your study area instead of a global description of the phenomenon.**

We have rewritten as follows:

>ENSO is a recurrent climate phenomenon causing anomalies around the globe [@capotondi_understanding_2015].
ENSO consists of alternating episodes.
El Niño represents the warm and La Niña the cold episode of this cycle. 
ENSO neutral refers to periods when neither El Niño nor La Niña episodes are present. 
The study area is part of the core region where the ENSO phenomenon exerts its greatest terrestrial influence [@richter_monitoring_2005]. 
Generally, El Niño episodes bring more rain than usual and La Niña episodes enforces the already dry conditions in the study area [<50 mm of annual precipitation along the coast of NW Peru; Fig. 1D;  rollenbeck_climatic_2015].

7. **In general the aim as well as hypotheses are not getting very clear in the introduction. E.g. Line 40: “To tackle the outlined issues...” Be more specific here what issues you’re talking about. In the sentences above you talk about the effect of different soil properties. But you are not addressing this in your methods. Also it’s not getting clear why you want to study edaphic and topographic properties on floristic composition. How does this link to your major aim to study bidoversity with different ENSO phases? This needs a lot more clarification. Also the motivation why different ENSO phases are studied is not getting clear. Make clear why this is relevant.**

Thank you for pushing us to:

- make our aims and hypotheses clearer.
- clarify why it is important to study biodiversity during different ENSO episodes.

We have added several new sentences to the introduction.

With regard to different ENSO episodes, we added:

>Therefore, one objective of this manuscript is to study the effect of different ENSO episodes, including La Niña episodes, on the floristic composition, comprised of all observed vascular plants, of the same ecosystem.

With regard to the predictive mapping, we added a new paragraph (see also the reply to comment #15) while already mentioning the corresponding aim as follows:

>We aim at spatially predicting, for the first time, how different ENSO episodes might change and reshape the spatial distribution of the floristic composition in the study area.

With regard to the impact of environmental variables, we added:

>Hence, this study investigates how much the impact of edaphic and topographic variables varies in explaining the floristic composition between dry and humid ENSO years.
Additionally, an irrigation-fertilization experiment should help to quantify in more detail the beneficial effects of the water-nutrient interaction on vegetation diversity and cover.

Additionally, we have rewritten the aims of our manuscript as follows:

>In summary, in this manuscript we study the effects of different ENSO episodes on biodiversity patterns and floristic composition  as well as the effect of environmental variables on the floristic composition under varying rainfall patterns.
Specifically, we tested the following three hypotheses:
>
(1) We investigated the change in the different levels of biodiversity along a pronounced climatic gradient across different ENSO episodes. 
Overall, we expected a boost in biodiversity as soon as water was no longer the predominant limiting factor.
We were particularly interested in how different ENSO episodes modify the spatial distribution of the floristic composition.
Overall, we expected a more pronounced development of vegetation formations along the climatic gradient with wetter conditions [@engelbrecht_drought_2007].
>
(2) Water availability limits plant cover and species richness along the entire climatic gradient in the (semi-)arid study area.
Therefore, we were interested in evaluating the effects of soil characteristics and topography changed in shaping floristic composition and vegetation cover (biomass production) with varying rainfall patterns as caused by different ENSO episodes. 
We expected that the influence of topographic and edaphic variables might gain in influence in wetter years, i.e., as soon as water is no longer the limiting factor.
>
(3) To quantify the beneficial effects of the water-nutrient interaction on vegetation cover development and species richness, we executed an irrigation-fertilization experiment simulating different ENSO rainfall scenarios.
We expected that the water input of the rainfall scenario representing a Super-Niño episode in combination with the fertilizer would be most conducive to plant cover and species richness development.

As now written in more detail in our hypotheses, we also wanted to test if the impact of edaphic and topographic variables change as soon as water is no longer the limiting factor in the study area.
Different ENSO episodes create different rainfall scenarios, and therefore provide an excellent opportunity to test if and how the influence of environmental variables change the floristic composition and biodiversity.

Please also note that we describe in section *2.2.2 Environmental variables* which edaphic variables were sampled in the field.

8. **Line 45/46: Don’t mix up hypothesis and methods here**

Please refer to the reply to the previous comment and to comment #18.

9. **Line 50/52: Why do you expect that topography/soil properties have a higher effect on biodiversity in wetter years? The hypothesis is not getting clear.**

Please refer to the reply to comment #7.

10. **Line 53-57: This is a description of the method and unclear what the corresponding aim is. Better describe the hypothesis/aim here and describe the methods applied to approach this later on.**

Please refer to the reply to comment #7.

11. **In the introduction it is unclear what kind of biodiversity you’re studying (not even mentioned that you focus on vegetation!). But that’s relevant for the understanding/hypotheses. E.g. for trees, I would not expect that biodiversity changes during ENSO cycles. So please state what you’re studying and why you expect changes.**

You are absolutely right.
We have added following sentence to the introduction to make clear what we study:

> Therefore, one aim of this manuscript is to study the effect of different ENSO, including La Niña, episodes on the floristic composition, comprised of all observed vascular plants, of the same ecosystem.

Please refer also to the reply to comment #7.

12. **Section 2.2.1: You often talk about plant development. Be more specific here. Do you mean biomass/coverage/diversity?**

We have clarified as follows:

>In addition to the plot survey at the landscape scale, we conducted a small-scale irrigation-fertilization experiment to quantify the beneficial effects of the water-nutrient interaction on species richness, plant cover and biomass production.

13. **2.2.1: Experimental in addition is great to test effect of water availability in a more systematic way than under real-world conditions. But was it vegetation on the experimental plots from the same ecosystem? Or how can the two surveys be compared?**

We have clarified as follows:

>We established the experiment in the middle of the study area on on a protected dryland forest inside the campus of the University of Piura (5°10'S, 80°38'W, see Piura in Fig. 1).
Hence, the experiment was located in the same ecosystem as the landscape study (see previous paragraph) and more specifically in the middle of the observed precipitation gradient, ranging from 26 mm (Paita; annual median precipitation) to 272 mm (Chulucanas; Fig. 1).

Please refer also to Appendix 2 for a more detailed description of the experiment.

14. **Section 2.2.2: Because the hypotheses/aims are not well described in the introduction, it is not clear now why soil samples were taken.**

Please refer to the reply to comment #7.

15. **Section 2.2.2: Why did you use NDVI from MODIS (250m resolution) when you have much smaller plots? Wouldn’t something like Landsat/Sentinel-2 be more appropriate? Also see comment above: Not clear at this point why these variables were calculated because the aims/hypotheses were not made clear. Also at the beginning of the method chapter a short summary would help gaining understanding of what you did for which purposes before going into detail here.**

We have not used Landsat/Sentinel-2 imagery due to following reasons:

- Sentinel-2 data have only become available globally since March 2017.
For consistency reasons, we believe it is better to use the same sensor for each modeled year instead of mixing data from different sensors for different years.
- MODIS data has a much better temporal resolution than Landsat imagery, i.e., as opposed to MODIS data, Landsat data was not available for the vegetation period in each sampled year whereas there were thress MODIS images available for each vegetation period.
Please note also that there is not one pair of locations that share the same MODIS pixel.

16. **2.3.2: Here you mention for the first time, that one aim is to do spatio-temporal prediction. This should have been already mentioned in the introduction. Also there please make clear WHY this is an aim of your study. Predictive mapping even occurs in the title but it is not made clear why you’re doing this, other than  for visualization as you mention in the introduction.**

You are absolutely right, we should have already emphasized in the introduction the importance of predictive mapping, and why we are doing it.
Therefore, we have added following paragraph to the introduction:

>Statistical learning has a long tradition in biodiversity research for detecting patterns in ecological data and predicting from these patterns [@guisan_predictive_2000].
It is especially useful for quantifying species-environment relationships and for predicting the occurrence of species or species richness (predictive mapping), and therefore is an indispensable instrument for biodiversity management [@fremout_mapping_2020].
Furthermore, predictive mapping can reveal ecological changes in space and time which otherwise might be hard or impossible to detect [@lovelace_geocomputation_2019a].
Therefore, it is a crucial tool for gaining detailed insights into ecosystem dynamics, which other more descriptive methods such as global biodiversity measures and ordination techniques cannot capture.
We aim at spatially predicting, for the first time, how different ENSO episodes might change and reshape the spatial distribution of the floristic composition in the study area.

17. **2.3.2: By using NDVI as the only proxy for assemblages, all you do is to assume that diversity increases with vegetation cover and/or biomass (as the NDVI is increasing with cover/biomass but not with diversity if the biomass is the same). Probably there is a good correlation between cover and diversity but this scientific explanation should be made clear here. However, in line 256-257 you write that 2007 was wetter and biomass increased (consequently increased NDVI) but diversity decreased. How does this fit to your NDVI~diversity model? **

In fact, we have modeled the first axis of the DCA as a function of the interaction NDVI:year.
The input for the DCA was a plot-species matrix consisting of the decimal-scale transformed cover values, hence the resulting scores comprise both cover and diversity (number of species), and represent the floristic composition of the observations condensed to a single number.
Therefore, a score can also represent a plot whose number of species has decreased while the overall plant cover has increased.
In fact, Fig. 2 in the manuscript provides a visual representation how the position of one plot can change in ordination space with changing cover and species richness values.
Please refer also to the reply to the next comment in which we included a figure that shows the relationship between the response, i.e., the first DCA axis, and NDVI by year (Fig. \@ref(fig:dca-ndvi)).
Thank you again for advising us to include this figure which is crucial since it makes transparent the observed relationship between the scores of the first axis and NDVI and how this relationship changes between years.

18. **Also, why do you include the year as predictor? By doing so, you don’t set up a model that would be in any way applicable beyond the years you sampled. Why are you not assuming that the relationship between NDVI and diversity stays the same independently of the year? **

We included the year-NDVI interaction because we wanted to allow the model to have a different intercept and a different slope for each year (Fig. \@ref(fig:dca-ndvi)).
Additionally, we used a spline smoother to allow for the observed non-linearity (Fig. \@ref(fig:dca-ndvi)).
And you are absolutely right that it is crucial to give the reader the opportunity to have a look at the relationship between NDVI and the response variable.
Therefore, we have also put Fig. \@ref(fig:dca-ndvi) into the Supplementary Information and have cross-referenced it in the main text.

```{r dca-ndvi, echo=FALSE, message=FALSE, fig.align="center", fig.cap="Relationship between the response (DCA1) and NDVI by year. To aid visual interpretation, line smoothers were added.", out.width="70%"}
data = readRDS(here("images/15_model_input_data.rds"))
ggplot(data, aes(x = ndvi, y = DCA1, group = year)) +
  geom_point(aes(color = as.factor(year)), alpha = 0.4) + 
  labs(color = "year") +
  xlab(label = "NDVI") +
  ylab(label = "DCA1") +
  geom_smooth(aes(color = as.factor(year)), se = FALSE) + 
  theme_bw() +
  coord_fixed(ratio = 0.15)
```

The aim of our study was to visualize by means of predictive mapping how the floristic composition changed between years, i.e., how different ENSO phases shape the spatial distribution of vegetation assemblages.
Therefore, we were not interested in predicting unobserved or future years.
But if we were, it would be possible to do so by taking a mean intercept and a mean slope.
However, doing so, would cancel out the majority of the interannual differences.

To make our aims clearer, we have rephrased the corresponding hypothesis as follows:

(1) We investigated the change in the different levels of biodiversity along a pronounced climatic gradient across different ENSO episodes. 
Overall, we expected a boost in biodiversity as soon as water was no longer the predominant limiting factor.
We were particularly interested in how different ENSO episodes modify the spatial distribution of the floristic composition.
Overall, we expected a more pronounced development of vegetation formations along the climatic gradient with wetter conditions [@engelbrecht_drought_2007].

19. **More general question: why do you need the spatial prediction? Isn’t it sufficient to look at the plots that were sampled to test your hypotheses?**

Only the predictive mapping approach reveals how different ENSO episodes changed the spatial distribution of floristic composition and diversity.
An ordination by itself only reveals that diversity has changed along the gradient but cannot reveal spatial patterns.
But thank you for pointing us to formulate our hypotheses and aims in more detail.
Please refer also to the replies to comments #16, 18 and to the reply to comment #4 of reviewer #2.
And please note also that we discuss the outcome of the spatial prediction in detail in section *4.2 Floristic gradient mapping across ENSO episodes*.

20. **Numbers in Fig.2 not readable**

We now show only a few plot numbers which should illustrate how the position in ordination space has changed for a few selected examples throughout the years (Fig. \@ref(fig:dca)).

```{r dca, echo=FALSE, fig.cap="Plotting the scores of the second DCA axis against the scores of the first DCA axis. To better visualize the cross-annual changes, the scores of all years were added in light gray to the background of each panel. The black dots refer to the scores of the year given in the strip of each panel. Numbers above the dots are the plot IDs which were ordered in accordance with the distance to the sea, i.e. plot number 1 is closest and plot number 50 farthest away from the sea. To avoid cluttering the figure with too many numbers, we have only added a few selected plot numbers.", out.width="100%", fig.align="center"}

library("lattice")
load(file = here("figures/14_londo_dca.rda"))
print(p_1)
```

21. **Figure 4: Scatter plot NDVI~richness would be helpful in addition as this represents the basis for the spatial prediction. Also it is unclear to me what the numbers (and colors) mean. What are the negative and positive values? According to the text below it appears to me as if the number correspond to vegetation types. But if so, I don’t understand why its expressed as a gradual color scale and with positive and negative numbers.**

The scores of a DCA axis are of type numeric, and hence can take on any value between + and - infinity.
Since the score of a DCA can be understood as the floristic composition condensed to a single number, we have interpreted certain DCA value ranges as belonging to a specific vegetation formation.

We added following explanation to the corresponding results section:

>Since the predicted DCA scores represent the floristic composition of a pixel, clusters of pixels sharing similar values (visible as similar color tones in Fig. 4) can be understood as belonging to the same vegetation formation.

22. **Figure 5: I think the indication of the year is missing (at least I assume the 4 plots correspond to the 4 years)**

Thanks for spotting.
The year labels should have been there in the first place.
Changed as requested.

23. **Also in the discussion/conclusion it remains unclear what the study is exactly about. Change in vegetation patterns due to ENSO, improved agriculture, nature conservation? The claimed implications are often not justified by data. e.g. you have higher diversity when more rain is there but does this have implications for nature conservation as claimed in the paper?**

We have rewritten the Conclusions in such a way that they now mirror the hypotheses and therefore also the overall structure of our paper, i.e, they now present the most important result for each hypothesis and draw a general conclusion from the chosen integrated approach.
We also made clear how our results can be of use in nature conservation and improved agriculture at the same time.
Please note also that we have deleted a complete section in the discussion to also streamline the discussion with the posed hypotheses (see also the reply to comment #26).
The Conclusions now read as follows:

>Alpha and gamma diversity increased with wetter conditions during the humid La Niña (2012) and moderate El Niño episodes (2016), though, a few dominant species largely prevailed under the very humid conditions of the coastal coastal El Niño episode (2017), which then even resulted in a slightly decreased alpha diversity.
Still, plant cover was greatest by far during the most humid conditions.
Beta diversity also increased with wetter conditions but did not change significantly between years in accordance with the test for multivariate homogeneity of groups dispersions.
As useful as such global measures are in general, they are in essence a single descriptive figure which naturally lack the capability of capturing spatial diversity patterns.
By contrast, our spatial predictions revealed drastic changes in the distribution of species composition, beta diversity and plant cover.
In effect, a large water surplus disrupted the strictly ordered vegetation assemblages which at first became apparent along the climatic gradient with wetter conditions.
Hence, spatial predictions might provide more detailed insights into ecosystem dynamics, especially in semi-arid ecosystems where changes in plant cover and distribution patterns are more pronounced than changes in alpha and gamma diversity.
>
>The influence of topographic and edaphic variables on plant diversity and productivity also gained in importance with wetter conditions but this effect was not further amplified with very humid conditions.
Our irrigation-fertilization experiment confirms these findings but also revealed the true beneficial effect of fertilizing under wet Super Niño conditions when plant cover was three times as high as in the unfertilized plots.
>
>In conclusion, our predictive mapping approach can identify locations suitable for restoration and conservation, and thus could form the basis for an informed conservation management (forest management plans).
At the same time, the results of our experiment can equip farmers with information regarding sustainable agrarian management which might increase the productivity on already existing farmland which in turn might help to prevent the destruction of the last remnants of TDF in the study area.

24. **Line 285: You write that grasses have a drought memory. So is it possible that the changes in biodiversity you observed are not due to current ENSO conditions but due to the conditions in the previous year(s)? This would change the outcome completely. Also, only one year per ENSO cycle was sampled. So I wonder if a general conclusion about the relationship between ENSO and biodiversity is justified.**

We are sorry for the confusion.
But in fact, we have largely focused on current-year precipitation effects on vegetation development.
The ordination and modeling results provide ample evidence for these effects.
Still, it might be possible that to a much lesser extent previous year precipitation patterns can also have an effect.
However, this was not the aim of the study, and therefore the collected data cannot provide evidence for such effects.
A former reviewer, your predecessor, acknowledged this already but still asked us to discuss possible time-lagged responses.
Therefore, we have now rewritten as follows:

> The ordination and modeling results of our study provide ample evidence how current-year precipitation affects biodiversity dynamics.
Still, it might be possible that to a much lesser extent previous year rainfall might also have an effect in the study area [@tenhumberg_timelagged_2018].
For instance, @dudney_lagging_2017 observed an increase in grass and a decrease in forb abundance when rainfall was high one year earlier.
Similarly, @walter_plants_2011 suggest that grasses might have a 'drought-memory'.

Since we have observed each ENSO episode only once, we cannot draw general conclusions, and we have not done so in the paper.
Instead we described the observed biodiversity patterns for each episode.
Despite that, it is relatively safe to assume that (though we do not claim so in the manuscript):

- the floristic composition remains uniform across the entire gradient during dry years.
- a certain order of vegetation formations appears along the gradient in more humid years.
- the strict order of the vegetation formations are partly interrupted in very humid years.

25. **326-327: Most important finding is that fertilizer increases biomass? Probably well known… By the way Figure 6 shows that plant coverage increases with water input more than when fertilized.  Where is the corresponding plot for biomass that was used for this statement? Also don't understand why talking about crops here when the study in general (though the experimental sites were placed somewhere else) was about vegetation plots where agricultural area was explicitly excluded (Line 89)**

Indeed, it is probably well known that fertilizing increases biomass.
However, the true impact of fertilizing in the study area only becomes apparent with quite humid conditions (Super Niño) when plant cover was three times higher than in the unfertilized plots.

Please note also that the corresponding figure is now in color (Fig. \@ref(fig:irrigation)).
Additionally, the solid salmon line and the dashed gray lines represent plant coverage in %.
By using the dashes, it is now visible that in the Control panel fertilization had no effect and that the two lines overlay each other.
Thanks for helping us to make the figure better understandable.

You are also right, that the figure actually shows the increase in plant cover which we now clarified in the text (see below).
Additionally, we now explicitly state in the discussion that agricultural activity increases and that our experiment might help to improve the productivity on the already existing farmland instead of making disappear the last TDF remnants found in the study area.
Please note that the experiment was carried out in the study area, please refer also to the reply to comment #12.
We have rewritten as follows:

>In fact, the water-nitrogen interaction can result in a plant coverage three times as high as achieved when adding the same amount of water but without additional nutrients (Fig. 6).
This is of utmost importance for sustainable yet productive agricultural management in the study area, where farming activity increases along the gradient.
Hence, instead of making disappear the last remnants of TDF in the study area, it would be desirable to increase the productivity on already existing farmland.
The decision on which crops (annual herbal and grass species) to grow and how much fertilizer to use could be further improved by reliable climate predictions on the occurrence of ENSO episodes.

```{r irrigation, echo=FALSE, out.width="100%", fig.align="center", warning=FALSE, message=FALSE, fig.cap="Results of the irrigation-fertilization experiment. The blue line refers to the water input, note that the left y-axis was log-transformed. The solid salmon and the gray dashed lines represent the observed plant coverage in \\% (right y-axis)."}
# include_graphics(here("figures/17_nutrient_irrigation_experiment.png"))
load(here("figures/17_nutrient_irrigation_experiment.rda"))

# to plot manually axes labels outside of the panel, we have to disable the
# clipping beforehand
my.setting = trellis.par.get("clip")
my.setting$panel = "off"
trellis.par.set("clip", my.setting)
# trellis.par.get("clip")

# plot
print(p_1)
```

26. **Section 4.4.4: This section seems to be off scope because none of your own results is coming in here.**

<!-- 
Reviewer#2 this time has been reviewer#1 in the previous round.
However, reviewer#2 in the first round asked us to discuss ENSO indices.
Hence, we can delete the corresponding section.
-->

To discuss especially findings supported by our data (as also requested by you earlier), we have deleted the corresponding section.

# Reply to reviewer #2

1. **I am glad that the authors have taken my (and of the reviewer 2) previous comments and criticisms seriously. I think the manuscript is greatly improved as a result. This paper is an interesting contribution to understand plant community ecology and their response to climate change using spatial and temporal data and the integration to macroecological and experimental approaches.**

Thank you very much.
We are glad to hear this.
And thank you again for taking the time to improve our manuscript with your valuable feedback.

2. **The introduction is focused on the effect of water availability gradient on floristic composition, but the authors analyzed simultaneously plant diversity and biomass (or cover). Then, I think that it is necessary to include another paragraph in the introduction to present the effects of climate and explicitly how water availability influences biomass productivity and dynamics (or development).**

We have rewritten the introduction as follows:

>> Naturally, water availability drives floristic composition and biomass production in the (semi-)arid regions such as NW Peru [@muenchow_coupling_2013; @muenchow_woody_2013].
In fact, water is probably the most important limiting abiotic factor in NW Peru's tropical dry forests where it controls plant recruitment, establishment and survival, seedling production, vegetation cover as well as species diversity [@balvanera_distribution_2011;@salazar_tree_2020].
However, water availability does not only vary in space but also in time, especially in regions heavily affected by the El Niño Southern Oscillation (ENSO).

3. **The authors also introduce the “ecosystem diversity” term, which is calculated as the axis values of a DCA analysis for each plot and subsequently, these are used to model floristic composition along a spatial gradient. Also, they analyzed “beta-diversity” as a range of DCA axes and compare them among years. Both approaches are based on the analysis of variation of the floristic composition obtained from a DCA and correspond to beta-diversity analyses (Anderson et al. 2011) but differ in the spatial scale in which are thought. On the one hand, the “beta-diversity” analyzed by the authors is the total variation on samples among plots and compare their range among years. Then, this approach is beta-diversity at the regional scale (i.e. the entire study area scale). On the other hand, DCA axes' values for each plot were used to modeling spatial variation of floristic composition. Here, the local variation of each plot was taken into account (i.e. plot scale). It is a difference among both approaches, and it may be clarified in the text.**

You are absolutely right.
To avoid misunderstandings, we now refrain from using the term ecosystem diversity, and now solely use floristic composition throughout the document.
We have rewritten the corresponding methods section as follows:

>One objective of this study was the spatio-temporal mapping of floristic composition represented by the scores of the first DCA axis along the climatic gradient in the study area.
Please note that the resulting prediction map not only shows species composition but also differences in beta diversity on the predicted pixel level.
This is because the units of a DCA axis are given in SDs (see also previous section).

4. **In this way, we recommend including an additional analysis to assess differences among years in the floristic composition. You can use a test of multivariate homogeneity of groups dispersions (see Anderson et al. 2006). Also, you can see Anderson et al. (2011) for a discussion among variation among samples and turnover in beta-diversity analyses.**

Thank you for this interesting suggestion.

In the method section we added following sentence:

>>To test if beta diversity and floristic composition, respectively, varied significantly between years, we analyzed the multivariate homogeneity of group dispersions [@anderson_multivariate_2006].

We added following sentence to the results:

>>Overall, there were no significant interannual differences in beta diversity in accordance with the test for multivariate homogeneity of groups dispersions (ANOVA *F*-test: 0.05069).

And we rewrote the discussion as follows...

>>Beta diversity (expressed as the range of the first DCA axis), increased with wetter conditions (Fig. 3 & Table 1.
However, beta diversity did no change significantly between years (section 3.1).
Concordantly, the change of the shape of the DCA point scatter between the relatively humid years (panels 2012 & 2016 of Fig. 3 and the very humid neutral year (panel 2017 of Fig. 3 can be most likely attributed to the increase of plant cover (see also next paragraph).

... and

>> Overall, our spatial mapping approach concedes detailed insights into how floristic composition and beta diversity distribution patterns change in response to varied rainfall patterns caused by different ENSO episodes.
Similarly, @lovelace_geocomputation_2019a (chapter 13) predicted floristic composition on a Peruvian fog oasis which revealed that the vegetation belts along the mountain slope are partly interrupted. 
Consequently, predictive mapping can contribute to a more nuanced understanding of ecosystem dynamics.
Naturally, such insights cannot be captured by global beta diversity measures such as the analysis of multivariate homogeneity of groups dispersions which in our case simply yielded not significant results (see sections 3.1 & 4.1).

Finally, we also added following sentences to the Conclusions:

>Beta diversity also increased with wetter conditions but did not change significantly between years in accordance with the test for multivariate homogeneity of groups dispersions.
As useful as such global measures are in general, they are in essence a single descriptive figure which naturally lack the capability of capturing spatial diversity patterns.
By contrast, our spatial predictions revealed drastic changes in the distribution of species composition and beta diversity.
In effect, a large water surplus disrupted the strictly ordered vegetation formations which at first became apparent along the climatic gradient with wetter conditions.
Hence, spatial predictions might provide more detailed insights into ecosystem dynamics, especially in semi-arid ecosystems where changes in plant cover and distribution patterns are more pronounced than changes in alpha and gamma diversity.

5. **Also, I recommend running a GAM between mean local alpha diversity and distance to the sea and even including a year as a categorical variable (and potentially their interaction) to add statistical support to the effect of spatial variation at this level of diversity.**

We have now also modeled alpha diversity as a function of the interaction between distance to the sea and year using a GAM and a spline smoother consuming three degrees of freedom (5-fold 100-repeated spatial cross-validation yielded an NRMSE of ~36%).
However, the corresponding prediction map becomes rather smooth with only a small amount of variation (Figure \@ref(fig:spri-dist)).
This is because distance to the sea remains the same independent of the year and simply because distance to the sea gradually increases from west to east.

```{r spri-dist, echo=FALSE, out.width="100%", fig.align="center", message=FALSE, warning=FALSE, fig.cap="Spatial prediction of species richness along the cross-section for different ENSO episodes using distance to the sea as predictor. The urban boundaries of the cities Paita, Piura and Chulucanas are displayed in black. The Piura River and its tributaries are represented by light blue lines with a black contour.",}
library("sp")
library("sf")
load(here("figures/15_spri_preds.rda"))
print(spri_dist)
```

Hence, we have again used NDVI and the interaction of NDVI and year as the predictors (Figure \@ref(fig:spri-ndvi)).
However, the spatial cross-validation is only partly convincing with an NRMSE of 40%.
Still, the predicted species patterns look reasonable.
We have put the spatial prediction map of alpha diversity into the Supplementary Information.
Please note that we additionally used topographic predictors (slope, catchment area, aspect, curvatures) to test if this would yield even better results when doing the spatial cross-validation.
However, this was not the case.

```{r spri-ndvi, echo=FALSE, out.width="100%", fig.align="center", message=FALSE, warning=FALSE, fig.cap="Spatial prediction of species richness along the cross-section for different ENSO episodes using the NDVI and year interaction as predictors (used model: GAM with a spline smoother with 3 degrees of freedom and a Poisson distribution; NRMSE of a 5-fold 100-repeated CV: 40\\%). The urban boundaries of the cities Paita, Piura and Chulucanas are displayed in black. The Piura River and its tributaries are represented by light blue lines with a black contour."}
print(spri_ndvi)
```

We have added the following sentence to the discussion:

>> Finally, the spatial prediction of species richness (Appendix 4) supports the findings of the floristic gradient mapping since it revealed similar spatial diversity patterns and dynamics.

6. **Finally, I think that the integration of macroecological and experimental approaches to this study is very interesting. Then, the results of local species richness of the experimental approach should be included in the main text to improve the discussion about the effect of water availability and nutrient enrichment on local diversity of plant communities.**

We have moved a revised version of the figure showing the development of species richness during the experiment from the Appendix to the main text (Fig. \@ref(fig:spri-exp)).
Additionally, we have added sentences to several sections throughout the document as follows:

Methods:

>> For presenting the results of our irrigation-fertilization experiment we simply used descriptive statistics by plotting the increase in the number of species, plant cover as well as water input against time for the duration of the experiment.

Results:

>> Species richness followed a similar pattern.
That means, highest numbers in species were recorded in the the fertilized Super-El Niño plots, though with just seven observed species the overall number remained low (Fig. 7).

Discussion:

>> Our irrigation-fertilization experiment confirms the findings of the landscape scale, i.e., the increase in plant cover was much more pronounced than the increase in species richness (see Figs. 6 & 7).

```{r spri-exp, echo=FALSE, fig.cap="Mean number of species per treatment level plotted for the duration of the experiment. Each dot represents a sampling visit during which among others the numbr of species was recorded.", fig.align="center"}
spri_exp = readRDS(here("images/17_irrigation_spri.rds"))
print(spri_exp)
```

7. **Line 16: “change and reshape” could be changed by modify**

We have deleted the corresponding sentence.

8. **Line 50-52: The prediction could be better explained.**

Please refer to the reply to comment #3 and also to the replies to comments #16-19 of reviewer #1.

9. **Line 86-87: The word “randomly” maybe removed because the sample design was stratified.**

Changed as requested.

10. **Line 111: change “determination” by identification**

Thanks for spotting.
Changed as requested.

11. **Line 130-133: The comparison of precipitation values among worldclim and local station could be interesting to visualize as supplementary information.**

In fact, we have not used the WorldClim data in our analysis which is why we have deleted the corresponding sentence from the manuscript.
Still, when comparing the climate station data with the WorldClim data it becomes apparent that WorldClim probably underestimates the precipitation west of Piura (middle blue dot in Figure \@ref(fig:wclim)).

```{r wclim, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="The black line represents the precipitation as derived from WorldClim along the transect. The points represent the annual median precipitation values at the climatic stations.", fig.width=6, fig.height=4, fig.align="center"}

# attach data
# env_data containing wordclim precip
env_data = readRDS(here("images/11_env_data.rds")) %>%
  tidyr::separate(col = "pnr", into = c("id", "year"), sep = "_")

# precipitation data of the climatic stations
con = dbConnect(RSQLite::SQLite(), here("data/tables.gpkg"))
# dbListTables(con)
precip = RSQLite::dbReadTable(con, "precipitation")
towns = st_read(con, "towns_jvs") 
dbDisconnect(con)

# get center of towns-polygon
piura = st_centroid(towns) %>%
  st_coordinates %>%
  as.data.frame %>%
  slice(2)

names(piura) = tolower(names(piura))
piura = cbind(piura, name = "piura")

xy = filter(env_data, id %in% c(1, 46) & year == 2011) %>% 
  dplyr::select(x, y)
xy = cbind(xy, "name" = c("paita", "chulu"))
xy = bind_rows(xy, piura) %>%
  arrange(x)

med_precip = group_by(precip, year, station) %>% 
  summarize(precip = sum(precip, na.rm = TRUE)) %>%
  group_by(station) %>% 
  summarize(precip = median(precip)) %>%
  inner_join(., xy, by = c("station" = "name")) %>% 
  arrange(precip)
med_precip$station = c("Paita", "Piura", "Chulucanas")

# ok, use coords of towns
ggplot(env_data) +
  geom_line(aes(x = x, y = prec)) +
  geom_point(data = med_precip, 
             aes(x = x, y = precip, color = station), cex = 1.5) +
  ylab(label = "Precipitation (mm)") +
  xlab(label = "Longitude") +
  theme_bw()
```

12. **Line 138-140: “This measure is given in units of standard deviation (SD) with four SDs corresponding to a complete species turnover”. I think that the number of SD that separate two samples with complete species turnover depends of the analyzed dataset and four SD is not a rule. Please add a citation for this sentence.**

We have clarified as follows:

>> Beta diversity is expressed as the range of the first Detrended Correspondence Analysis (DCA) axis (see below). 
The unit of a DCA axis is given in units of standard deviation (SD).
4 SDs correspond to a complete species turnover  while a difference between 1 to 1.4 SD units already correspond to half a change in species composition [@legendre_numerical_2012].

13. **Line 146: We should include species richness in your experimental approaches results**

Please refer to the reply to comment #6.

14. **Lines 195-196: The figures 2 and 3 could be switched to maintain the results consistency.**

Thanks for spotting.
Changed as requested.

15. **Lines 235: In the figure 5 should be included the year label of each panel.**

Thanks for spotting, they year labels should have been there in the first place.
Changed as requested.

16. **Lines 244: The results of species richness of the experimental approach may be included on the main text.**

Please refer to the reply to comments #6.

17. **Lines 262: Here, the author can include the name of dominant species to illustrate to readers into the ecological traits that increases which increases water availability in this study system. **

We have rewritten the sentence as follows:

>> Interestingly, only a few grass (e.g., *Aristida adscensionis*, *Eragrostis cilianensis*) and forb species (e.g., *Tiquilia paronychioides*) were in large parts responsible for this increase (see section 3.2).

# References

### 2020-08-10 text/ecography/round_1/enso_comp.Rmd
migrate gitlab repo to github

---
output:
  bookdown::pdf_document2:
    keep_tex: yes
    latex_engine: xelatex
    # float is needed for floating figures
    extra_dependencies: ["fontspec", "sectsty", "float"]
    toc: false
    fig_caption: yes
    number_sections: true
bibliography: transect_refs.bib
csl: ecography.csl
---

```{r setup, include=FALSE}
library("knitr")
library("kableExtra")
library("tidyverse")
library("here")
# make sure that figures are allowed to float (hold the assigned position and are not always put on top of a page)
knitr::opts_chunk$set(fig.pos = 'H')
```

# Introduction

![Caption for the picture.](/home/jannes/work/acad/science/projects/ecology/latin_america/peru/transect_piura/transect_17/figures/12_map_grid.png)

Studying the relationship between environmental gradients and species turnover has become a cornerstone of biogeography and community ecology since Humbold's and Bonpland's travels to South America more than 200 years ago [@macarthur_geographical_1984; @whittaker_gradient_1967]. 
Naturally, water availability drives floristic composition and biomass production in the (semi-)arid regions such as NW Peru [@muenchow_coupling_2013; @muenchow_woody_2013].
However, water availability does not only vary in space but also in time, especially in regions heavily affected by the El Niño Southern Oscillation (ENSO).
<!--
A super-humid El Niño (EN) episode benefits other species than a super-dry and fire-bringing La Niña (LN) episode [@richter_monitoring_2005].
-->

ENSO is a recurrent climate phenomenon causing anomalies around the globe [@capotondi_understanding_2015].
EN represents the warm and LN the cold phase of this cycle. 
ENSO neutral refers to periods when neither EN nor LN are present. 
LN usually causes the reverse effects of EN. 
The effects depend on the region and include among others excessive rainfalls, droughts as well as temperature, wind and monsoon anomalies. 
<!--
The anomalies show recurrent patterns to a certain degree but are still highly variable especially on a local scale. 
For instance, in 2016 the predicted Super-Niño rainfalls largely failed to materialize along the coasts of Ecuador and Peru.
Just a year later, unexpected, torrential rainfalls hit the same coasts during a supposedly ENSO neutral year with a magnitude comparable to that of the Super-Niño event in 1997/98. Originating from a so-called coastal EN, the 2017 rainfalls came as a complete surprise, and caused the unprepared population unbearable hardship [@fraser_peru_2017; @ramirez_understanding_2017].
-->

Though extreme climate events like EN and LN change and reshape ecosystems drastically [@Kogan.2017], ecological research mostly focusses on the effects of EN.
Nevertheless, the eco-climatic impact of LN is comparable in magnitude with that of EN.
For instance, LN episodes were most likely the main driver for the hiatus in global warming at the beginning of the 21^st^ century [@kosaka_recent_2013].
In any case, ENSO certainly drives a large amount of the interannual variability of woody plant growth across the tropics [@rifai_enso_2018] though the specific response to El Niño may vary from region to region [@phillips_changes_2009].
Overall, the effect of  extreme events is even pronounced in (semi)-arid ecosystems [@holmgren_extreme_2006], which is especially true for ENSO's terrestrial core region in NW Peru and SW Ecuador [@muenchow_woody_2013; @rollenbeck_climatic_2015] -- one of the most variable regions on Earth in terms of climate [@wehrden_global_2012].
This region is also home to the highly tropical dry forest (TDF) ecosystems.
TDFs are characterized by a high rate of endemism and feature an outstanding biodiversity which is almost as high as that of humid tropical forests [@espinosa_what_2011;@pennington_tropical_2018]. 
At the same time, they are the most threatened tropical ecosystem [@miles_global_2006]. 
In Peru, 95% of TDFs have already disappeared [@portillo-quintero_extent_2010]. 
Even so, tropical dry ecosystems are chronically understudied [@escribano-avila_biodiversity_2017;@muenchow_review_2018].

Species richness is strongly associated with precipitation in the tropics [@esquivel-muelbert_biogeographic_2017;@esquivel-muelbert_seasonal_2017].
This is especially true for dry tropical ecosystems where water shortage is the main constraint to plant growth.
Still, edaphic and topographic properties are sometimes equally important in structuring vegetation of tropical ecosystems [@laurance_influence_2010; @soethe_nutrient_2008; @condit_species_2013].
Yet, the corresponding effects on the floristic composition of tropical (semi-)arid vegetation communities remain scarcely investigated [@muenchow_soil_2013; @pena-claros_soil_2012;@ulrich_climate_2014]. 
This is especially true in terms of how soil properties influence vegetation communities as soon as water is no longer the limiting factor.
For instance, nitrogen may increase primary production even in the presence of droughts and grazing pressure [@kinugasa_increasing_2012; @whitford_effects_2011].

To tackle the outlined issues at least in part, we examined how different ENSO episodes affect the floristic composition along a spatio-temporal gradient in NW Peru. 
For this, we surveyed the floristic composition on permanent plots in four years representing different ENSO episodes (dry (2011) and humid (2012) LN year; moderate EN year (2016); very humid coastal EN year (2017)).
Specifically, we investigated the:

(1) change in the different levels of biodiversity, namely alpha, beta, gamma and ecosystem diversity. 
The latter was represented by ordination scores and additionally visualized by means of predictive mapping.
Overall, we expected a boost in biodiversity as soon as water was no longer a limiting factor.
Regarding the spatial distribution of the floristic composition we expected a more pronounced development of vegetation assemblages along the climatic gradient with wetter conditions [@engelbrecht_drought_2007].

(2) effects of soil characteristics and topography in shaping floristic composition and vegetation cover (biomass production). 
We expected that the influence of topographic and edaphic variables might gain in influence in wetter years.

(3) We complemented the plot survey at the landscape scale with an irrigation-fertilization experiment, thereby considering the socio-economic perspective of the ENSO variability. 
Specifically, we quantified the effect of the water-nutrient interaction on biomass development under different rainfall scenarios.
We expected that the water input of the rainfall scenario representing a Super-Niño event in combination with the fertilizer would be most conducive to biomass and biodiversity development.

# Methods

## Study area

\begin{figure}[H]
\includegraphics[width=25in]{/home/jannes/work/acad/science/projects/ecology/latin_america/peru/transect_piura/transect_17/figures/12_map_grid.png}
\caption{The study area.}
\label{fig:study-area-map}
\end{figure}

The study area stretches for 120 km along a pronounced bioclimatic gradient in NW Peru and features a desert along the Pacific coast in the west and TDFs near the Andean foothills in the east (Fig. \@ref(fig:study-area-map)). 
The climatic gradient is at the same time a gradient of increasing human influence especially in terms of agricultural activity towards the Andean foothills [@muenchow_coupling_2013; @muenchow_woody_2013].
Agricultural activity includes field farming and extensive grazing caused by free-ranging livestock.

The study area is part of the core region where the ENSO phenomenon exerts its greatest terrestrial influence [@richter_monitoring_2005]. 
Generally, EN brings more rain than usual and LN enforces the already dry conditions in the study area.
However, even this pattern is highly variable and can occasionally be reversed, see e.g., the years 2005, 2008 and 2012 in Fig. \@ref(fig:study-area-map)d. Additionally, so-called "local" or "coastal EN events" can take place as observed in 2017. 
Coastal EN events refer to the regional warming of coastal waters in the Gulf of Guayaquil in December as opposed to the Pacific basin-wide EN when warm North Australian waters reach the South American coast.

```{r, echo=FALSE, message=FALSE}
precip = readRDS(here("images/12_precip.rds"))
rain_17 = group_by(precip, station) %>% 
  # compute the median for each station
  mutate(median = median(precip)) %>% 
  # just keep coastal EN year
  filter(year == 2017)
```

Therefore, to put the ecological results into context, it is important to keep in mind the varying rainfall patterns as produced by different ENSO episodes across the studied years. 
The observed years differed largely in terms of precipitation.
2011 corresponded to a dry LN year and the whole study area experienced drier conditions than normal.
By contrast, 2012 was a humid LN year, and 1.5 times and 3 times more rain fell in Piura and Chulucanas, i.e., in the middle to eastern part of the study area (Fig. \@ref(fig:study-area-map)).
2016 was a moderate EN year and rainfall was double the median input in Paita and Piura, whereas in Chulucanas the median value was almost reached. 
2017 coincided with a coastal EN event and is the only studied year during which the whole study area experienced exceptionally wet conditions: Paita received almost `r rain_17[rain_17$station == "Paita", "dev"]  %>% pull %>% round` times (`r rain_17[rain_17$station == "Paita", "precip"] %>% pull %>% round` mm), Piura almost `r rain_17[rain_17$station == "Piura", "dev"]  %>% pull %>% round` times (`r rain_17[rain_17$station == "Piura", "precip"] %>% pull %>% round` mm) and Chulucanas `r rain_17[rain_17$station == "Chulucanas", "dev"]  %>% pull %>% round` times (`r rain_17[rain_17$station == "Chulucanas", "precip"] %>% pull %>% round` mm) more rain than on median (Fig. \@ref(fig:study-area-map)d). 
Moreover, it is noteworthy that vegetation development changes drastically in the study area during wet episodes but remains more or less uniform during neutral years [@holmgren_extreme_2006; @muenchow_coupling_2013; @richter_monitoring_2005].

## Data compilation

### Landscape and small-scale experiment vegetation sampling

We sampled 50 vegetation plots of 30 × 30 m^2^ size randomly along the climatic gradient between Paita and Chulucanas (Fig. \@ref(fig:study-area-map)).
To ensure that all parts of the study area were equally likely to be sampled, we stratified the sampling by distance to the Pacific Ocean in ten classes (see Appendix 1).
Prior to the sampling, we excluded riparian, agricultural and urban areas.
Each plot was located with a handheld Garmin GPS device in the field.
Field sampling took place towards the end of the vegetation period (March-May) in 2011, 2012, 2016 and 2017.
In each plot, we recorded all vascular plant species and determined their cover as portion of the plot area in percentage points. 
Specifically, plant cover refers to the relative projected area covered by a species [@damgaard_estimating_2014].
Prior to the analysis, we transformed the plant cover values in accordance with the decimal scale by Londo [@londo_decimal_1976].
The nomenclature follows the conventions of the Tropicos online database [@missouribotanicalgarden_tropicos_2018].

In addition to the plot survey at the landscape scale, we conducted a small-scale irrigation-nutrient experiment to quantify the beneficial effects of the water-nutrient interaction on vegetation development and biomass production.
We established the experiment in the middle of the precipitation gradient, ranging from `r rain_17[rain_17$station == "Paita", "median"] %>% pull %>% round` mm (Paita; annual median precipitation) to `r rain_17[rain_17$station == "Chulucanas", "median"] %>% pull %>% round` mm (Chulucanas; Fig. \@ref(fig:study-area-map)), on the campus of the University of Piura (5°10'S, 80°38'W, see Piura in Fig. \@ref(fig:study-area-map)).
The environmental conditions are typical for the region in terms of climate, topography and soil properties (see @muenchow_soil_2013; @muenchow_woody_2013). 
Inside a sandy area devoid of woody vegetation, we placed 12 experimental plots of 3 × 3 m^2^, which were regularly spaced along a rectangular grid of 1 ha. 
Plots were assigned to one of three irrigation treatments: The first treatment represented a Super EN year (example of 1997/98: 1780 mm), the second represented a moderate EN event (example of 1991/1992: 258 mm). 
The last treatment represented the baseline with no additional water input and corresponded to the year in which the irrigation experiment was conducted (2013) which was a neutral year (54 mm; see also Fig. \@ref(fig:study-area-map)d). 
The irrigation water stemmed from rain-collecting tanks.
In addition to the water treatment, two randomly chosen plots of each irrigation treatment received a fertilization treatment of 200kg/ha granular nitrogen fertilizer (NH~4~NO~3~). 
The experiment was carried out between December 2012 and May 2013.
Plant development was monitored two times per month and included the determination of plant species [@missouribotanicalgarden_tropicos_2018] and corresponding coverage (%).
Please refer to Appendix 2 for a detailed description of the irrigation-nutrient experiment.

### Environmental variables

We randomly took and mixed three soil samples from 15cm and 30cm depth in each plot in 2011. 
Soil composition should not vary too much between the years in which vegetation sampling took place since the study area largely features arenosols, i.e., a very weekly developed soil type.
On the other hand, soil chemistry can vary largely between two years.
For instance, El Niño events can increase soil respiration, i.e., soil CO\textsubscript{2} emissions, by a factor of 100, especially in the vicinity of trees [@salazar_tree_2019].
Subsequently, we measured in the laboratory variables known to be important for vegetation development: pH, electrical conductivity (µS), carbon-to-nitrogen ratio, soil texture (%), the concentrations of the cations Ca, Mg, K and Na (cmol kg^-1^), and skeletal content (%; measured as gravimetric proportion of stones \>2 mm; see [@muenchow_woody_2013] for a more detailed description).

We computed the topographic variables aspect, altitude, catchment slope, catchment area, soil wetness index, plan and vertical curvature from a digital elevation model with a 30 × 30 m^2^ resolution [@landprocessesdistributedactivearchivecenter_lp_2018].
We downloaded the normalized difference vegetation index (NDVI) raster imagery from the Moderate Resolution Imaging Spectrometer (MODIS) available for the vegetation period (March, April) for each studied year (2011, 2012, 2016, 2017) at a 250 × 250 m^2^ resolution for our study area.
Since three NDVI scenes were available for each vegetation period, we computed the mean NDVI for each vegetation period.

The interpolated precipitation values from the WorldClim data provided a spatial distribution of the average annual rainfall [@hijmans_very_2005].
Local precipitation values were collected from automated climate stations (Fig. \@ref(fig:study-area-map)d) operated by the University of Piura near Paita (1997-2018), in Piura (1991-2018) and in Chulucanas (1997-2018).

## Data analysis

### Biodiversity and plant cover analysis

We calculated four measures of biodiversity for each observed year, namely alpha, beta, gamma and ecosystem diversity [@whittaker_gradient_1967].
Alpha diversity refers to the mean number of species per plot.
Beta diversity is expressed as the range of the first Detrended Correspondence Analysis (DCA) axis (see below). 
This measure is given in units of standard deviation (SD) with four SDs corresponding to a complete species turnover.
Gamma diversity is the total number of observed species across all plots.
Ecosystem diversity was assessed with the help of a DCA [@hill_detrended_1980] to which the decimal scale transformed cover values of the species-plot matrix were subjected [@londo_decimal_1976]. 
DCA tries to condense as much as possible the observed variance of the input matrix in a low-dimensional space (usually 2-3 axes), thereby extracting the main apparent gradients.

For presenting the results of our nutrient-irrigation experiment we simply used descriptive statistics by plotting the increase in plant cover as well as water input against time for the duration of the experiment.

### Modeling and predictive mapping

<!-- JM: say, the nth aim in correspondence with the hypotheses made at the beginning -->
One aim of this study was to spatio-temporally predict ecosystem diversity (floristic composition) represented by ordination scores along the climatic gradient in the study area.
@muenchow_coupling_2013 have already shown that NDVI is an excellent predictor for vegetation assemblages in the study area.
Therefore, we modeled the scores of the first DCA axis representing the main observed floristic gradient in the study area with the help of a Generalized Additive Model (GAM) as a function of the interaction between NDVI and year: 
$$
\begin{aligned}
\text{Scores}_{i}\ \sim\ \mathrm{N}\left( \mu_{i},\ \sigma^{2} \right) \\
\mu_{i} = \alpha + f_{\text{year}_{i}}\left( \text{NDVI}_{i} \right) + \beta_{1}\text{year}_{i}
\end{aligned}
$$

(Formula 1)

where *i* refers to the *i*th observation, *$\alpha$* is the intercept, *$\beta$*~1~ is the estimated coefficient for year and *f* is a spline smoother for the NDVI:year interaction.

The sole purpose of the model is the spatial prediction of the ordination scores, not statistical inference or the interpretation of the models' coefficients.
We will use variation partitioning later on to assess how the influence of environmental predictors has changed over the studied years (see section \@ref(impact-of-environmental-variables)). 
Consequently, here we are only interested in a good predictive performance of the model to make sure that it is able to generalize, i.e., that it is able to predict reasonably well unseen data [@james_introduction_2013; @kuhn_applied_2013]. 
Only this can be the basis for a reliable map of the predicted ordination scores.
To make sure this is the case, we used spatial cross-validation.
Overall, cross-validation accounts for over-optimistic predictions on the training set by randomly splitting a data set into partitions used for training and testing [@brenning_spatial_2012;@lovelace_geocomputation_2019a]. 
Spatial cross-validation is an extension that measures a model's ability to spatially predict the response variable in the presence of spatial autocorrelation [@schratz_hyperparameter_2019]. 
We ran a 100-repeated 5-fold spatial cross-validation in which the partitioning is based on *k*-means clustering (*k* = 5; [@russ_data_2010]).
We also made sure that all annual observations of one plot were jointly placed in either the test or the training data set [@meyer_improving_2018]. 
We used the normalized root mean-square error (NRMSE) in percent as performance measure for the spatial cross-validation.

$$
NRMSE = \frac{\sqrt{\frac{\sum_{i = 1}^{n}{(e_{i} - y_{i})²}}{n}}}{max(y) - \ min(y)}*100
$$ 
(Formula 2)

where *n* is the number of observations, *e* corresponds to the residuals, and *y* refers to the observed values.

### Impact of environmental variables

<!-- JM: add a first sentence, the nth aim of the study was to ... in correspondence with the posed hypotheses -->
Variation partitioning quantifies the variation explained of the response, in our case the first two DCA axes, by one predictor group, in our case edaphic variables, while controlling for the effect of another predictor group [@borcard_numerical_2018; @peres-neto_variation_2006], in our case topographic variables.
Internally, variation partitioning computes a Redundancy Analysis (RDA) separately for each predictor group and for all predictor groups together, and subesequently quantifies how much each predictor group explains alone and jointly with the other predictor group of the variation of the response.
This is frequently displayed with the help of a Venn diagramm.
We ran the variation partitioning for each studied year (2011, 2012, 2016, 2017) to assess how edaphic and topographic variables influence vegetation composition under increasingly humid conditions. 
Prior to the variation partitioning, we log-transformed electrical conductivity, the CN ratio and the catchment area.
Subsequently, we excluded all collinear variables which exceeded a variance inflation factor \>3 leaving us with seven edaphic and eight topographic variables. 
We refrained from a forward selection of explanatory variables [@blanchet_forward_2008] because (i) we only used a reasonable number of meaningful, non-collinear variables that (ii) were available for all four years thus guaranteeing inter-annual comparability.

All statistical analyses were conducted in the open source software R [@rcoreteam_language_2019] using its packages mgcv [@wood_generalized_2017], RQGIS ([@muenchow_rqgis_2017a], sperrorest [@brenning_spatial_2012] and vegan [@oksanen_vegan_2019]. 
<!-- JM: add link to github repository, zenodo repository, dockerhub -->
To comply with the good practices of reproducibility we will make available the data and the code of this work on an open accessible github repository (https://github.com/) upon acceptance of the manuscript.

# Results

## Biodiversity and plant cover

```{r desc-stats, echo=FALSE}
div = readRDS(here("images/13_desc_stats_tab.rds"))
cap = paste("Different biodiversity measures across the studied years.",
            "min = minimum observed number of species per plot.",
            "max = maximum observed number of species per plot.")
kable(div, "latex", booktabs = TRUE, escape = TRUE, 
      # align = "c",
      # col.names = linebreak(nms, align = "c"),
      caption = cap)
```

Mean alpha diversity was lowest during the dry LN (2011) and highest during the moderate EN (2016; Table \@ref(tab:desc-stats)).
Compared to the humid LN (2012) and the moderate EN (2016), species richness was lower in the east of the study area during the very humid neutral year(2017), however, higher within a distance of 20-25 km to the coast (left panel of Fig. \@ref(fig:spri)). 
Species turnover (beta diversity), expressed as the range of the first DCA axis (see also Fig. \@ref(fig:dca)), was also lowest during the dry LN (2011), slightly increased in 2012 (humid LN) and 2016 (moderate EN) and reached its highest value in 2017 (very humid neutral year).
Gamma diversity increased from 2011 (dry LN) to 2012 (humid LN), remained almost the same in 2016 (moderate EN), and decreased in 2017 (very humid neutral year; Table \@ref(tab:desc-stats)).
By contrast, the latter showed by far the highest plant coverage along the entire gradient (right panel of Fig. \@ref(fig:spri) & section \@ref(spatio-temporal-mapping)).

```{r dca, echo=FALSE, fig.cap="Plotting the scores of the second DCA axis against the scores of the first DCA axis. To better visualize the cross-annual changes, the scores of all years were added in light gray to the background of each panel. The black dots refer to the scores of the year given in the strip of each panel. Numbers above the dots are the plot IDs which were ordered in accordance with the distance to the sea, i.e. plot number 1 is closest and plot number 50 farthest away from the sea.", out.width="70%", fig.align="center"}

knitr::include_graphics(here("figures/14_londo_dca.png"))
```

The first two DCA axes explained 68% of the observed variance with the first axis contributing alone 48%. 
The first axis is mainly associated with precipitation while the second axis is related to topographic variables such as vertical curvature and catchment area.
Overall, the DCA scatter cloud becomes more compressed over the years, and the plots become increasingly positioned along the first axis (Fig. \@ref(fig:dca)). 
Put differently, the importance of the second axis steadily decreased from 3.22 during the dry LN (2011) to 1.35 units of standard deviation during the very humid neutral year (2017).
By contrast, the first axis was more important in all phases which were wetter than the dry LN (Table \@ref(tab:desc-stats)).

```{r spri, echo=FALSE, out.width="80%", fig.align="center", fig.cap="Species richness (left panel) and plant cover (right panel, in \\%) along the spatio-temporal gradient. To aid visual interpretation, loess smoothers were added."}
include_graphics(here("figures/13_spri_cover.png"))
```

## Spatio-temporal mapping

Our model explained 76% of the observed variance with a spline smoother absorbing 3 *df* (Formula 1 & section \@ref(modeling-and-predictive-mapping)).
The spatially cross-validated NRMSE also indicated a satisfying fit of 27% (SD 8.53% over 100 repetitions).

```{r pred-map, echo=FALSE, out.width="80%", fig.align="center", fig.cap="Spatial prediction of the floristic composition along the cross-section for different ENSO episodes The urban boundaries of the cities Paita, Piura and Chulucanas are displayed in black. The Piura River and its tributaries are represented by light blue lines with a black contour.", fig.align="centerfig.align=fig.align="}
include_graphics(here("figures/15_pred_map.png"))
```

A clear trend towards more pronounced differences in vegetation formations became apparent between the studied years (Fig. \@ref(fig:pred-map)). 
The dry LN (2011) showed a floristic composition that was almost uniform across great parts of the study area (Panel 2011 of Fig. \@ref(fig:pred-map)).
This changed during the wet LN (2012) where three distinct vegetation formations developed (Panel 2012 of Fig. \@ref(fig:pred-map)). 
In the west, we observed a semi- to full desert with a sparse herb and grass cover, which at times was accompanied by isolated trees and bushes.
This formation developed into an open xeric shrubland (green/yellow pixels in Fig. \@ref(fig:pred-map)) where *Acacia macracantha* and *Encelia canescens* were the dominating species.
Further east, the shrubland turned into a TDF formation (orange/red pixels in Fig. \@ref(fig:pred-map)) with *Cordia lutea*, *Prosopis pallida*, *Cenchrus echinatus* and *Antephora hermaphrodita* being the most common species. 
The differentiation of these three vegetation formations was even more pronounced during the moderate EN (Panel 2016 of Fig. \@ref(fig:pred-map)). 
This trend was both amplified and disrupted during the very humid neutral year (2017) during which a dominant grass-herb formation stretched from the coast until far behind the city of Piura (dark blue pixels in Panel 2017 of Fig. \@ref(fig:pred-map)). 
TDF formations were also less dominant (orange to red pixels).
Though species composition remained roughly the same during the very humid neutral year (2017) compared to previous years, plant cover has increased drastically, and particularly in the western part of the study area where highest coverage values were observed for the grasses *Aristida adscensionis* and *Eragrostis cilianensis* and the seedlings of the shrub *Encelia canescens*.
Among herbal species, *Crotalaria incana*, *Exodeconus maritimus* and *Tiquilia paronychioides* showed the highest cover values.

## Influence of environmental variables

The influence of edaphic and topographic variables on floristic composition was relatively small in the driest year (2011) with 25% and 28% of explained variance, respectively (Fig. \@ref(fig:varpart)). 
The explained variance roughly doubled in all other years, however, did not further increase with an excessive supply of water in the most humid year (2017; Fig. \@ref(fig:varpart)).

```{r varpart, echo=FALSE, out.width="50%", fig.align="center", fig.cap="Fractions explaining the floristic composition represented by the first two DCA axes in accordance with a RDA variation partitioning. Numbers within the circles represent the adjusted $R^2$. Significance levels of the pure fractions are given in the soil and topography circles as . ($<0.1$), * ($<0.05$), ** ($ <0.01$), and *** ($<0.001$)."}
include_graphics(here("figures/16_varpart.png"))
```

## Irrigation-fertilization experiment

The experiment revealed that edaphic variables are conducive to vegetation development (see also Appendix 3). 
The fertilized plots representing moderate EN years exhibited a slightly greater plant coverage than in their unfertilized counterparts (middle panel of Fig. \@ref(fig:irrigation)). 
Adding even more water revealed the true beneficial effects of fertilizing.
The fertilized plots simulating Super EN conditions showed a threefold increase in plant cover compared to their unfertilized counterparts (right panel of Fig. \@ref(fig:irrigation)).
Nevertheless, there was an upper limit regarding the beneficial effects of nutrient addition. 
Biomass production stopped increasing or even decreased in the fertilized Super-EN plot after 8 March 2013, which corresponded to 1487 mm of irrigation and 200 kg ha^-1^ of NH~4~NO~3~.

```{r irrigation, echo=FALSE, out.width="80%", fig.align="center", fig.cap="Results of the irrigation-nutrient experiment. The dashed line refers to the water input, note that the left y-axis was log-transformed. The solid lines represent the observed plant coverage in \\% (right y-axis)."}
include_graphics(here("figures/17_nutrient_irrigation_experiment.png"))
```

# Discussion

## Biodiversity and biomass

Beta diversity (expressed as the range of the first DCA axis), increased with wetter conditions (Fig. \@ref(fig:dca) & Table \@ref(tab:desc-stats)).
Naturally, higher species turnover was mainly driven by short-lived, less abundant species.
Concordantly, @cleland_sensitivity_2013 reports similar findings from temperate grasslands.
The change of the shape of the DCA point scatter between the relatively humid years (panels 2012 & 2016 of Fig. \@ref(fig:dca)) and the very humid neutral year (panel 2017 of Fig. \@ref(fig:dca)) can be most likely attributed to the increase of plant cover (see also next paragraph).

Alpha and gamma diversity increased with higher seasonal precipitation amounts. 
The relationship, however, is non-linear, i.e., the increase in diversity reached a plateau with the rainfall values recorded under humid LN (2012) and under moderate EN (2016) conditions. 
Then even more water input as observed during the very humid neutral year (2017) resulted in a slightly decreased diversity but also in a massive and sudden increase of overall plant cover due to higher biomass production.
Interestingly, only a few grass and forb species were in large parts responsible for this increase.
This is in contrast to other dryland studies which showed that dormant and annual species increased biodiversity under more humid conditions [@gutierrez_variation_2000; @holmgren_extreme_2006].
On the other hand, this finding is in line with the theory that under favorable environmental conditions a limited amount of species become the dominant actors of an ecosystem, whereas under intermediate levels of stress and disturbance a higher species richness is supported [@brown_why_2014]. 
Though obviously a few dominant species outcompeted other species in 2017, they still followed a ruderal strategy by growing as fast and producing as much offspring as possible in the short period of favorable conditions [@schmidtlein_mapping_2012].
Finally, free-ranging livestock might also have an impact on vegetation dynamics in the study area, especially in its eastern parts with more than 100mm of annual precipitation.
Livestock might alter floristic composition and diversity while supporting especially plants adapted to grazing.

Aside from the spectacular germination of annuals, perennial woody plants also largely benefit from wetter periods, e.g. through increased flower and fruit production [@holmgren_nino_2001]. 
However, we only recorded plant cover, where the corresponding interannual difference is only marginal compared to previous years. 
Still, rainy periods are very conducive to the regeneration of woody vegetation in (semi-)arid areas [@bowers_demographic_1997; @brown_reorganization_1997; @sitters_rainfalltuned_2012].
For example, the growth and recruitment rate of *Prosopis pallida* is almost twice as high during wet EN years than in neutral or dry years in northern Peru [@lopez_climatic_2006].

Overall, water surplus converts the (semi-)arid landscape interspersed with shrubs and trees temporarily into a tree savannah.
Shallow underground water near the Pacific coast is largely responsible for the occurrence of perennial species in the study area.
In fact, dendrochronological studies have shown that the growth of the most frequent tree species in the study area, *Prosopis pallida*, is nearly independent of the interannual climatic variability because frequently the water input of the low annual precipitation and nearby rivers suffices to replenish the groundwater [@brown_water_1990; @salazar_effect_2018; @throop_response_2012].

Though we largely focussed on how current-year precipitation affects biodiversity dynamics, previous year rainfall might also have an effect [@tenhumberg_timelagged_2018].
For instance, @dudney_lagging_2017 observed an increase in grass and a decrease in forb abundance when rainfall was high one year earlier.
Similarly, @walter_plants_2011 suggest that grasses might have a 'drought-memory'.
Time-lagged responses can be also observed in woody species.
For example, tree ring growth of *Prosopis pallida* is highly correlated with rainfall of the previous year [@salazar_effect_2018].

## Floristic gradient mapping across ENSO episodes

The spatial prediction map highlights that specific vegetation formations become more pronounced with increasing water input. 
In the driest year (2011), a rather uniform distribution of the same floristic composition (green pixels in Fig. \@ref(fig:pred-map)) developed along large parts of the study area. 
The corresponding DCA scores close to 0 indicates that plant community composition was primarily made up of species that are largely independent of the climatic gradient [@hill_detrended_1980; @vonwehrden_pluralism_2009].
This is because in 2011 rainfall barely occurred along large parts of the study area, and therefore especially perennial species were recorded which can and must endure drought years [@salazar_effect_2018].
By contrast, during the humid LN (2012) and the moderate EN (2016) the same three distinct vegetation formations developed along the gradient. 
This can be explained by similar rainfall patterns though 2012 was a humid LN and 2016 a moderate EN year. 
Finally, a large water surplus during the very humid neutral year (2017) led to the disruption of the strictly ordered formations along the humidity gradient. 
On the one hand, dense grass and herb cover dominated large parts of the study area. The increased cover dominance of annual species is visible in larger negative DCA score values. 
On the other hand, vegetation patches (yellow, green, and orange pixels) can be found near the coast which are otherwise more typical of the eastern part of the gradient (section \@ref(spatio-temporal-mapping) & Fig. \@ref(fig:pred-map)).
TDF formations (orange and red pixels) are less visible in 2017 because the dense grass/herb cover temporarily converted the landscape into a savannah-like vegetation formation [@espinosa_what_2011; @salazar_effect_2018].
Another interpretation could be also that the NDVI values, our main predictor for the floristic composition, is similar for the dense cover of annual plants and perennial woody plants in 2017 [@pena_assessing_2015; @salazar_intraspecific_2018].

## Changing impact of environmental variables with wetter conditions

<!-- 
Interpreting the ordination axes is interesting, however, slighly deviating from the main purpose of the paper.
Therefore, I have commented this part.

The first ordination axis represents the main floristic gradient in the study area (Fig. \@ref(fig:dca)) and is associated with the increasing amount of precipitation from east to west (see section \@ref(biodiversity-and-plant-cover)). 
The second DCA axis is associated with topographic variables (section \@ref(biodiversity-and-plant-cover); see also @muenchow_woody_2013 and @muenchow_coupling_2013). 
Topographic heterogeneity was most important during the dry LN (panel 2011 of Fig. \@ref(fig:dca)) and became less important in more humid years as evidenced by the concentration of the points along the first axis (panels 2012, 2016 & 2017 of Fig. \@ref(fig:dca)).
-->

Variation partitioning clearly reveals that environmental predictors (topography, soil) gain in importance for explaining plant diversity and productivity dynamics as soon as water is no longer restricting plant growth (section \@ref(influence-of-environmental-variables) & Fig. \@ref(fig:varpart)).
Interestingly, the shared explained variance in the floristic dataset is roughly the same for the years experiencing wetter conditions (humid LN (2012), moderate EN (2016) and the very humid neutral year (2017; coastal EN)) though 2017 was the most humid by far. 
This indicates that there is an upper limit to the beneficial effects of the water-soil and water-topography interaction.
Topographic variables such as curvature and catchment area play some role in structuring the vegetation composition in the study area especially in dry years [@muenchow_coupling_2013; @muenchow_woody_2013]. 
Similar findings could be also observed in Mexico where topography also partly explained the composition of dry forests [@mendez-toribio_effects_2016].
However, the most important topographic variable in the study area was altitude since it is highly correlated with mean annual precipitation. 
Of course, edaphic variables such as pH are also related to precipitation.
The influence of edaphic variables on vegetation composition could even increase with more developed soils.
However, large parts of the study area consist of sandy soils (arenosols). 
Only in the east of the study area, soils show more signs of humification and a brownish color due to iron oxide release from primary minerals.

Naturally, nutrients can play a major role in drylands [@ronnenberg_effects_2011; @whitford_effects_2011] as confirmed by our irrigation-nutrient experiment.
In fact, the water-nitrogen interaction can result in a biomass production three times as high as the production achieved when adding the same amount of water but without additional nutrients (Fig. \@ref(fig:irrigation)).
This is of utmost importance for sustainable yet productive agricultural management. 
In combination with a reliable climate prediction, it supports the decision on which crops (annual herbal and grass species) to grow and how much fertilizer to use.
Other studies have also confirmed the importance of soil for the composition of the vegetation in drylands.
For example, @ulrich_climate_2014 found that soil fertility and variability in rainfall and temperature are the most important predictors for beta diversity.
Similarly, soil plays an important role in Colombian dry forests [@gonzalez-m_disentangling_2018] 
Finally, @espinosa_what_2011 found out that soil moisture, soil temperature and nitrogen explained in large parts the floristic composition of Tumbesian dry forests, which is quite close to our study area.

## Climate change and opportunities for restoration and sustainable agrarian management

Climate change will most likely increase the frequency and intensity of El Niño events [@cai_increasing_2014; @IPCC.2014; @wang_continued_2017].
This and climate change in general pose the risk to turn tropical forests - dry and moist - from carbon sinks to sources [@phillips_changes_2009;@alvarez-davila_forest_2017;@salazar_tree_2019]. 
This is also true for the study area [@salazar_tree_2019] where EN is frequently associated with higher rainfall and LN with increased aridity.
However, both can also cause the opposite effects. 
For instance, 2005 was a very dry EN year, 2012 was a humid LN year, and 2017 was first predicted to be a LN year and then turned out to be a very wet neutral year (coastal EN; see Fig. \@ref(fig:study-area-map)d for many more examples). 
Hence, deviations from the mean expectation were the norm for the study area in the last two decades.
However, this is only true when solely relying on the popular ONI index -- an index, which measures sea surface temperature anomalies in the central Pacific. 
Naturally, local anomalies possibly overlap, dilute and modify the ENSO signal stemming from the central Pacific. 
Studies have shown that indices computed from the EN region 1+2, which corresponds to an area in the tropical Pacific close to the South American coast, already provide better correlations with climate anomalies in northern Peru and southern Ecuador [@kao_contrasting_2009;@salazar_effect_2018]. 
At least two further local indexes seem to be also promising for predicting climate anomalies along the Peruvian coast.
The first is the Índice Costero El Niño (ICEN; see http://www.met.igp.gob.pe/publicaciones/2014/ElNino_v2_27_03_2014_7.pdf), and the second is an ENSO index for the insurance of agricultural products [@mortensen_enso_2018].

In any case, constant monitoring is a prerequisite for a better understanding of ENSO-affected ecosystems. 
Since ENSO exerts its most direct influence on tropical regions, and all corresponding ecosystems are endangered without exception, the correct prediction of the expected ENSO effects are a cornerstone of informed conservation and reforestation management. 
Conserving and reversing the state of degraded ecosystems also helps to mitigate the effects of floodings and landslides due to excessive rainfalls, and thereby contributes to protect the population and infrastructure from unnecessary harm [@fraser_peru_2017; @ramirez_understanding_2017].
Currently, the regional government has set up eleven restoration projects in northern Peru which are trying to protect the endemic vegetation and increase vegetation cover through reforestation.
Additionally, NGOs and mining companies are also working on reforestation projects.
Most of these projects rely on extreme events (like 2017 EN) to boost plant growth to overcome the otherwise very limited access to water.
Our experimental results indicated that water availability showed a non-linear relationship with biodiversity and plant cover.
Therefore, restoration projects might consider water regulation in combination with a sensible amount of fertilization [@dahlsjo_local_2019].
The same applies to a sustainable agrarian management.
Indeed, agricultural development has been very intense over the last 20 years in the study area.
Personal communications with the main local companies indicated that extreme events are an important issue.
On the one hand, El Niño has the potential to boost crop productivity.
On the other hand, especially short and very intense rainfall might create havoc during the crop season as was the case during the 2017 EN.
Additionally, our results clearly reveal patterns of the spatial variability of biomass production in response to water availability along the climatic gradient, which might be also an important information for agrarian activities.

# Conclusions

<!--
EN and LN episodes are highly variable in terms of frequency, intensity and impact in ENSO's terrestrial core region.
Climate change is believed to even increase the frequency of extreme EN and LN events (IPCC 2014, Cai et al. 2014, Wang et al. 2017).
Failing to predict the occurrence of such extreme events causes the local population unbearable hardship as was the case along the Ecuadorian and Peruvian coast in 2017 [@fraser_peru_2017; @ramirez_understanding_2017].
Reforestation and renaturation are among the most effective countermeasures to prevent and mitigate hardship caused by flooding and catastrophic mass movements since intact ecosystems can store larger quantities of water and additionally contribute to increased slope stability.
Moreover, our study has shown that the nitrogen-water interaction has increased biomass production manifold, a result which is highly valuable for sustainable agrarian management.
However, both agrarian and effective conservation management depend on reliable climatic predictions, especially regarding the effect and intensity of the ENSO episode.
For these predictions, authorities, stakeholders and conservationists should not solely rely on the ONI index but also consider local indices.
Overall, it is time to overthink our sometimes overly simplistic understanding of ENSO (Richter et al. 2018).
-->

# References {#references .ListParagraph}

### 2020-08-10 text/ecography/round_2/enso_comp.Rmd
migrate gitlab repo to github

--- 2020-08-10 text/ecography/round_1/enso_comp.Rmd
@@
     number_sections: true
+header-includes:
+  # adjust figure captions
+  - \usepackage{caption}
+  - \captionsetup{font={footnotesize,bf,singlespacing}}
+  # Ecography requires double line spacing and line numbering
+  - \usepackage{setspace}
+  - \doublespacing
+  - \usepackage{lineno}
+  - \linenumbers
+
+title: "Monitoring and predictive mapping of floristic biodiversity along a climatic gradient in ENSO's terrestrial core region, NW Peru"
 bibliography: transect_refs.bib
 csl: ecography.csl
+# biblio-style: "natbib"
+link-citations: true
 ---
@@
 
-![Caption for the picture.](/home/jannes/work/acad/science/projects/ecology/latin_america/peru/transect_piura/transect_17/figures/12_map_grid.png)
-
 Studying the relationship between environmental gradients and species turnover has become a cornerstone of biogeography and community ecology since Humbold's and Bonpland's travels to South America more than 200 years ago [@macarthur_geographical_1984; @whittaker_gradient_1967].
@@
 ENSO is a recurrent climate phenomenon causing anomalies around the globe [@capotondi_understanding_2015].
-EN represents the warm and LN the cold phase of this cycle. 
-ENSO neutral refers to periods when neither EN nor LN are present. 
-LN usually causes the reverse effects of EN. 
+El Niño represents the warm and La Niña the cold phase of this cycle. 
+ENSO neutral refers to periods when neither El Niño nor La Niña are present. 
+La Niña usually causes the reverse effects of El Niño. 
 The effects depend on the region and include among others excessive rainfalls, droughts as well as temperature, wind and monsoon anomalies. 
@@
 
-Though extreme climate events like EN and LN change and reshape ecosystems drastically [@Kogan.2017], ecological research mostly focusses on the effects of EN.
-Nevertheless, the eco-climatic impact of LN is comparable in magnitude with that of EN.
-For instance, LN episodes were most likely the main driver for the hiatus in global warming at the beginning of the 21^st^ century [@kosaka_recent_2013].
+Though extreme climate events like El Niño and La Niña change and reshape ecosystems drastically [@Kogan.2017], ecological research mostly focusses on the effects of El Niño.
+Nevertheless, the eco-climatic impact of La Niña is comparable in magnitude with that of El Niño.
+For instance, La Niña episodes were most likely the main driver for the hiatus in global warming at the beginning of the 21^st^ century [@kosaka_recent_2013].
 In any case, ENSO certainly drives a large amount of the interannual variability of woody plant growth across the tropics [@rifai_enso_2018] though the specific response to El Niño may vary from region to region [@phillips_changes_2009].
@@
 To tackle the outlined issues at least in part, we examined how different ENSO episodes affect the floristic composition along a spatio-temporal gradient in NW Peru. 
-For this, we surveyed the floristic composition on permanent plots in four years representing different ENSO episodes (dry (2011) and humid (2012) LN year; moderate EN year (2016); very humid coastal EN year (2017)).
+For this, we surveyed the floristic composition on permanent plots in four years representing different ENSO episodes (dry (2011) and humid (2012) La Niña year; moderate El Niño year (2016); very humid coastal El Niño year (2017)).
 Specifically, we investigated the:
@@
 
-\begin{figure}[H]
-\includegraphics[width=25in]{/home/jannes/work/acad/science/projects/ecology/latin_america/peru/transect_piura/transect_17/figures/12_map_grid.png}
-\caption{The study area.}
-\label{fig:study-area-map}
-\end{figure}
+```{r study-area-map, echo=FALSE, fig.cap="The study area (Panel A) located in Peru (Panel B) given as a false-color composite representing the change in NDVI across the observed years.
+The 2011 NDVI serves as the basis for the interannual change (Panel C).
+Green pixels occur where the interannual difference in NDVI was greatest between 2017 and 2011.
+Blue color tones indicate that NDVI differences were greatest between 2012 and 2011 (red: between 2016 and 2011).
+Panel D: Ratio between annual precipitation and median precipitation (median Paita: 25.9mm, median Piura: 71.4mm, median Chulucanas: 272.4mm) measured at the three automated climatic stations (see section \\\\@ref(environmental-variables)).
+In case the precipitation was smaller than the median, we computed the ratio as: 1 / (precipitation / median(precipitation)) \\* -1.
+The years in which vegetation sampling took place are printed in bold (x-axis).
+The ENSO phases are in correspondence with the ONI index of the NOAA Climatic Prediction Center (http://origin.cpc.ncep.noaa.gov/products/analysis\\_monitoring/ensostuff/ONI\\_v5.php).
+We have intentionally left out the 1997/1998 Super-Niño event since this would have dominated the y-axis."}
+# knitr::include_graphics(here("figures/12_map_grid.png"))
+suppressPackageStartupMessages(library("raster"))
+library("png")
+back = as.raster(readPNG(here("figures/12_ternary_background.png")))
+front = as.raster(readPNG(here("figures/12_ternary_foreground.png")))
+p = readRDS(here("figures/12_map_grid.rds"))
+p
+```
 
@@
 The study area is part of the core region where the ENSO phenomenon exerts its greatest terrestrial influence [@richter_monitoring_2005]. 
-Generally, EN brings more rain than usual and LN enforces the already dry conditions in the study area.
+Generally, El Niño brings more rain than usual and La Niña enforces the already dry conditions in the study area.
 However, even this pattern is highly variable and can occasionally be reversed, see e.g., the years 2005, 2008 and 2012 in Fig.
 \@ref(fig:study-area-map)d.
-Additionally, so-called "local" or "coastal EN events" can take place as observed in 2017.
-Coastal EN events refer to the regional warming of coastal waters in the Gulf of Guayaquil in December as opposed to the Pacific basin-wide EN when warm North Australian waters reach the South American coast.
+Additionally, so-called "local" or "coastal El Niño events" can take place as observed in 2017.
+Coastal El Niño events refer to the regional warming of coastal waters in the Gulf of Guayaquil in December as opposed to the Pacific basin-wide El Niño when warm North Australian waters reach the South American coast.
 
@@
 The observed years differed largely in terms of precipitation.
-2011 corresponded to a dry LN year and the whole study area experienced drier conditions than normal.
-By contrast, 2012 was a humid LN year, and 1.5 times and 3 times more rain fell in Piura and Chulucanas, i.e., in the middle to eastern part of the study area (Fig. \@ref(fig:study-area-map)).
-2016 was a moderate EN year and rainfall was double the median input in Paita and Piura, whereas in Chulucanas the median value was almost reached. 
-2017 coincided with a coastal EN event and is the only studied year during which the whole study area experienced exceptionally wet conditions: Paita received almost `r rain_17[rain_17$station == "Paita", "dev"]  %>% pull %>% round` times (`r rain_17[rain_17$station == "Paita", "precip"] %>% pull %>% round` mm), Piura almost `r rain_17[rain_17$station == "Piura", "dev"]  %>% pull %>% round` times (`r rain_17[rain_17$station == "Piura", "precip"] %>% pull %>% round` mm) and Chulucanas `r rain_17[rain_17$station == "Chulucanas", "dev"]  %>% pull %>% round` times (`r rain_17[rain_17$station == "Chulucanas", "precip"] %>% pull %>% round` mm) more rain than on median (Fig.
+2011 corresponded to a dry La Niña year and the whole study area experienced drier conditions than normal.
+By contrast, 2012 was a humid La Niña year, and 1.5 times and 3 times more rain fell in Piura and Chulucanas, i.e., in the middle to eastern part of the study area (Fig. \@ref(fig:study-area-map)).
+2016 was a moderate El Niño year and rainfall was double the median input in Paita and Piura, whereas in Chulucanas the median value was almost reached. 
+2017 coincided with a coastal El Niño event and is the only studied year during which the whole study area experienced exceptionally wet conditions: Paita received almost `r rain_17[rain_17$station == "Paita", "dev"]  %>% pull %>% round` times (`r rain_17[rain_17$station == "Paita", "precip"] %>% pull %>% round` mm), Piura almost `r rain_17[rain_17$station == "Piura", "dev"]  %>% pull %>% round` times (`r rain_17[rain_17$station == "Piura", "precip"] %>% pull %>% round` mm) and Chulucanas `r rain_17[rain_17$station == "Chulucanas", "dev"]  %>% pull %>% round` times (`r rain_17[rain_17$station == "Chulucanas", "precip"] %>% pull %>% round` mm) more rain than on median (Fig.
 \@ref(fig:study-area-map)d).
@@
 Inside a sandy area devoid of woody vegetation, we placed 12 experimental plots of 3 × 3 m^2^, which were regularly spaced along a rectangular grid of 1 ha. 
-Plots were assigned to one of three irrigation treatments: The first treatment represented a Super EN year (example of 1997/98: 1780 mm), the second represented a moderate EN event (example of 1991/1992: 258 mm).
+Plots were assigned to one of three irrigation treatments: The first treatment represented a Super El Niño year (example of 1997/98: 1780 mm), the second represented a moderate El Niño event (example of 1991/1992: 258 mm).
 The last treatment represented the baseline with no additional water input and corresponded to the year in which the irrigation experiment was conducted (2013) which was a neutral year (54 mm; see also Fig.
@@
 Spatial cross-validation is an extension that measures a model's ability to spatially predict the response variable in the presence of spatial autocorrelation [@schratz_hyperparameter_2019]. 
-We ran a 100-repeated 5-fold spatial cross-validation in which the partitioning is based on *k*-means clustering (*k* = 5; [@russ_data_2010]).
+We ran a 100-repeated 5-fold spatial cross-validation in which the partitioning is based on *k*-means clustering (*k* = 5; @russ_data_2010).
 We also made sure that all annual observations of one plot were jointly placed in either the test or the training data set [@meyer_improving_2018]. 
@@
 All statistical analyses were conducted in the open source software R [@rcoreteam_language_2019] using its packages mgcv [@wood_generalized_2017], RQGIS ([@muenchow_rqgis_2017a], sperrorest [@brenning_spatial_2012] and vegan [@oksanen_vegan_2019].
-<!-- JM: add link to github repository, zenodo repository, dockerhub -->
-To comply with the good practices of reproducibility we will make available the data and the code of this work on an open accessible github repository (https://github.com/) upon acceptance of the manuscript.
 
@@
 
-Mean alpha diversity was lowest during the dry LN (2011) and highest during the moderate EN (2016; Table \@ref(tab:desc-stats)).
-Compared to the humid LN (2012) and the moderate EN (2016), species richness was lower in the east of the study area during the very humid neutral year(2017), however, higher within a distance of 20-25 km to the coast (left panel of Fig.
+Mean alpha diversity was lowest during the dry La Niña (2011) and highest during the moderate El Niño (2016; Table \@ref(tab:desc-stats)).
+Compared to the humid La Niña (2012) and the moderate El Niño (2016), species richness was lower in the east of the study area during the very humid neutral year(2017), however, higher within a distance of 20-25 km to the coast (left panel of Fig.
 \@ref(fig:spri)).
 Species turnover (beta diversity), expressed as the range of the first DCA axis (see also Fig.
-\@ref(fig:dca)), was also lowest during the dry LN (2011), slightly increased in 2012 (humid LN) and 2016 (moderate EN) and reached its highest value in 2017 (very humid neutral year).
-Gamma diversity increased from 2011 (dry LN) to 2012 (humid LN), remained almost the same in 2016 (moderate EN), and decreased in 2017 (very humid neutral year; Table \@ref(tab:desc-stats)).
+\@ref(fig:dca)), was also lowest during the dry La Niña (2011), slightly increased in 2012 (humid La Niña) and 2016 (moderate El Niño) and reached its highest value in 2017 (very humid neutral year).
+Gamma diversity increased from 2011 (dry La Niña) to 2012 (humid La Niña), remained almost the same in 2016 (moderate El Niño), and decreased in 2017 (very humid neutral year; Table \@ref(tab:desc-stats)).
 By contrast, the latter showed by far the highest plant coverage along the entire gradient (right panel of Fig. \@ref(fig:spri) & section \@ref(spatio-temporal-mapping)).
@@
 Numbers above the dots are the plot IDs which were ordered in accordance with the distance to the sea, i.e.
-plot number 1 is closest and plot number 50 farthest away from the sea.", out.width="70%", fig.align="center"}
+plot number 1 is closest and plot number 50 farthest away from the sea.", out.width="100%", fig.align="center"}
 
-knitr::include_graphics(here("figures/14_londo_dca.png"))
+# include_graphics(here("figures/14_londo_dca.png"))
+library("lattice")
+load(file = here("figures/14_londo_dca.rda"))
+print(p_1)
 ```
@@
 Overall, the DCA scatter cloud becomes more compressed over the years, and the plots become increasingly positioned along the first axis (Fig. \@ref(fig:dca)). 
-Put differently, the importance of the second axis steadily decreased from 3.22 during the dry LN (2011) to 1.35 units of standard deviation during the very humid neutral year (2017).
-By contrast, the first axis was more important in all phases which were wetter than the dry LN (Table \@ref(tab:desc-stats)).
+Put differently, the importance of the second axis steadily decreased from 3.22 during the dry La Niña (2011) to 1.35 units of standard deviation during the very humid neutral year (2017).
+By contrast, the first axis was more important in all phases which were wetter than the dry La Niña (Table \@ref(tab:desc-stats)).
 
-```{r spri, echo=FALSE, out.width="80%", fig.align="center", fig.cap="Species richness (left panel) and plant cover (right panel, in \\%) along the spatio-temporal gradient.
+```{r spri, echo=FALSE, out.width="100%", out.height="30%", message=FALSE, fig.align="center", fig.cap="Species richness (left panel) and plant cover (right panel, in \\%) along the spatio-temporal gradient.
 To aid visual interpretation, loess smoothers were added."}
-include_graphics(here("figures/13_spri_cover.png"))
+# include_graphics(here("figures/13_spri_cover.png"))
+library("lattice")
+library("latticeExtra")
+my.panel.loess = function(x, y, span = 3/5, degree = 0.5, ...) {
+  loess.fit = loess.smooth(x, y, span = span, dgree = degree )
+  panel.lines(loess.fit$x, loess.fit$y, ...)
+}
+pal = rev(c("lightblue",  "#FDD39B", "pink", "lightgray"))
+a = readRDS(here("figures/13_spri.rds"))
+b = readRDS(here("figures/13_cover.rds"))
+plot(a + b)
 ```
@@
 
-```{r pred-map, echo=FALSE, out.width="80%", fig.align="center", fig.cap="Spatial prediction of the floristic composition along the cross-section for different ENSO episodes The urban boundaries of the cities Paita, Piura and Chulucanas are displayed in black.
-The Piura River and its tributaries are represented by light blue lines with a black contour.", fig.align="centerfig.align=fig.align="}
-include_graphics(here("figures/15_pred_map.png"))
+```{r pred-map, echo=FALSE, out.width="100%", fig.align="center", message=FALSE, fig.cap="Spatial prediction of the floristic composition along the cross-section for different ENSO episodes The urban boundaries of the cities Paita, Piura and Chulucanas are displayed in black.
+The Piura River and its tributaries are represented by light blue lines with a black contour."}
+# include_graphics(here("figures/15_pred_map.png"))
+library("sp")
+library("sf")
+load(here("figures/15_pred_map.rda"))
+print(p)
 ```
@@
 A clear trend towards more pronounced differences in vegetation formations became apparent between the studied years (Fig. \@ref(fig:pred-map)). 
-The dry LN (2011) showed a floristic composition that was almost uniform across great parts of the study area (Panel 2011 of Fig. \@ref(fig:pred-map)).
-This changed during the wet LN (2012) where three distinct vegetation formations developed (Panel 2012 of Fig. \@ref(fig:pred-map)). 
+The dry La Niña (2011) showed a floristic composition that was almost uniform across great parts of the study area (Panel 2011 of Fig. \@ref(fig:pred-map)).
+This changed during the wet La Niña (2012) where three distinct vegetation formations developed (Panel 2012 of Fig. \@ref(fig:pred-map)). 
 In the west, we observed a semi- to full desert with a sparse herb and grass cover, which at times was accompanied by isolated trees and bushes.
@@
 \@ref(fig:pred-map)) with *Cordia lutea*, *Prosopis pallida*, *Cenchrus echinatus* and *Antephora hermaphrodita* being the most common species.
-The differentiation of these three vegetation formations was even more pronounced during the moderate EN (Panel 2016 of Fig. \@ref(fig:pred-map)). 
+The differentiation of these three vegetation formations was even more pronounced during the moderate El Niño (Panel 2016 of Fig. \@ref(fig:pred-map)). 
 This trend was both amplified and disrupted during the very humid neutral year (2017) during which a dominant grass-herb formation stretched from the coast until far behind the city of Piura (dark blue pixels in Panel 2017 of Fig.
@@
 
-```{r varpart, echo=FALSE, out.width="50%", fig.align="center", fig.cap="Fractions explaining the floristic composition represented by the first two DCA axes in accordance with a RDA variation partitioning.
+```{r varpart, echo=FALSE, out.width="70%", fig.align="center", message=FALSE, fig.cap="Fractions explaining the floristic composition represented by the first two DCA axes in accordance with a RDA variation partitioning.
 Numbers within the circles represent the adjusted $R^2$.
@@
 ($<0.1$), * ($<0.05$), ** ($ <0.01$), and *** ($<0.001$)."}
-include_graphics(here("figures/16_varpart.png"))
+# include_graphics(here("figures/16_varpart.png"))
+# Variation partitioning for the four londo cts
+# attach packages
+library("vegan")
+library("dplyr")
+# read in functions
+source(here("code/funs/varpart_helper_funs.R"))
+
+load(here("images/16_input_varpart_plot.rda"))
+par(mfrow = c(2, 2))
+par(mar = c(0, 0, 0, 0))
+#par(tcl = -0.25)
+#par(mgp = c(2, 0.6, 0))
+for (i in seq_along(inp)) {
+  mod = ordi_varpart(inp[[i]], choices = 1:2, expl_1 = soil, expl_2 = topo,
+                     dca = TRUE)
+  # showvarparts(2, bg = c("hotpink","skyblue"))
+  # plot(mod$vp, bg = c("hotpink","skyblue"))
+  labs = mod$vp$part$indfract[, 3]
+  labs = round(labs, 2)
+  labs[c(1, 3)] = paste(labs[c(1, 3)],
+                        c(as.character(mod$a), as.character(mod$c)))
+  plot(mod$vp, labels = labs, bg = c("hotpink","skyblue"),
+       Xnames = "", cex = 1)  
+  text(x = 0.5, y = 1.1, labels = years[i], font = 2, cex = 1)
+  text(x = c(-0.4, 1.1), y = c(0.4, 0.4), labels = c("soil", "topography"),
+       cex = 1)
+}
 ```
@@
 The experiment revealed that edaphic variables are conducive to vegetation development (see also Appendix 3). 
-The fertilized plots representing moderate EN years exhibited a slightly greater plant coverage than in their unfertilized counterparts (middle panel of Fig. \@ref(fig:irrigation)). 
+The fertilized plots representing moderate El Niño years exhibited a slightly greater plant coverage than in their unfertilized counterparts (middle panel of Fig. \@ref(fig:irrigation)). 
 Adding even more water revealed the true beneficial effects of fertilizing.
-The fertilized plots simulating Super EN conditions showed a threefold increase in plant cover compared to their unfertilized counterparts (right panel of Fig. \@ref(fig:irrigation)).
+The fertilized plots simulating Super El Niño conditions showed a threefold increase in plant cover compared to their unfertilized counterparts (right panel of Fig. \@ref(fig:irrigation)).
 Nevertheless, there was an upper limit regarding the beneficial effects of nutrient addition. 
-Biomass production stopped increasing or even decreased in the fertilized Super-EN plot after 8 March 2013, which corresponded to 1487 mm of irrigation and 200 kg ha^-1^ of NH~4~NO~3~.
+Biomass production stopped increasing or even decreased in the fertilized Super-El Niño plot after 8 March 2013, which corresponded to 1487 mm of irrigation and 200 kg ha^-1^ of NH~4~NO~3~.
 
-```{r irrigation, echo=FALSE, out.width="80%", fig.align="center", fig.cap="Results of the irrigation-nutrient experiment.
+```{r irrigation, echo=FALSE, out.width="100%", fig.align="center", warning=FALSE, message=FALSE, fig.cap="Results of the irrigation-nutrient experiment.
 The dashed line refers to the water input, note that the left y-axis was log-transformed.
 The solid lines represent the observed plant coverage in \\% (right y-axis)."}
-include_graphics(here("figures/17_nutrient_irrigation_experiment.png"))
+# include_graphics(here("figures/17_nutrient_irrigation_experiment.png"))
+load(here("figures/17_nutrient_irrigation_experiment.rda"))
+print(p_1)
 ```
@@
 Alpha and gamma diversity increased with higher seasonal precipitation amounts. 
-The relationship, however, is non-linear, i.e., the increase in diversity reached a plateau with the rainfall values recorded under humid LN (2012) and under moderate EN (2016) conditions. 
+The relationship, however, is non-linear, i.e., the increase in diversity reached a plateau with the rainfall values recorded under humid La Niña (2012) and under moderate El Niño (2016) conditions. 
 Then even more water input as observed during the very humid neutral year (2017) resulted in a slightly decreased diversity but also in a massive and sudden increase of overall plant cover due to higher biomass production.
@@
 Still, rainy periods are very conducive to the regeneration of woody vegetation in (semi-)arid areas [@bowers_demographic_1997; @brown_reorganization_1997; @sitters_rainfalltuned_2012].
-For example, the growth and recruitment rate of *Prosopis pallida* is almost twice as high during wet EN years than in neutral or dry years in northern Peru [@lopez_climatic_2006].
+For example, the growth and recruitment rate of *Prosopis pallida* is almost twice as high during wet El Niño years than in neutral or dry years in northern Peru [@lopez_climatic_2006].
 
@@
 This is because in 2011 rainfall barely occurred along large parts of the study area, and therefore especially perennial species were recorded which can and must endure drought years [@salazar_effect_2018].
-By contrast, during the humid LN (2012) and the moderate EN (2016) the same three distinct vegetation formations developed along the gradient. 
-This can be explained by similar rainfall patterns though 2012 was a humid LN and 2016 a moderate EN year. 
+By contrast, during the humid La Niña (2012) and the moderate El Niño (2016) the same three distinct vegetation formations developed along the gradient. 
+This can be explained by similar rainfall patterns though 2012 was a humid La Niña and 2016 a moderate El Niño year. 
 Finally, a large water surplus during the very humid neutral year (2017) led to the disruption of the strictly ordered formations along the humidity gradient. 
@@
 \@ref(fig:varpart)).
-Interestingly, the shared explained variance in the floristic dataset is roughly the same for the years experiencing wetter conditions (humid LN (2012), moderate EN (2016) and the very humid neutral year (2017; coastal EN)) though 2017 was the most humid by far.
+Interestingly, the shared explained variance in the floristic dataset is roughly the same for the years experiencing wetter conditions (humid La Niña (2012), moderate El Niño (2016) and the very humid neutral year (2017; coastal El Niño)) though 2017 was the most humid by far.
 This indicates that there is an upper limit to the beneficial effects of the water-soil and water-topography interaction.
@@
 
-## Climate change and opportunities for restoration and sustainable agrarian management
+## ENSO, climate change and ENSO indices
 
@@
 This and climate change in general pose the risk to turn tropical forests - dry and moist - from carbon sinks to sources [@phillips_changes_2009;@alvarez-davila_forest_2017;@salazar_tree_2019]. 
-This is also true for the study area [@salazar_tree_2019] where EN is frequently associated with higher rainfall and LN with increased aridity.
+This is also true for the study area [@salazar_tree_2019] where El Niño is frequently associated with higher rainfall and La Niña with increased aridity.
 However, both can also cause the opposite effects. 
-For instance, 2005 was a very dry EN year, 2012 was a humid LN year, and 2017 was first predicted to be a LN year and then turned out to be a very wet neutral year (coastal EN; see Fig.
+For instance, 2005 was a very dry El Niño year, 2012 was a humid La Niña year, and 2017 was first predicted to be a La Niña year and then turned out to be a very wet neutral year (coastal El Niño; see Fig.
 \@ref(fig:study-area-map)d for many more examples).
@@
 Naturally, local anomalies possibly overlap, dilute and modify the ENSO signal stemming from the central Pacific. 
-Studies have shown that indices computed from the EN region 1+2, which corresponds to an area in the tropical Pacific close to the South American coast, already provide better correlations with climate anomalies in northern Peru and southern Ecuador [@kao_contrasting_2009;@salazar_effect_2018].
+Studies have shown that indices computed from the El Niño region 1+2, which corresponds to an area in the tropical Pacific close to the South American coast, already provide better correlations with climate anomalies in northern Peru and southern Ecuador [@kao_contrasting_2009;@salazar_effect_2018].
 At least two further local indexes seem to be also promising for predicting climate anomalies along the Peruvian coast.
@@
 
-In any case, constant monitoring is a prerequisite for a better understanding of ENSO-affected ecosystems. 
+# Conclusions
+
+Constant monitoring is a prerequisite for a better understanding of ENSO-affected ecosystems. 
 Since ENSO exerts its most direct influence on tropical regions, and all corresponding ecosystems are endangered without exception, the correct prediction of the expected ENSO effects are a cornerstone of informed conservation and reforestation management.
 Conserving and reversing the state of degraded ecosystems also helps to mitigate the effects of floodings and landslides due to excessive rainfalls, and thereby contributes to protect the population and infrastructure from unnecessary harm [@fraser_peru_2017; @ramirez_understanding_2017].
-Currently, the regional government has set up eleven restoration projects in northern Peru which are trying to protect the endemic vegetation and increase vegetation cover through reforestation.
-Additionally, NGOs and mining companies are also working on reforestation projects.
-Most of these projects rely on extreme events (like 2017 EN) to boost plant growth to overcome the otherwise very limited access to water.
+Currently, the regional government has set up eleven restoration projects in northern Peru which are trying to protect the endemic vegetation and increase vegetation cover through reforestation [@cordova_piura_2019].
+Additionally, NGOs and mining companies are also working on reforestation projects [@gutierrez_anuncian_2015].
+Most of these projects rely on extreme events (like the 2017 El Niño) to boost plant growth to overcome the otherwise very limited access to water.
 Our experimental results indicated that water availability showed a non-linear relationship with biodiversity and plant cover.
@@
 Indeed, agricultural development has been very intense over the last 20 years in the study area.
-Personal communications with the main local companies indicated that extreme events are an important issue.
+Personal communications with the main local companies (e.g., https://www.ecosac.com.pe/) indicated that extreme events are an important issue.
 On the one hand, El Niño has the potential to boost crop productivity.
-On the other hand, especially short and very intense rainfall might create havoc during the crop season as was the case during the 2017 EN.
-Additionally, our results clearly reveal patterns of the spatial variability of biomass production in response to water availability along the climatic gradient, which might be also an important information for agrarian activities.
-
-# Conclusions
+On the other hand, especially short and very intense rainfall might create havoc during the crop season as was the case during the 2017 El Niño.
+Additionally, our results clearly reveal patterns in the spatial variability of biomass production in response to water availability along the climatic gradient, which might be also an important information for agrarian and restoration activities.
 

### 2020-08-10 text/ecography/round_2/supplementary_information.Rmd
migrate gitlab repo to github

---
output:
  bookdown::pdf_document2:
    keep_tex: no
    latex_engine: xelatex
    # float is needed for floating figures
    extra_dependencies: ["fontspec", "sectsty", "float"]
    toc: false
    fig_caption: no
    number_sections: true
header-includes:
  # adjust figure captions
  - \usepackage{caption}
  - \captionsetup{font={footnotesize,bf,singlespacing}}
  # Ecography requires double line spacing and line numbering
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage{lineno}
  - \linenumbers

title: Supplementary Information
bibliography: transect_refs.bib
csl: ecography.csl
# biblio-style: "natbib"
link-citations: true
---

```{r setup, include=FALSE}
library("knitr")
library("kableExtra")
library("tidyverse")
library("here")
# make sure that figures are allowed to float (hold the assigned position and are not always put on top of a page)
# however, to make sure that the last figure is now positioned after the references, I needed to set fig.caption: no in the yaml header, see:
# browseURL("https://stackoverflow.com/questions/29696172/how-to-hold-figure-position-with-figure-caption-in-pdf-output-of-knitr")
knitr::opts_chunk$set(fig.pos = "H")
```

\subsection*{Appendix 1. Study area stratified by distance to sea in 10 classes.}

```{r sampling-zones, echo=FALSE, fig.cap="Study area stratified by distance to sea in 10 classes."}
p_1 = readRDS(here("images/18_sample_stratification.rds"))
print(p_1)
```

\subsection*{Appendix 2: Detailed description of the methodological approach used for the irrigation-nutrient experiment.}

We established 12 experimental plots of 3 × 3 m^2 in a sandy area typical of the region.
Plots were located on the campus of the Universidad de Piura (Sechura Desert; 5°10'S, 80°38'W) on a sandy plain devoid of woody vegetation.
We built a barbed wire fence around the experimental area to avoid the intrusion of domestic animals such as deer wandering the campus.
The soils of the region are nutrient-poor (e.g., mean nitrogen content: 0.005 mg g^-1^) with high water infiltration rates [@muenchow_woody_2013].
The vegetation surrounding the experimental area mirrors the typical plant formation around Piura, hence a transition between tropical desert and open-shrubland vegetation [@muenchow_coupling_2013].
Prior to the irrigation experiment, we clipped the scarce aboveground biomass.
Between December 2012 and May 2013 four plots received the simulated rainfall amount of the Super El Niño event in 1997/98 (1780 mm), another four received the rainfall amount of the moderate El Niño event in 1991/92 (258 l mm) and the remaining four plots served as baseline plots without any additional water input.
Natural rainfall amounted to 53.9 mm during the study period (data: meteorological station of the University of Piura).
If natural rainfall occurred, it was subtracted from the artificial water input in the case of the Normal and Super Niño plots.
The irrigation water stemmed from rain-collecting tanks.
Irrigation was applied using hand sprinklers [@ronnenberg_effects_2011] before and after sunset and followed the intensity and frequency of the rainfall events of the El Niño episodes in 1991/92 and 1997/98 as recorded by the automatic climatic station of the University of Piura.
On the one hand, this reproduced the temporal pattern of El Niño rainfalls, which commonly took place at night time.
An on the other hand, this procedure impeded undesired evaporation and subsequent salinization effects.
The plan sandy underground prevented any runoff from irrigated plots to adjacent plots.

In addition to the water treatment, two randomly chosen plots of each irrigation treatment received a fertilization treatment of 200kg/ha granular nitrogen fertilizer (NH~4~NO~3~) using a hand-held spreader [@whitford_effects_2011]. To prevent a quick washing out of the fertilizer, we spread the granular nitrogen fertilizer four times during the study period (each time: 50 kg/h).

Plant development was monitored on a two-week basis and included the determination of plant species and corresponding coverage (%).
Biomass was harvested between mid-April and the beginning of May.
We randomly threw circles á 0.25 m^2^ four times into each plot and collected all plants within the circular area including the subterranean biomass. 
Soil particles were thoroughly removed from the plants.
Afterwards the biomass was subjected to 72 h of air drying.
Before weighing, biomass was finally oven-dried at a temperature of 105 °C for 24 hours.
From the total biomass we subtracted the subterranean biomass of the geophyte *Proboscidea altheifolia*, as it was already established prior to the irrigation experiment in most cases.

\subsection*{Appendix 3. Number of species plotted against time during the duration of the experiment.}

```{r alpha-exp, echo=FALSE, fig.cap="Number of species plotted against time during the duration of the experiment.", message=FALSE}
p_1 = readRDS(file = here("images/18_irrigation_spri.rds"))
print(p_1)
```

# References

### 2020-08-10 text/ecography/round_2/title_page.Rmd
migrate gitlab repo to github

---
output:
  bookdown::pdf_document2:
    keep_tex: yes
    latex_engine: xelatex
    # float is needed for floating figures
    extra_dependencies: ["fontspec", "sectsty", "float"]
    toc: false
    fig_caption: yes
#    number_sections: true
header-includes:
  # adjust figure captions
  - \usepackage{caption}
  - \captionsetup{font={footnotesize,bf,singlespacing}}
  # Ecography requires double line spacing and line numbering
  - \usepackage{setspace}
  - \doublespacing
  #- \usepackage{lineno}
  #- \linenumbers

title: "Monitoring and predictive mapping of floristic biodiversity along a climatic gradient in ENSO's terrestrial core region, NW Peru"
#bibliography: transect_refs.bib
#csl: ecography.csl
# biblio-style: "natbib"
#link-citations: true
---

Jannes Muenchow^1^, Petra Dieker^2^, Thea Böttcher^3^, Jonas Brock^4^, Gregor Didenko^1^, Tobias Fremout^5^, Desiree Jakubka^6^, Anke Jentsch^7^, Daniel Nüst^8^, Michael Richter^9^, Eric Frank Rodríguez^10^, Rodolfo Arismendiz Rodríguez^11^, Rütger Rollenbeck^12^, Pablo Salazar Zarsosa^11^, Patrick Schratz^1^, Alexander Brenning^1^

^1^ Institute of Geography, Friedrich Schiller University Jena, Löbdergraben 32, 07743 Jena, Germany. jannes.muenchow\@uni-jena.de. ORCID: https://orcid.org/0000-0001-7834-4717

^2^ Thünen Institute, Institute of Biodiversity, Bundesallee 65, 38116 Braunschweig, Germany

^3^ Institute of Biology, Humboldt-University of Berlin, Philippstr. 13, 10115 Berlin, Germany

^4^ Department of Ecological Modelling^,^ Helmholtz Centre for Environmental Research -- UFZ, Permoserstr. 15, 04318 Leipzig, Germany

^5^ Division of Forest, Nature and Landscape, KU Leuven, Heverlee (Leuven), Belgium

^6^ Institute of Ecology and Evolution, Friedrich-Schiller University Jena, Dornburger Str. 159, 07743 Jena, Germany

^7^ Disturbance Ecology, BayCEER, University of Bayreuth, Universitätsstr. 30, 95440 Bayreuth, Germany

^8^ Institute for Geoinformatics, University of Münster, Heisenbergstraße 2
48149 Münster, Germany

^9^ Institute of Geography, Friedrich Alexander University Erlangen-Nuremberg, 91058 Wetterkreuz 15, Erlangen, Germany

^10^ Herbarium Truxillense (HUT), National University of Trujillo, Trujillo, Peru

^11^ Faculty of Engineering, University of Piura, Piura, Peru

^12^ Laboratory for Climatology and Remote Sensing (LCRS), Faculty of Geography, University of Marburg, Deutschhausstr.10, 35032 Marburg, Germany

\section*{Abstract}

Though tropical dry forests (TDF) are the most endangered tropical ecosystem, they are still understudied.
Therefore, we studied the spatio-temporal effects of different El Niño Southern Oscillation (ENSO) episodes on TDF biodiversity in the terrestrial core region of ENSO in NW Peru.
For this, we sampled 50 permanent vegetation plots along a climatic gradient in four years representing different ENSO episodes: dry La Niña (2011), humid La Niña (2012), moderate El Niño (2016) and a very humid neutral year (coastal El Niño; 2017).
To visually represent the spatio-temporal changes in floristic composition, we used an innovative approach by modeling the ordination scores using a Generalized Additive Model and spatial cross-validation.
Additionally, we applied variation partitioning to test if topographic or edaphic variables gained in importance as soon as water was no longer the limiting factor.
Plant species richness increased under humid conditions during the humid La Niña (2012) and moderate El Niño (2016) year, and slightly decreased under the very humid conditions of the coastal El Niño year (2017).
Vegetation composition changed also under more humid conditions.
Vegetation cover was manifold higher in 2017 compared to all previous years.
Edaphic and topographic variables gained in importance with increased water availability in 2012 and 2016, however, this effect was not further amplified under very wet conditions (2017).
As nutrients can be equally important for primary production as water, we additionally executed an irrigation-nutrient experiment.
The experiment showed that biomass production under wet Super Niño conditions was three times higher when a fertilizer treatment was added.
Overall, our results can help to 1) protect local human populations from devastating ENSO effects through informed conservation management and the restoration of degraded ecosystems; 2) equip farmers with information regarding sustainable agrarian management.

**Keywords**: tropical dry forest, Peru, ecosystem diversity, ordination, statistical learning, predictive mapping, irrigation-fertilization experiment, productivity

*Acknowledgments* - We thank Msc Luis Urbina Zapata and Msc Nestor Atarama for their excellent work in the field. 
We are grateful to NASA LP DAAC for granting access to remote sensing imagery (ASTER, MODIS).

*Funding* - We are also most thankful to the Deutsche Forschungsgemeinschaft (project Ri 370/19-1) and a Friedrich Schiller University of Jena program to support junior researchers (project DRM/2017-05) for funding this work.

*Author contributions* - JM, MR and AJ conceived the ideas.
JM, TB, DJ, RAR and PSZ conducted the ﬁeldwork and collected the data with the help of collaborators.
JM, PD, AB, TF, JB, GD, RR, DJ, PS, DN and EFR analyzed the data.
JM, DN and PS were responsible for establishing the research compendium.
JM led the writing with the assistance of all co-authors.

### 2020-08-10 text/ecography/round_2/wip_discussion.Rmd
migrate gitlab repo to github

--- 2020-08-10 text/ecography/round_1/enso_comp.Rmd
@@
----
-output:
-  bookdown::pdf_document2:
-    keep_tex: yes
-    latex_engine: xelatex
-    # float is needed for floating figures
-    extra_dependencies: ["fontspec", "sectsty", "float"]
-    toc: false
-    fig_caption: yes
-    number_sections: true
-bibliography: transect_refs.bib
-csl: ecography.csl
----
-
-```{r setup, include=FALSE}
-library("knitr")
-library("kableExtra")
-library("tidyverse")
-library("here")
-# make sure that figures are allowed to float (hold the assigned position and are not always put on top of a page)
-knitr::opts_chunk$set(fig.pos = 'H')
-```
-
-# Introduction
-
-![Caption for the picture.](/home/jannes/work/acad/science/projects/ecology/latin_america/peru/transect_piura/transect_17/figures/12_map_grid.png)
-
-Studying the relationship between environmental gradients and species turnover has become a cornerstone of biogeography and community ecology since Humbold's and Bonpland's travels to South America more than 200 years ago [@macarthur_geographical_1984; @whittaker_gradient_1967].
-Naturally, water availability drives floristic composition and biomass production in the (semi-)arid regions such as NW Peru [@muenchow_coupling_2013; @muenchow_woody_2013].
-However, water availability does not only vary in space but also in time, especially in regions heavily affected by the El Niño Southern Oscillation (ENSO).
-<!--
-A super-humid El Niño (EN) episode benefits other species than a super-dry and fire-bringing La Niña (LN) episode [@richter_monitoring_2005].
--->
-
-ENSO is a recurrent climate phenomenon causing anomalies around the globe [@capotondi_understanding_2015].
-EN represents the warm and LN the cold phase of this cycle. 
-ENSO neutral refers to periods when neither EN nor LN are present. 
-LN usually causes the reverse effects of EN. 
-The effects depend on the region and include among others excessive rainfalls, droughts as well as temperature, wind and monsoon anomalies. 
-<!--
-The anomalies show recurrent patterns to a certain degree but are still highly variable especially on a local scale. 
-For instance, in 2016 the predicted Super-Niño rainfalls largely failed to materialize along the coasts of Ecuador and Peru.
-Just a year later, unexpected, torrential rainfalls hit the same coasts during a supposedly ENSO neutral year with a magnitude comparable to that of the Super-Niño event in 1997/98.
-Originating from a so-called coastal EN, the 2017 rainfalls came as a complete surprise, and caused the unprepared population unbearable hardship [@fraser_peru_2017; @ramirez_understanding_2017].
--->
-
-Though extreme climate events like EN and LN change and reshape ecosystems drastically [@Kogan.2017], ecological research mostly focusses on the effects of EN.
-Nevertheless, the eco-climatic impact of LN is comparable in magnitude with that of EN.
-For instance, LN episodes were most likely the main driver for the hiatus in global warming at the beginning of the 21^st^ century [@kosaka_recent_2013].
-In any case, ENSO certainly drives a large amount of the interannual variability of woody plant growth across the tropics [@rifai_enso_2018] though the specific response to El Niño may vary from region to region [@phillips_changes_2009].
-Overall, the effect of  extreme events is even pronounced in (semi)-arid ecosystems [@holmgren_extreme_2006], which is especially true for ENSO's terrestrial core region in NW Peru and SW Ecuador [@muenchow_woody_2013; @rollenbeck_climatic_2015] -- one of the most variable regions on Earth in terms of climate [@wehrden_global_2012].
-This region is also home to the highly tropical dry forest (TDF) ecosystems.
-TDFs are characterized by a high rate of endemism and feature an outstanding biodiversity which is almost as high as that of humid tropical forests [@espinosa_what_2011;@pennington_tropical_2018]. 
-At the same time, they are the most threatened tropical ecosystem [@miles_global_2006]. 
-In Peru, 95% of TDFs have already disappeared [@portillo-quintero_extent_2010]. 
-Even so, tropical dry ecosystems are chronically understudied [@escribano-avila_biodiversity_2017;@muenchow_review_2018].
-
-Species richness is strongly associated with precipitation in the tropics [@esquivel-muelbert_biogeographic_2017;@esquivel-muelbert_seasonal_2017].
-This is especially true for dry tropical ecosystems where water shortage is the main constraint to plant growth.
-Still, edaphic and topographic properties are sometimes equally important in structuring vegetation of tropical ecosystems [@laurance_influence_2010; @soethe_nutrient_2008; @condit_species_2013].
-Yet, the corresponding effects on the floristic composition of tropical (semi-)arid vegetation communities remain scarcely investigated [@muenchow_soil_2013; @pena-claros_soil_2012;@ulrich_climate_2014].
-This is especially true in terms of how soil properties influence vegetation communities as soon as water is no longer the limiting factor.
-For instance, nitrogen may increase primary production even in the presence of droughts and grazing pressure [@kinugasa_increasing_2012; @whitford_effects_2011].
-
-To tackle the outlined issues at least in part, we examined how different ENSO episodes affect the floristic composition along a spatio-temporal gradient in NW Peru. 
-For this, we surveyed the floristic composition on permanent plots in four years representing different ENSO episodes (dry (2011) and humid (2012) LN year; moderate EN year (2016); very humid coastal EN year (2017)).
-Specifically, we investigated the:
-
-(1) change in the different levels of biodiversity, namely alpha, beta, gamma and ecosystem diversity. 
-The latter was represented by ordination scores and additionally visualized by means of predictive mapping.
-Overall, we expected a boost in biodiversity as soon as water was no longer a limiting factor.
-Regarding the spatial distribution of the floristic composition we expected a more pronounced development of vegetation assemblages along the climatic gradient with wetter conditions [@engelbrecht_drought_2007].
-
-(2) effects of soil characteristics and topography in shaping floristic composition and vegetation cover (biomass production). 
-We expected that the influence of topographic and edaphic variables might gain in influence in wetter years.
-
-(3) We complemented the plot survey at the landscape scale with an irrigation-fertilization experiment, thereby considering the socio-economic perspective of the ENSO variability. 
-Specifically, we quantified the effect of the water-nutrient interaction on biomass development under different rainfall scenarios.
-We expected that the water input of the rainfall scenario representing a Super-Niño event in combination with the fertilizer would be most conducive to biomass and biodiversity development.
-
-# Methods
-
-## Study area
-
-\begin{figure}[H]
-\includegraphics[width=25in]{/home/jannes/work/acad/science/projects/ecology/latin_america/peru/transect_piura/transect_17/figures/12_map_grid.png}
-\caption{The study area.}
-\label{fig:study-area-map}
-\end{figure}
-
-The study area stretches for 120 km along a pronounced bioclimatic gradient in NW Peru and features a desert along the Pacific coast in the west and TDFs near the Andean foothills in the east (Fig.
-\@ref(fig:study-area-map)).
-The climatic gradient is at the same time a gradient of increasing human influence especially in terms of agricultural activity towards the Andean foothills [@muenchow_coupling_2013; @muenchow_woody_2013].
-Agricultural activity includes field farming and extensive grazing caused by free-ranging livestock.
-
-The study area is part of the core region where the ENSO phenomenon exerts its greatest terrestrial influence [@richter_monitoring_2005]. 
-Generally, EN brings more rain than usual and LN enforces the already dry conditions in the study area.
-However, even this pattern is highly variable and can occasionally be reversed, see e.g., the years 2005, 2008 and 2012 in Fig.
-\@ref(fig:study-area-map)d.
-Additionally, so-called "local" or "coastal EN events" can take place as observed in 2017.
-Coastal EN events refer to the regional warming of coastal waters in the Gulf of Guayaquil in December as opposed to the Pacific basin-wide EN when warm North Australian waters reach the South American coast.
-
-```{r, echo=FALSE, message=FALSE}
-precip = readRDS(here("images/12_precip.rds"))
-rain_17 = group_by(precip, station) %>% 
-  # compute the median for each station
-  mutate(median = median(precip)) %>% 
-  # just keep coastal EN year
-  filter(year == 2017)
-```
-
-Therefore, to put the ecological results into context, it is important to keep in mind the varying rainfall patterns as produced by different ENSO episodes across the studied years. 
-The observed years differed largely in terms of precipitation.
-2011 corresponded to a dry LN year and the whole study area experienced drier conditions than normal.
-By contrast, 2012 was a humid LN year, and 1.5 times and 3 times more rain fell in Piura and Chulucanas, i.e., in the middle to eastern part of the study area (Fig. \@ref(fig:study-area-map)).
-2016 was a moderate EN year and rainfall was double the median input in Paita and Piura, whereas in Chulucanas the median value was almost reached. 
-2017 coincided with a coastal EN event and is the only studied year during which the whole study area experienced exceptionally wet conditions: Paita received almost `r rain_17[rain_17$station == "Paita", "dev"]  %>% pull %>% round` times (`r rain_17[rain_17$station == "Paita", "precip"] %>% pull %>% round` mm), Piura almost `r rain_17[rain_17$station == "Piura", "dev"]  %>% pull %>% round` times (`r rain_17[rain_17$station == "Piura", "precip"] %>% pull %>% round` mm) and Chulucanas `r rain_17[rain_17$station == "Chulucanas", "dev"]  %>% pull %>% round` times (`r rain_17[rain_17$station == "Chulucanas", "precip"] %>% pull %>% round` mm) more rain than on median (Fig.
-\@ref(fig:study-area-map)d).
-Moreover, it is noteworthy that vegetation development changes drastically in the study area during wet episodes but remains more or less uniform during neutral years [@holmgren_extreme_2006; @muenchow_coupling_2013; @richter_monitoring_2005].
-
-## Data compilation
-
-### Landscape and small-scale experiment vegetation sampling
-
-We sampled 50 vegetation plots of 30 × 30 m^2^ size randomly along the climatic gradient between Paita and Chulucanas (Fig. \@ref(fig:study-area-map)).
-To ensure that all parts of the study area were equally likely to be sampled, we stratified the sampling by distance to the Pacific Ocean in ten classes (see Appendix 1).
-Prior to the sampling, we excluded riparian, agricultural and urban areas.
-Each plot was located with a handheld Garmin GPS device in the field.
-Field sampling took place towards the end of the vegetation period (March-May) in 2011, 2012, 2016 and 2017.
-In each plot, we recorded all vascular plant species and determined their cover as portion of the plot area in percentage points. 
-Specifically, plant cover refers to the relative projected area covered by a species [@damgaard_estimating_2014].
-Prior to the analysis, we transformed the plant cover values in accordance with the decimal scale by Londo [@londo_decimal_1976].
-The nomenclature follows the conventions of the Tropicos online database [@missouribotanicalgarden_tropicos_2018].
-
-In addition to the plot survey at the landscape scale, we conducted a small-scale irrigation-nutrient experiment to quantify the beneficial effects of the water-nutrient interaction on vegetation development and biomass production.
-We established the experiment in the middle of the precipitation gradient, ranging from `r rain_17[rain_17$station == "Paita", "median"] %>% pull %>% round` mm (Paita; annual median precipitation) to `r rain_17[rain_17$station == "Chulucanas", "median"] %>% pull %>% round` mm (Chulucanas; Fig.
-\@ref(fig:study-area-map)), on the campus of the University of Piura (5°10'S, 80°38'W, see Piura in Fig.
-\@ref(fig:study-area-map)).
-The environmental conditions are typical for the region in terms of climate, topography and soil properties (see @muenchow_soil_2013; @muenchow_woody_2013). 
-Inside a sandy area devoid of woody vegetation, we placed 12 experimental plots of 3 × 3 m^2^, which were regularly spaced along a rectangular grid of 1 ha. 
-Plots were assigned to one of three irrigation treatments: The first treatment represented a Super EN year (example of 1997/98: 1780 mm), the second represented a moderate EN event (example of 1991/1992: 258 mm).
-The last treatment represented the baseline with no additional water input and corresponded to the year in which the irrigation experiment was conducted (2013) which was a neutral year (54 mm; see also Fig.
-\@ref(fig:study-area-map)d).
-The irrigation water stemmed from rain-collecting tanks.
-In addition to the water treatment, two randomly chosen plots of each irrigation treatment received a fertilization treatment of 200kg/ha granular nitrogen fertilizer (NH~4~NO~3~). 
-The experiment was carried out between December 2012 and May 2013.
-Plant development was monitored two times per month and included the determination of plant species [@missouribotanicalgarden_tropicos_2018] and corresponding coverage (%).
-Please refer to Appendix 2 for a detailed description of the irrigation-nutrient experiment.
-
-### Environmental variables
-
-We randomly took and mixed three soil samples from 15cm and 30cm depth in each plot in 2011. 
-Soil composition should not vary too much between the years in which vegetation sampling took place since the study area largely features arenosols, i.e., a very weekly developed soil type.
-On the other hand, soil chemistry can vary largely between two years.
-For instance, El Niño events can increase soil respiration, i.e., soil CO\textsubscript{2} emissions, by a factor of 100, especially in the vicinity of trees [@salazar_tree_2019].
-Subsequently, we measured in the laboratory variables known to be important for vegetation development: pH, electrical conductivity (µS), carbon-to-nitrogen ratio, soil texture (%), the concentrations of the cations Ca, Mg, K and Na (cmol kg^-1^), and skeletal content (%; measured as gravimetric proportion of stones \>2 mm; see [@muenchow_woody_2013] for a more detailed description).
-
-We computed the topographic variables aspect, altitude, catchment slope, catchment area, soil wetness index, plan and vertical curvature from a digital elevation model with a 30 × 30 m^2^ resolution [@landprocessesdistributedactivearchivecenter_lp_2018].
-We downloaded the normalized difference vegetation index (NDVI) raster imagery from the Moderate Resolution Imaging Spectrometer (MODIS) available for the vegetation period (March, April) for each studied year (2011, 2012, 2016, 2017) at a 250 × 250 m^2^ resolution for our study area.
-Since three NDVI scenes were available for each vegetation period, we computed the mean NDVI for each vegetation period.
-
-The interpolated precipitation values from the WorldClim data provided a spatial distribution of the average annual rainfall [@hijmans_very_2005].
-Local precipitation values were collected from automated climate stations (Fig.
-\@ref(fig:study-area-map)d) operated by the University of Piura near Paita (1997-2018), in Piura (1991-2018) and in Chulucanas (1997-2018).
-
-## Data analysis
-
-### Biodiversity and plant cover analysis
-
-We calculated four measures of biodiversity for each observed year, namely alpha, beta, gamma and ecosystem diversity [@whittaker_gradient_1967].
-Alpha diversity refers to the mean number of species per plot.
-Beta diversity is expressed as the range of the first Detrended Correspondence Analysis (DCA) axis (see below). 
-This measure is given in units of standard deviation (SD) with four SDs corresponding to a complete species turnover.
-Gamma diversity is the total number of observed species across all plots.
-Ecosystem diversity was assessed with the help of a DCA [@hill_detrended_1980] to which the decimal scale transformed cover values of the species-plot matrix were subjected [@londo_decimal_1976]. 
-DCA tries to condense as much as possible the observed variance of the input matrix in a low-dimensional space (usually 2-3 axes), thereby extracting the main apparent gradients.
-
-For presenting the results of our nutrient-irrigation experiment we simply used descriptive statistics by plotting the increase in plant cover as well as water input against time for the duration of the experiment.
-
-### Modeling and predictive mapping
-
-<!-- JM: say, the nth aim in correspondence with the hypotheses made at the beginning -->
-One aim of this study was to spatio-temporally predict ecosystem diversity (floristic composition) represented by ordination scores along the climatic gradient in the study area.
-@muenchow_coupling_2013 have already shown that NDVI is an excellent predictor for vegetation assemblages in the study area.
-Therefore, we modeled the scores of the first DCA axis representing the main observed floristic gradient in the study area with the help of a Generalized Additive Model (GAM) as a function of the interaction between NDVI and year:
-$$
-\begin{aligned}
-\text{Scores}_{i}\ \sim\ \mathrm{N}\left( \mu_{i},\ \sigma^{2} \right) \\
-\mu_{i} = \alpha + f_{\text{year}_{i}}\left( \text{NDVI}_{i} \right) + \beta_{1}\text{year}_{i}
-\end{aligned}
-$$
-
-(Formula 1)
-
-where *i* refers to the *i*th observation, *$\alpha$* is the intercept, *$\beta$*~1~ is the estimated coefficient for year and *f* is a spline smoother for the NDVI:year interaction.
-
-The sole purpose of the model is the spatial prediction of the ordination scores, not statistical inference or the interpretation of the models' coefficients.
-We will use variation partitioning later on to assess how the influence of environmental predictors has changed over the studied years (see section \@ref(impact-of-environmental-variables)). 
-Consequently, here we are only interested in a good predictive performance of the model to make sure that it is able to generalize, i.e., that it is able to predict reasonably well unseen data [@james_introduction_2013; @kuhn_applied_2013].
-Only this can be the basis for a reliable map of the predicted ordination scores.
-To make sure this is the case, we used spatial cross-validation.
-Overall, cross-validation accounts for over-optimistic predictions on the training set by randomly splitting a data set into partitions used for training and testing [@brenning_spatial_2012;@lovelace_geocomputation_2019a].
-Spatial cross-validation is an extension that measures a model's ability to spatially predict the response variable in the presence of spatial autocorrelation [@schratz_hyperparameter_2019]. 
-We ran a 100-repeated 5-fold spatial cross-validation in which the partitioning is based on *k*-means clustering (*k* = 5; [@russ_data_2010]).
-We also made sure that all annual observations of one plot were jointly placed in either the test or the training data set [@meyer_improving_2018]. 
-We used the normalized root mean-square error (NRMSE) in percent as performance measure for the spatial cross-validation.
-
-$$
-NRMSE = \frac{\sqrt{\frac{\sum_{i = 1}^{n}{(e_{i} - y_{i})²}}{n}}}{max(y) - \ min(y)}*100
-$$ 
-(Formula 2)
-
-where *n* is the number of observations, *e* corresponds to the residuals, and *y* refers to the observed values.
-
-### Impact of environmental variables
-
-<!-- JM: add a first sentence, the nth aim of the study was to ... in correspondence with the posed hypotheses -->
-Variation partitioning quantifies the variation explained of the response, in our case the first two DCA axes, by one predictor group, in our case edaphic variables, while controlling for the effect of another predictor group [@borcard_numerical_2018; @peres-neto_variation_2006], in our case topographic variables.
-Internally, variation partitioning computes a Redundancy Analysis (RDA) separately for each predictor group and for all predictor groups together, and subesequently quantifies how much each predictor group explains alone and jointly with the other predictor group of the variation of the response.
-This is frequently displayed with the help of a Venn diagramm.
-We ran the variation partitioning for each studied year (2011, 2012, 2016, 2017) to assess how edaphic and topographic variables influence vegetation composition under increasingly humid conditions. 
-Prior to the variation partitioning, we log-transformed electrical conductivity, the CN ratio and the catchment area.
-Subsequently, we excluded all collinear variables which exceeded a variance inflation factor \>3 leaving us with seven edaphic and eight topographic variables. 
-We refrained from a forward selection of explanatory variables [@blanchet_forward_2008] because (i) we only used a reasonable number of meaningful, non-collinear variables that (ii) were available for all four years thus guaranteeing inter-annual comparability.
-
-All statistical analyses were conducted in the open source software R [@rcoreteam_language_2019] using its packages mgcv [@wood_generalized_2017], RQGIS ([@muenchow_rqgis_2017a], sperrorest [@brenning_spatial_2012] and vegan [@oksanen_vegan_2019].
-<!-- JM: add link to github repository, zenodo repository, dockerhub -->
-To comply with the good practices of reproducibility we will make available the data and the code of this work on an open accessible github repository (https://github.com/) upon acceptance of the manuscript.
-
-# Results
-
-## Biodiversity and plant cover
-
-```{r desc-stats, echo=FALSE}
-div = readRDS(here("images/13_desc_stats_tab.rds"))
-cap = paste("Different biodiversity measures across the studied years.",
-            "min = minimum observed number of species per plot.",
-            "max = maximum observed number of species per plot.")
-kable(div, "latex", booktabs = TRUE, escape = TRUE, 
-      # align = "c",
-      # col.names = linebreak(nms, align = "c"),
-      caption = cap)
-```
-
-Mean alpha diversity was lowest during the dry LN (2011) and highest during the moderate EN (2016; Table \@ref(tab:desc-stats)).
-Compared to the humid LN (2012) and the moderate EN (2016), species richness was lower in the east of the study area during the very humid neutral year(2017), however, higher within a distance of 20-25 km to the coast (left panel of Fig.
-\@ref(fig:spri)).
-Species turnover (beta diversity), expressed as the range of the first DCA axis (see also Fig.
-\@ref(fig:dca)), was also lowest during the dry LN (2011), slightly increased in 2012 (humid LN) and 2016 (moderate EN) and reached its highest value in 2017 (very humid neutral year).
-Gamma diversity increased from 2011 (dry LN) to 2012 (humid LN), remained almost the same in 2016 (moderate EN), and decreased in 2017 (very humid neutral year; Table \@ref(tab:desc-stats)).
-By contrast, the latter showed by far the highest plant coverage along the entire gradient (right panel of Fig. \@ref(fig:spri) & section \@ref(spatio-temporal-mapping)).
-
-```{r dca, echo=FALSE, fig.cap="Plotting the scores of the second DCA axis against the scores of the first DCA axis.
-To better visualize the cross-annual changes, the scores of all years were added in light gray to the background of each panel.
-The black dots refer to the scores of the year given in the strip of each panel.
-Numbers above the dots are the plot IDs which were ordered in accordance with the distance to the sea, i.e.
-plot number 1 is closest and plot number 50 farthest away from the sea.", out.width="70%", fig.align="center"}
-
-knitr::include_graphics(here("figures/14_londo_dca.png"))
-```
-
-The first two DCA axes explained 68% of the observed variance with the first axis contributing alone 48%. 
-The first axis is mainly associated with precipitation while the second axis is related to topographic variables such as vertical curvature and catchment area.
-Overall, the DCA scatter cloud becomes more compressed over the years, and the plots become increasingly positioned along the first axis (Fig. \@ref(fig:dca)). 
-Put differently, the importance of the second axis steadily decreased from 3.22 during the dry LN (2011) to 1.35 units of standard deviation during the very humid neutral year (2017).
-By contrast, the first axis was more important in all phases which were wetter than the dry LN (Table \@ref(tab:desc-stats)).
-
-```{r spri, echo=FALSE, out.width="80%", fig.align="center", fig.cap="Species richness (left panel) and plant cover (right panel, in \\%) along the spatio-temporal gradient.
-To aid visual interpretation, loess smoothers were added."}
-include_graphics(here("figures/13_spri_cover.png"))
-```
-
-## Spatio-temporal mapping
-
-Our model explained 76% of the observed variance with a spline smoother absorbing 3 *df* (Formula 1 & section \@ref(modeling-and-predictive-mapping)).
-The spatially cross-validated NRMSE also indicated a satisfying fit of 27% (SD 8.53% over 100 repetitions).
-
-```{r pred-map, echo=FALSE, out.width="80%", fig.align="center", fig.cap="Spatial prediction of the floristic composition along the cross-section for different ENSO episodes The urban boundaries of the cities Paita, Piura and Chulucanas are displayed in black.
-The Piura River and its tributaries are represented by light blue lines with a black contour.", fig.align="centerfig.align=fig.align="}
-include_graphics(here("figures/15_pred_map.png"))
-```
-
-A clear trend towards more pronounced differences in vegetation formations became apparent between the studied years (Fig. \@ref(fig:pred-map)). 
-The dry LN (2011) showed a floristic composition that was almost uniform across great parts of the study area (Panel 2011 of Fig. \@ref(fig:pred-map)).
-This changed during the wet LN (2012) where three distinct vegetation formations developed (Panel 2012 of Fig. \@ref(fig:pred-map)). 
-In the west, we observed a semi- to full desert with a sparse herb and grass cover, which at times was accompanied by isolated trees and bushes.
-This formation developed into an open xeric shrubland (green/yellow pixels in Fig. \@ref(fig:pred-map)) where *Acacia macracantha* and *Encelia canescens* were the dominating species.
-Further east, the shrubland turned into a TDF formation (orange/red pixels in Fig.
-\@ref(fig:pred-map)) with *Cordia lutea*, *Prosopis pallida*, *Cenchrus echinatus* and *Antephora hermaphrodita* being the most common species.
-The differentiation of these three vegetation formations was even more pronounced during the moderate EN (Panel 2016 of Fig. \@ref(fig:pred-map)). 
-This trend was both amplified and disrupted during the very humid neutral year (2017) during which a dominant grass-herb formation stretched from the coast until far behind the city of Piura (dark blue pixels in Panel 2017 of Fig.
-\@ref(fig:pred-map)).
-TDF formations were also less dominant (orange to red pixels).
-Though species composition remained roughly the same during the very humid neutral year (2017) compared to previous years, plant cover has increased drastically, and particularly in the western part of the study area where highest coverage values were observed for the grasses *Aristida adscensionis* and *Eragrostis cilianensis* and the seedlings of the shrub *Encelia canescens*.
-Among herbal species, *Crotalaria incana*, *Exodeconus maritimus* and *Tiquilia paronychioides* showed the highest cover values.
-
-## Influence of environmental variables
-
-The influence of edaphic and topographic variables on floristic composition was relatively small in the driest year (2011) with 25% and 28% of explained variance, respectively (Fig.
-\@ref(fig:varpart)).
-The explained variance roughly doubled in all other years, however, did not further increase with an excessive supply of water in the most humid year (2017; Fig. \@ref(fig:varpart)).
-
-```{r varpart, echo=FALSE, out.width="50%", fig.align="center", fig.cap="Fractions explaining the floristic composition represented by the first two DCA axes in accordance with a RDA variation partitioning.
-Numbers within the circles represent the adjusted $R^2$.
-Significance levels of the pure fractions are given in the soil and topography circles as .
-($<0.1$), * ($<0.05$), ** ($ <0.01$), and *** ($<0.001$)."}
-include_graphics(here("figures/16_varpart.png"))
-```
-
-## Irrigation-fertilization experiment
-
-The experiment revealed that edaphic variables are conducive to vegetation development (see also Appendix 3). 
-The fertilized plots representing moderate EN years exhibited a slightly greater plant coverage than in their unfertilized counterparts (middle panel of Fig. \@ref(fig:irrigation)). 
-Adding even more water revealed the true beneficial effects of fertilizing.
-The fertilized plots simulating Super EN conditions showed a threefold increase in plant cover compared to their unfertilized counterparts (right panel of Fig. \@ref(fig:irrigation)).
-Nevertheless, there was an upper limit regarding the beneficial effects of nutrient addition. 
-Biomass production stopped increasing or even decreased in the fertilized Super-EN plot after 8 March 2013, which corresponded to 1487 mm of irrigation and 200 kg ha^-1^ of NH~4~NO~3~.
-
-```{r irrigation, echo=FALSE, out.width="80%", fig.align="center", fig.cap="Results of the irrigation-nutrient experiment.
-The dashed line refers to the water input, note that the left y-axis was log-transformed.
-The solid lines represent the observed plant coverage in \\% (right y-axis)."}
-include_graphics(here("figures/17_nutrient_irrigation_experiment.png"))
-```
-
-# Discussion
-
-## Biodiversity and biomass
-
-Beta diversity (expressed as the range of the first DCA axis), increased with wetter conditions (Fig. \@ref(fig:dca) & Table \@ref(tab:desc-stats)).
-Naturally, higher species turnover was mainly driven by short-lived, less abundant species.
-Concordantly, @cleland_sensitivity_2013 reports similar findings from temperate grasslands.
-The change of the shape of the DCA point scatter between the relatively humid years (panels 2012 & 2016 of Fig.
-\@ref(fig:dca)) and the very humid neutral year (panel 2017 of Fig.
-\@ref(fig:dca)) can be most likely attributed to the increase of plant cover (see also next paragraph).
-
-Alpha and gamma diversity increased with higher seasonal precipitation amounts. 
-The relationship, however, is non-linear, i.e., the increase in diversity reached a plateau with the rainfall values recorded under humid LN (2012) and under moderate EN (2016) conditions. 
-Then even more water input as observed during the very humid neutral year (2017) resulted in a slightly decreased diversity but also in a massive and sudden increase of overall plant cover due to higher biomass production.
-Interestingly, only a few grass and forb species were in large parts responsible for this increase.
-This is in contrast to other dryland studies which showed that dormant and annual species increased biodiversity under more humid conditions [@gutierrez_variation_2000; @holmgren_extreme_2006].
-On the other hand, this finding is in line with the theory that under favorable environmental conditions a limited amount of species become the dominant actors of an ecosystem, whereas under intermediate levels of stress and disturbance a higher species richness is supported [@brown_why_2014].
-Though obviously a few dominant species outcompeted other species in 2017, they still followed a ruderal strategy by growing as fast and producing as much offspring as possible in the short period of favorable conditions [@schmidtlein_mapping_2012].
-Finally, free-ranging livestock might also have an impact on vegetation dynamics in the study area, especially in its eastern parts with more than 100mm of annual precipitation.
-Livestock might alter floristic composition and diversity while supporting especially plants adapted to grazing.
-
-Aside from the spectacular germination of annuals, perennial woody plants also largely benefit from wetter periods, e.g. through increased flower and fruit production [@holmgren_nino_2001]. 
-However, we only recorded plant cover, where the corresponding interannual difference is only marginal compared to previous years. 
-Still, rainy periods are very conducive to the regeneration of woody vegetation in (semi-)arid areas [@bowers_demographic_1997; @brown_reorganization_1997; @sitters_rainfalltuned_2012].
-For example, the growth and recruitment rate of *Prosopis pallida* is almost twice as high during wet EN years than in neutral or dry years in northern Peru [@lopez_climatic_2006].
-
-Overall, water surplus converts the (semi-)arid landscape interspersed with shrubs and trees temporarily into a tree savannah.
-Shallow underground water near the Pacific coast is largely responsible for the occurrence of perennial species in the study area.
-In fact, dendrochronological studies have shown that the growth of the most frequent tree species in the study area, *Prosopis pallida*, is nearly independent of the interannual climatic variability because frequently the water input of the low annual precipitation and nearby rivers suffices to replenish the groundwater [@brown_water_1990; @salazar_effect_2018; @throop_response_2012].
-
-Though we largely focussed on how current-year precipitation affects biodiversity dynamics, previous year rainfall might also have an effect [@tenhumberg_timelagged_2018].
-For instance, @dudney_lagging_2017 observed an increase in grass and a decrease in forb abundance when rainfall was high one year earlier.
-Similarly, @walter_plants_2011 suggest that grasses might have a 'drought-memory'.
-Time-lagged responses can be also observed in woody species.
-For example, tree ring growth of *Prosopis pallida* is highly correlated with rainfall of the previous year [@salazar_effect_2018].
-
-## Floristic gradient mapping across ENSO episodes
-
-The spatial prediction map highlights that specific vegetation formations become more pronounced with increasing water input. 
-In the driest year (2011), a rather uniform distribution of the same floristic composition (green pixels in Fig. \@ref(fig:pred-map)) developed along large parts of the study area. 
-The corresponding DCA scores close to 0 indicates that plant community composition was primarily made up of species that are largely independent of the climatic gradient [@hill_detrended_1980; @vonwehrden_pluralism_2009].
-This is because in 2011 rainfall barely occurred along large parts of the study area, and therefore especially perennial species were recorded which can and must endure drought years [@salazar_effect_2018].
-By contrast, during the humid LN (2012) and the moderate EN (2016) the same three distinct vegetation formations developed along the gradient. 
-This can be explained by similar rainfall patterns though 2012 was a humid LN and 2016 a moderate EN year. 
-Finally, a large water surplus during the very humid neutral year (2017) led to the disruption of the strictly ordered formations along the humidity gradient. 
-On the one hand, dense grass and herb cover dominated large parts of the study area. The increased cover dominance of annual species is visible in larger negative DCA score values. 
-On the other hand, vegetation patches (yellow, green, and orange pixels) can be found near the coast which are otherwise more typical of the eastern part of the gradient (section \@ref(spatio-temporal-mapping) & Fig.
-\@ref(fig:pred-map)).
-TDF formations (orange and red pixels) are less visible in 2017 because the dense grass/herb cover temporarily converted the landscape into a savannah-like vegetation formation [@espinosa_what_2011; @salazar_effect_2018].
-Another interpretation could be also that the NDVI values, our main predictor for the floristic composition, is similar for the dense cover of annual plants and perennial woody plants in 2017 [@pena_assessing_2015; @salazar_intraspecific_2018].
-
-## Changing impact of environmental variables with wetter conditions
-
-<!-- 
-Interpreting the ordination axes is interesting, however, slighly deviating from the main purpose of the paper.
-Therefore, I have commented this part.
-
-The first ordination axis represents the main floristic gradient in the study area (Fig.
-\@ref(fig:dca)) and is associated with the increasing amount of precipitation from east to west (see section \@ref(biodiversity-and-plant-cover)).
-The second DCA axis is associated with topographic variables (section \@ref(biodiversity-and-plant-cover); see also @muenchow_woody_2013 and @muenchow_coupling_2013). 
-Topographic heterogeneity was most important during the dry LN (panel 2011 of Fig.
-\@ref(fig:dca)) and became less important in more humid years as evidenced by the concentration of the points along the first axis (panels 2012, 2016 & 2017 of Fig.
-\@ref(fig:dca)).
--->
-
-Variation partitioning clearly reveals that environmental predictors (topography, soil) gain in importance for explaining plant diversity and productivity dynamics as soon as water is no longer restricting plant growth (section \@ref(influence-of-environmental-variables) & Fig.
-\@ref(fig:varpart)).
-Interestingly, the shared explained variance in the floristic dataset is roughly the same for the years experiencing wetter conditions (humid LN (2012), moderate EN (2016) and the very humid neutral year (2017; coastal EN)) though 2017 was the most humid by far.
-This indicates that there is an upper limit to the beneficial effects of the water-soil and water-topography interaction.
-Topographic variables such as curvature and catchment area play some role in structuring the vegetation composition in the study area especially in dry years [@muenchow_coupling_2013; @muenchow_woody_2013].
-Similar findings could be also observed in Mexico where topography also partly explained the composition of dry forests [@mendez-toribio_effects_2016].
-However, the most important topographic variable in the study area was altitude since it is highly correlated with mean annual precipitation. 
-Of course, edaphic variables such as pH are also related to precipitation.
-The influence of edaphic variables on vegetation composition could even increase with more developed soils.
-However, large parts of the study area consist of sandy soils (arenosols). 
-Only in the east of the study area, soils show more signs of humification and a brownish color due to iron oxide release from primary minerals.
-
-Naturally, nutrients can play a major role in drylands [@ronnenberg_effects_2011; @whitford_effects_2011] as confirmed by our irrigation-nutrient experiment.
-In fact, the water-nitrogen interaction can result in a biomass production three times as high as the production achieved when adding the same amount of water but without additional nutrients (Fig.
-\@ref(fig:irrigation)).
-This is of utmost importance for sustainable yet productive agricultural management. 
-In combination with a reliable climate prediction, it supports the decision on which crops (annual herbal and grass species) to grow and how much fertilizer to use.
-Other studies have also confirmed the importance of soil for the composition of the vegetation in drylands.
-For example, @ulrich_climate_2014 found that soil fertility and variability in rainfall and temperature are the most important predictors for beta diversity.
-Similarly, soil plays an important role in Colombian dry forests [@gonzalez-m_disentangling_2018] 
-Finally, @espinosa_what_2011 found out that soil moisture, soil temperature and nitrogen explained in large parts the floristic composition of Tumbesian dry forests, which is quite close to our study area.
-
-## Climate change and opportunities for restoration and sustainable agrarian management
+## ENSO and climate change and opportunities for restoration and sustainable agrarian management
 
 Climate change will most likely increase the frequency and intensity of El Niño events [@cai_increasing_2014; @IPCC.2014; @wang_continued_2017].
-This and climate change in general pose the risk to turn tropical forests - dry and moist - from carbon sinks to sources [@phillips_changes_2009;@alvarez-davila_forest_2017;@salazar_tree_2019]. 
-This is also true for the study area [@salazar_tree_2019] where EN is frequently associated with higher rainfall and LN with increased aridity.
+This and climate change in general pose the risk to turn tropical forests - dry and moist - from carbon sinks to sources [@phillips_changes_2009; @salazar_tree_accepted]. 
+This is also true for the study area [@salazar_tree_accepted] where EN is frequently associated with higher rainfall and LN with increased aridity.
 However, both can also cause the opposite effects. 
@@
 On the other hand, especially short and very intense rainfall might create havoc during the crop season as was the case during the 2017 EN.
-Additionally, our results clearly reveal patterns of the spatial variability of biomass production in response to water availability along the climatic gradient, which might be also an important information for agrarian activities.
+Our results show the spatial variability of biomass production in response to water availability along the climatic gradient, which might be also an important information for agrarian activities.
 
-# Conclusions
+## Conclusions
 
-<!--
-EN and LN episodes are highly variable in terms of frequency, intensity and impact in ENSO's terrestrial core region.
-Climate change is believed to even increase the frequency of extreme EN and LN events (IPCC 2014, Cai et al. 2014, Wang et al. 2017).
-Failing to predict the occurrence of such extreme events causes the local population unbearable hardship as was the case along the Ecuadorian and Peruvian coast in 2017 [@fraser_peru_2017; @ramirez_understanding_2017].
-Reforestation and renaturation are among the most effective countermeasures to prevent and mitigate hardship caused by flooding and catastrophic mass movements since intact ecosystems can store larger quantities of water and additionally contribute to increased slope stability.
-Moreover, our study has shown that the nitrogen-water interaction has increased biomass production manifold, a result which is highly valuable for sustainable agrarian management.
-However, both agrarian and effective conservation management depend on reliable climatic predictions, especially regarding the effect and intensity of the ENSO episode.
-For these predictions, authorities, stakeholders and conservationists should not solely rely on the ONI index but also consider local indices.
-Overall, it is time to overthink our sometimes overly simplistic understanding of ENSO (Richter et al. 2018).
--->
+## Title
 
-# References {#references .ListParagraph}
+Monitoring and predictive mapping of floristic biodiversity along a climatic gradient across four years in ENSO's terrestrial core region, NW Peru
+
+along a spatio-temporal gradient
+
+relationship with environmental variables
+
+machine learning, statistical learning, variation partitioning, DCA (ordination)

### 2020-08-10 text/ecography/round_3/enso_comp.Rmd
migrate gitlab repo to github

--- 2020-08-10 text/ecography/round_2/enso_comp.Rmd
@@
 title: "Monitoring and predictive mapping of floristic biodiversity along a climatic gradient in ENSO's terrestrial core region, NW Peru"
-bibliography: transect_refs.bib
-csl: ecography.csl
+bibliography: ../transect_refs.bib
+csl: ../ecography.csl
 # biblio-style: "natbib"
@@
 library("here")
+
+# needed for dates in experiment figure species richness, otw months are displayed in German
+Sys.setlocale("LC_TIME", "C")
+
 # make sure that figures are allowed to float (hold the assigned position and are not always put on top of a page)
@@
 Studying the relationship between environmental gradients and species turnover has become a cornerstone of biogeography and community ecology since Humbold's and Bonpland's travels to South America more than 200 years ago [@macarthur_geographical_1984; @whittaker_gradient_1967].
-Naturally, water availability drives floristic composition and biomass production in the (semi-)arid regions such as NW Peru [@muenchow_coupling_2013; @muenchow_woody_2013].
+Naturally, water availability drives floristic composition and biomass production in (semi-)arid regions such as NW Peru [@muenchow_coupling_2013; @muenchow_woody_2013], which is the area under study in this manuscript.
+In fact, water availability is probably the most important limiting abiotic factor in NW Peru's tropical dry forests, where it controls plant recruitment, establishment and survival, seedling production, vegetation cover as well as species diversity [@balvanera_distribution_2011;@espinosa_what_2011;@salazar_tree_2020].
 However, water availability does not only vary in space but also in time, especially in regions heavily affected by the El Niño Southern Oscillation (ENSO).
@@
 ENSO is a recurrent climate phenomenon causing anomalies around the globe [@capotondi_understanding_2015].
-El Niño represents the warm and La Niña the cold phase of this cycle. 
-ENSO neutral refers to periods when neither El Niño nor La Niña are present. 
-La Niña usually causes the reverse effects of El Niño. 
-The effects depend on the region and include among others excessive rainfalls, droughts as well as temperature, wind and monsoon anomalies. 
+ENSO consists of alternating episodes.
+El Niño represents the warm and La Niña the cold episode of this cycle. 
+ENSO neutral refers to periods when neither El Niño nor La Niña episodes are present. 
+The study area is part of the core region where the ENSO phenomenon exerts its greatest terrestrial influence [@richter_monitoring_2005]. 
+Generally, El Niño episodes bring more rain than usual and La Niña episodes enforce the already dry conditions in the study area [<50 mm of annual precipitation along the coast of NW Peru; Fig.
+\@ref(fig:study-area-map)D;  @rollenbeck_climatic_2015].
+
+Especially El Niño episodes provoke drastic ecological responses [@Kogan.2017] but occur irregularly (five times over the last 20 years in the study area), and are highly variable in terms of intensity.
 <!--
@@
 -->
+Nevertheless, the eco-climatic impact of La Niña episodes is comparable in magnitude with that of El Niño episodes.
+For instance, La Niña episodes were most likely the main driver for the hiatus in global warming at the beginning of the 21^st^ century [@kosaka_recent_2013].
+Therefore, one objective of this manuscript is to study the effect of different ENSO episodes, including La Niña episodes, on the floristic composition, comprised of all observed vascular plants, of the same ecosystem.
+In any case, ENSO drives a large amount of the interannual variability of woody plant growth across the tropics [@rifai_enso_2018] though the specific response to El Niño episodes may vary from region to region [@phillips_changes_2009].
+Overall, the effect of  extreme events is even more pronounced in (semi)-arid ecosystems [@holmgren_extreme_2006], and is especially true for ENSO's terrestrial core region in NW Peru and SW Ecuador [@muenchow_woody_2013; @rollenbeck_climatic_2015].
 
-Though extreme climate events like El Niño and La Niña change and reshape ecosystems drastically [@Kogan.2017], ecological research mostly focusses on the effects of El Niño.
-Nevertheless, the eco-climatic impact of La Niña is comparable in magnitude with that of El Niño.
-For instance, La Niña episodes were most likely the main driver for the hiatus in global warming at the beginning of the 21^st^ century [@kosaka_recent_2013].
-In any case, ENSO certainly drives a large amount of the interannual variability of woody plant growth across the tropics [@rifai_enso_2018] though the specific response to El Niño may vary from region to region [@phillips_changes_2009].
-Overall, the effect of  extreme events is even pronounced in (semi)-arid ecosystems [@holmgren_extreme_2006], which is especially true for ENSO's terrestrial core region in NW Peru and SW Ecuador [@muenchow_woody_2013; @rollenbeck_climatic_2015] -- one of the most variable regions on Earth in terms of climate [@wehrden_global_2012].
-This region is also home to the highly tropical dry forest (TDF) ecosystems.
+NW Peru is also home to the tropical dry forest (TDF) ecosystems.
 TDFs are characterized by a high rate of endemism and feature an outstanding biodiversity which is almost as high as that of humid tropical forests [@espinosa_what_2011;@pennington_tropical_2018]. 
-At the same time, they are the most threatened tropical ecosystem [@miles_global_2006]. 
-In Peru, 95% of TDFs have already disappeared [@portillo-quintero_extent_2010]. 
-Even so, tropical dry ecosystems are chronically understudied [@escribano-avila_biodiversity_2017;@muenchow_review_2018].
+At the same time, they are the most threatened tropical ecosystem [@miles_global_2006], and in general chronically understudied [@escribano-avila_biodiversity_2017;@muenchow_review_2018].
+Despite their unique biodiversity and threatened status, the TDFs of NW Peru are vastly understudied, and the spatio-temporal effect of extreme events like El Niño or La Niña episodes on the biodiversity patterns in this ecosystem remains largely unknown.
+
+Statistical learning has a long tradition in biodiversity research for detecting patterns in ecological data and predicting from these patterns [@guisan_predictive_2000].
+It is especially useful for quantifying species-environment relationships and for predicting the occurrence of species or species richness (predictive mapping), and therefore is an indispensable instrument for biodiversity management [@fremout_mapping_2020].
+Furthermore, predictive mapping can reveal ecological changes in space and time which otherwise might be hard or impossible to detect [@lovelace_geocomputation_2019a].
+Therefore, it is a crucial tool for gaining detailed insights into ecosystem dynamics, which other more descriptive methods such as global biodiversity measures and ordination techniques cannot capture.
+We aim at spatially predicting, for the first time, how different ENSO episodes might change and reshape the spatial distribution of the floristic composition in the study area.
 
@@
 For instance, nitrogen may increase primary production even in the presence of droughts and grazing pressure [@kinugasa_increasing_2012; @whitford_effects_2011].
+Hence, this study investigates how much the impact of edaphic and topographic variables varies in explaining the floristic composition between dry and humid ENSO years.
+Additionally, an irrigation-fertilization experiment should help to quantify in more detail the beneficial effects of the water-nutrient interaction on vegetation diversity and cover.
 
-To tackle the outlined issues at least in part, we examined how different ENSO episodes affect the floristic composition along a spatio-temporal gradient in NW Peru. 
-For this, we surveyed the floristic composition on permanent plots in four years representing different ENSO episodes (dry (2011) and humid (2012) La Niña year; moderate El Niño year (2016); very humid coastal El Niño year (2017)).
-Specifically, we investigated the:
+In summary, in this manuscript we study the effects of different ENSO episodes on biodiversity patterns and floristic composition  as well as the effect of environmental variables on the floristic composition under varying rainfall patterns.
+Specifically, we tested the following three hypotheses:
 
-(1) change in the different levels of biodiversity, namely alpha, beta, gamma and ecosystem diversity. 
-The latter was represented by ordination scores and additionally visualized by means of predictive mapping.
-Overall, we expected a boost in biodiversity as soon as water was no longer a limiting factor.
-Regarding the spatial distribution of the floristic composition we expected a more pronounced development of vegetation assemblages along the climatic gradient with wetter conditions [@engelbrecht_drought_2007].
+(1) We investigated the change in the different levels of biodiversity along a pronounced climatic gradient across different ENSO episodes. 
+Overall, we expected a boost in biodiversity as soon as water was no longer the predominant limiting factor.
+We were particularly interested in how different ENSO episodes modify the spatial distribution of the floristic composition.
+Overall, we expected a more pronounced development of vegetation formations along the climatic gradient with wetter conditions [@engelbrecht_drought_2007].
 
-(2) effects of soil characteristics and topography in shaping floristic composition and vegetation cover (biomass production). 
-We expected that the influence of topographic and edaphic variables might gain in influence in wetter years.
+(2) Water availability limits plant cover and species richness along the entire climatic gradient in the (semi-)arid study area.
+Therefore, we were interested in evaluating the effects of soil characteristics and topography changed in shaping floristic composition and vegetation cover (biomass production) with varying rainfall patterns as caused by different ENSO episodes.
+We expected that the influence of topographic and edaphic variables might gain in influence in wetter years, i.e., as soon as water is no longer the limiting factor.
 
-(3) We complemented the plot survey at the landscape scale with an irrigation-fertilization experiment, thereby considering the socio-economic perspective of the ENSO variability. 
-Specifically, we quantified the effect of the water-nutrient interaction on biomass development under different rainfall scenarios.
-We expected that the water input of the rainfall scenario representing a Super-Niño event in combination with the fertilizer would be most conducive to biomass and biodiversity development.
+(3) To quantify the beneficial effects of the water-nutrient interaction on vegetation cover development and species richness, we executed an irrigation-fertilization experiment simulating different ENSO rainfall scenarios.
+We expected that the water input of the rainfall scenario representing a Super-Niño episode in combination with the fertilizer would be most conducive to plant cover and species richness development.
 
@@
 Blue color tones indicate that NDVI differences were greatest between 2012 and 2011 (red: between 2016 and 2011).
-Panel D: Ratio between annual precipitation and median precipitation (median Paita: 25.9mm, median Piura: 71.4mm, median Chulucanas: 272.4mm) measured at the three automated climatic stations (see section \\\\@ref(environmental-variables)).
+Panel D: Ratio between annual precipitation and median precipitation (median Paita: 25.9mm, median Piura: 71.4mm, median Chulucanas: 272.4mm) measured at the three automated climatic stations (see section \\@ref(environmental-variables)).
 In case the precipitation was smaller than the median, we computed the ratio as: 1 / (precipitation / median(precipitation)) \\* -1.
@@
 The ENSO phases are in correspondence with the ONI index of the NOAA Climatic Prediction Center (http://origin.cpc.ncep.noaa.gov/products/analysis\\_monitoring/ensostuff/ONI\\_v5.php).
-We have intentionally left out the 1997/1998 Super-Niño event since this would have dominated the y-axis."}
+We have intentionally left out the 1997/1998 Super-Niño episode since this would have dominated the y-axis.
+Abbreviations: EN = El Niño episode; LN = La Niña episode."}
 # knitr::include_graphics(here("figures/12_map_grid.png"))
@@
 The climatic gradient is at the same time a gradient of increasing human influence especially in terms of agricultural activity towards the Andean foothills [@muenchow_coupling_2013; @muenchow_woody_2013].
-Agricultural activity includes field farming and extensive grazing caused by free-ranging livestock.
+Agricultural activity includes cultivation of annual crops and fruit trees and extensive grazing caused by free-ranging livestock.
 
-The study area is part of the core region where the ENSO phenomenon exerts its greatest terrestrial influence [@richter_monitoring_2005]. 
-Generally, El Niño brings more rain than usual and La Niña enforces the already dry conditions in the study area.
-However, even this pattern is highly variable and can occasionally be reversed, see e.g., the years 2005, 2008 and 2012 in Fig.
-\@ref(fig:study-area-map)d.
-Additionally, so-called "local" or "coastal El Niño events" can take place as observed in 2017.
-Coastal El Niño events refer to the regional warming of coastal waters in the Gulf of Guayaquil in December as opposed to the Pacific basin-wide El Niño when warm North Australian waters reach the South American coast.
+The ENSO phenomenon heavily impacts the study area [@richter_monitoring_2005]. 
+In general, El Niño episodes are associated with more rain than usual while La Niña episodes amplify the already dry conditions.
+However, even this pattern is highly variable and can occasionally be reversed (see e.g., the years 2005, 2008 and 2012 in Fig. \@ref(fig:study-area-map)d). 
+Additionally, so-called "local" or "coastal El Niño episodes" can take place as observed in 2017. 
+Coastal El Niño episodes refer to the regional warming of coastal waters in the Gulf of Guayaquil in December as opposed to the Pacific basin-wide El Niño episode, which occurs when warm North Australian waters reach the South American coast.
 
@@
 The observed years differed largely in terms of precipitation.
-2011 corresponded to a dry La Niña year and the whole study area experienced drier conditions than normal.
-By contrast, 2012 was a humid La Niña year, and 1.5 times and 3 times more rain fell in Piura and Chulucanas, i.e., in the middle to eastern part of the study area (Fig. \@ref(fig:study-area-map)).
-2016 was a moderate El Niño year and rainfall was double the median input in Paita and Piura, whereas in Chulucanas the median value was almost reached. 
-2017 coincided with a coastal El Niño event and is the only studied year during which the whole study area experienced exceptionally wet conditions: Paita received almost `r rain_17[rain_17$station == "Paita", "dev"]  %>% pull %>% round` times (`r rain_17[rain_17$station == "Paita", "precip"] %>% pull %>% round` mm), Piura almost `r rain_17[rain_17$station == "Piura", "dev"]  %>% pull %>% round` times (`r rain_17[rain_17$station == "Piura", "precip"] %>% pull %>% round` mm) and Chulucanas `r rain_17[rain_17$station == "Chulucanas", "dev"]  %>% pull %>% round` times (`r rain_17[rain_17$station == "Chulucanas", "precip"] %>% pull %>% round` mm) more rain than on median (Fig.
-\@ref(fig:study-area-map)d).
+2011 corresponded to a dry La Niña episode and the whole study area experienced drier conditions than normal.
+By contrast, 2012 was a humid La Niña episode, and 1.5 times and 3 times more rain fell in Piura and Chulucanas, i.e., in the middle to eastern part of the study area (Fig.
+\@ref(fig:study-area-map)A).
+2016 was a moderate El Niño episode and rainfall was double the median input in Paita and Piura, whereas in Chulucanas the median value was almost reached. 
+2017 coincided with a coastal El Niño episode and is the only studied year during which the whole study area experienced exceptionally wet conditions: Paita received almost `r rain_17[rain_17$station == "Paita", "dev"]  %>% pull %>% round` times (`r rain_17[rain_17$station == "Paita", "precip"] %>% pull %>% round` mm), Piura almost `r rain_17[rain_17$station == "Piura", "dev"]  %>% pull %>% round` times (`r rain_17[rain_17$station == "Piura", "precip"] %>% pull %>% round` mm) and Chulucanas `r rain_17[rain_17$station == "Chulucanas", "dev"]  %>% pull %>% round` times (`r rain_17[rain_17$station == "Chulucanas", "precip"] %>% pull %>% round` mm) more rain than on median (Fig.
+\@ref(fig:study-area-map)D).
 Moreover, it is noteworthy that vegetation development changes drastically in the study area during wet episodes but remains more or less uniform during neutral years [@holmgren_extreme_2006; @muenchow_coupling_2013; @richter_monitoring_2005].
@@
 
-We sampled 50 vegetation plots of 30 × 30 m^2^ size randomly along the climatic gradient between Paita and Chulucanas (Fig. \@ref(fig:study-area-map)).
-To ensure that all parts of the study area were equally likely to be sampled, we stratified the sampling by distance to the Pacific Ocean in ten classes (see Appendix 1).
+We sampled 50 vegetation plots of 30 × 30 m^2^ size along the climatic gradient between Paita and Chulucanas (Fig. \@ref(fig:study-area-map)).
+To ensure that all parts of the study area were equally likely to be sampled, we stratified the random sampling by distance to the Pacific Ocean in ten classes (see Appendix 1).
 Prior to the sampling, we excluded riparian, agricultural and urban areas.
@@
 In each plot, we recorded all vascular plant species and determined their cover as portion of the plot area in percentage points. 
-Specifically, plant cover refers to the relative projected area covered by a species [@damgaard_estimating_2014].
+Specifically, plant cover refers to the relative projected area covered by a species [@damgaard_estimating_2014], and is a proxy for biomass.
 Prior to the analysis, we transformed the plant cover values in accordance with the decimal scale by Londo [@londo_decimal_1976].
@@
 
-In addition to the plot survey at the landscape scale, we conducted a small-scale irrigation-nutrient experiment to quantify the beneficial effects of the water-nutrient interaction on vegetation development and biomass production.
-We established the experiment in the middle of the precipitation gradient, ranging from `r rain_17[rain_17$station == "Paita", "median"] %>% pull %>% round` mm (Paita; annual median precipitation) to `r rain_17[rain_17$station == "Chulucanas", "median"] %>% pull %>% round` mm (Chulucanas; Fig.
-\@ref(fig:study-area-map)), on the campus of the University of Piura (5°10'S, 80°38'W, see Piura in Fig.
-\@ref(fig:study-area-map)).
+In addition to the plot survey at the landscape scale, we conducted a small-scale irrigation-fertilization experiment to quantify the beneficial effects of the water-nutrient interaction on species richness, plant cover and biomass production.
+We established the experiment in the middle of the study area on on a protected dryland forest inside the campus of the University of Piura (5°10'S, 80°38'W, see Piura in Fig.
+\@ref(fig:study-area-map)A).
+Hence, the experiment was located in the same ecosystem as the landscape study (see previous paragraph) and more specifically in the middle of the observed precipitation gradient, ranging from `r rain_17[rain_17$station == "Paita", "median"] %>% pull %>% round` mm (Paita; annual median precipitation) to `r rain_17[rain_17$station == "Chulucanas", "median"] %>% pull %>% round` mm (Chulucanas; Fig.
+\@ref(fig:study-area-map)A).
 The environmental conditions are typical for the region in terms of climate, topography and soil properties (see @muenchow_soil_2013; @muenchow_woody_2013). 
 Inside a sandy area devoid of woody vegetation, we placed 12 experimental plots of 3 × 3 m^2^, which were regularly spaced along a rectangular grid of 1 ha. 
-Plots were assigned to one of three irrigation treatments: The first treatment represented a Super El Niño year (example of 1997/98: 1780 mm), the second represented a moderate El Niño event (example of 1991/1992: 258 mm).
-The last treatment represented the baseline with no additional water input and corresponded to the year in which the irrigation experiment was conducted (2013) which was a neutral year (54 mm; see also Fig.
-\@ref(fig:study-area-map)d).
+Plots were assigned to one of three irrigation treatments: The first treatment represented a Super El Niño episode (example of 1997/98: 1780 mm), the second represented a moderate El Niño episode (example of 1991/1992: 258 mm).
+The last treatment represented the baseline with no additional water input and corresponded to the year in which the irrigation experiment was conducted (2013) which was a neutral episode (54 mm; see also Fig.
+\@ref(fig:study-area-map)D).
 The irrigation water stemmed from rain-collecting tanks.
@@
 The experiment was carried out between December 2012 and May 2013.
-Plant development was monitored two times per month and included the determination of plant species [@missouribotanicalgarden_tropicos_2018] and corresponding coverage (%).
-Please refer to Appendix 2 for a detailed description of the irrigation-nutrient experiment.
+Monitoring was executed two times per month and included the identification of plant species [@missouribotanicalgarden_tropicos_2018] and corresponding coverage (%).
+Please refer to Appendix 2 for a detailed description of the irrigation-fertilization experiment.
 
@@
 On the other hand, soil chemistry can vary largely between two years.
-For instance, El Niño events can increase soil respiration, i.e., soil CO\textsubscript{2} emissions, by a factor of 100, especially in the vicinity of trees [@salazar_tree_2019].
-Subsequently, we measured in the laboratory variables known to be important for vegetation development: pH, electrical conductivity (µS), carbon-to-nitrogen ratio, soil texture (%), the concentrations of the cations Ca, Mg, K and Na (cmol kg^-1^), and skeletal content (%; measured as gravimetric proportion of stones \>2 mm; see [@muenchow_woody_2013] for a more detailed description).
+For instance, El Niño episodes can increase soil respiration, i.e., soil CO\textsubscript{2} emissions, by a factor of 100, especially in the vicinity of trees [@salazar_tree_2020].
+Subsequently, we measured in the laboratory variables known to be important for vegetation development: pH, electrical conductivity (µS), carbon-to-nitrogen ratio, soil texture (%), the concentrations of the cations Ca, Mg, K and Na (cmol kg^-1^), and skeletal content (%; measured as gravimetric proportion of stones \>2 mm; see @muenchow_woody_2013 for a more detailed description).
 
 We computed the topographic variables aspect, altitude, catchment slope, catchment area, soil wetness index, plan and vertical curvature from a digital elevation model with a 30 × 30 m^2^ resolution [@landprocessesdistributedactivearchivecenter_lp_2018].
-We downloaded the normalized difference vegetation index (NDVI) raster imagery from the Moderate Resolution Imaging Spectrometer (MODIS) available for the vegetation period (March, April) for each studied year (2011, 2012, 2016, 2017) at a 250 × 250 m^2^ resolution for our study area.
+We downloaded the normalized difference vegetation index (NDVI) raster imagery from the Moderate Resolution Imaging Spectrometer (MODIS; @landprocessesdistributedactivearchivecenter_lp_2018) available for the vegetation period (March, April) for each studied year (2011, 2012, 2016, 2017) at a 250 × 250 m^2^ resolution for our study area.
 Since three NDVI scenes were available for each vegetation period, we computed the mean NDVI for each vegetation period.
 
-The interpolated precipitation values from the WorldClim data provided a spatial distribution of the average annual rainfall [@hijmans_very_2005].
 Local precipitation values were collected from automated climate stations (Fig.
@@
 
-We calculated four measures of biodiversity for each observed year, namely alpha, beta, gamma and ecosystem diversity [@whittaker_gradient_1967].
+We calculated three measures of biodiversity for each observed year, namely alpha, beta and gamma diversity [@whittaker_gradient_1967].
 Alpha diversity refers to the mean number of species per plot.
-Beta diversity is expressed as the range of the first Detrended Correspondence Analysis (DCA) axis (see below). 
-This measure is given in units of standard deviation (SD) with four SDs corresponding to a complete species turnover.
 Gamma diversity is the total number of observed species across all plots.
-Ecosystem diversity was assessed with the help of a DCA [@hill_detrended_1980] to which the decimal scale transformed cover values of the species-plot matrix were subjected [@londo_decimal_1976]. 
+Beta diversity, in the sense of species turnover, is expressed as the range of the first Detrended Correspondence Analysis (DCA) axis.  
+The unit of a DCA axis is given in units of standard deviation (SD).
+4 SDs correspond to a complete species turnover while a difference between 1 to 1.4 SD units already correspond to half a change in species composition [@legendre_numerical_2012]. 
+Moreover, the floristic composition was assessed with the help of a DCA [@hill_detrended_1980] to which the decimal scale transformed cover values of the species-plot matrix were subjected [@londo_decimal_1976].
+Floristic composition implicitly represents the named biodiversity measures, and additionally considers plant cover.
 DCA tries to condense as much as possible the observed variance of the input matrix in a low-dimensional space (usually 2-3 axes), thereby extracting the main apparent gradients.
-
-For presenting the results of our nutrient-irrigation experiment we simply used descriptive statistics by plotting the increase in plant cover as well as water input against time for the duration of the experiment.
+To test if beta diversity and floristic composition, respectively, varied significantly between years, we analyzed the multivariate homogeneity of group dispersions [@anderson_multivariate_2006].
 
@@
 
-<!-- JM: say, the nth aim in correspondence with the hypotheses made at the beginning -->
-One aim of this study was to spatio-temporally predict ecosystem diversity (floristic composition) represented by ordination scores along the climatic gradient in the study area.
-@muenchow_coupling_2013 have already shown that NDVI is an excellent predictor for vegetation assemblages in the study area.
+One objective of this study was the spatio-temporal mapping of floristic composition represented by the scores of the first DCA axis along the climatic gradient in the study area.
+Please note that the resulting prediction map not only shows species composition but also differences in beta diversity on the predicted pixel level.
+This is because the units of a DCA axis are given in SDs (see also previous section).
+
+@muenchow_coupling_2013 have already shown that NDVI is an excellent predictor for vegetation formations in the study area.
 Therefore, we modeled the scores of the first DCA axis representing the main observed floristic gradient in the study area with the help of a Generalized Additive Model (GAM) as a function of the interaction between NDVI and year:
+
 $$
@@
 where *i* refers to the *i*th observation, *$\alpha$* is the intercept, *$\beta$*~1~ is the estimated coefficient for year and *f* is a spline smoother for the NDVI:year interaction. 
+Please refer to Appendix 3 for a visualization of the non-linear relationship between the response variable (ordination scores) and NDVI by year.
 
@@
 Consequently, here we are only interested in a good predictive performance of the model to make sure that it is able to generalize, i.e., that it is able to predict reasonably well unseen data [@james_introduction_2013; @kuhn_applied_2013].
-Only this can be the basis for a reliable map of the predicted ordination scores.
 To make sure this is the case, we used spatial cross-validation.
@@
 
-<!-- JM: add a first sentence, the nth aim of the study was to ... in correspondence with the posed hypotheses -->
 Variation partitioning quantifies the variation explained of the response, in our case the first two DCA axes, by one predictor group, in our case edaphic variables, while controlling for the effect of another predictor group [@borcard_numerical_2018; @peres-neto_variation_2006], in our case topographic variables.
-Internally, variation partitioning computes a Redundancy Analysis (RDA) separately for each predictor group and for all predictor groups together, and subesequently quantifies how much each predictor group explains alone and jointly with the other predictor group of the variation of the response.
+Internally, variation partitioning computes a Redundancy Analysis (RDA) separately for each predictor group and for all predictor groups together, and subsequently quantifies how much each predictor group explains alone and jointly with the other predictor group of the variation of the response.
 This is frequently displayed with the help of a Venn diagramm.
@@
 We refrained from a forward selection of explanatory variables [@blanchet_forward_2008] because (i) we only used a reasonable number of meaningful, non-collinear variables that (ii) were available for all four years thus guaranteeing inter-annual comparability.
+
+### Irrigation-fertilization experiment
+
+For presenting the results of our irrigation-fertilization experiment we used descriptive statistics and plotted the increase in the number of species, plant cover as well as water input against time.
 
@@
 
-Mean alpha diversity was lowest during the dry La Niña (2011) and highest during the moderate El Niño (2016; Table \@ref(tab:desc-stats)).
-Compared to the humid La Niña (2012) and the moderate El Niño (2016), species richness was lower in the east of the study area during the very humid neutral year(2017), however, higher within a distance of 20-25 km to the coast (left panel of Fig.
+Mean alpha diversity was lowest during the dry La Niña (2011) and highest during the moderate El Niño episode (2016; Table \@ref(tab:desc-stats)).
+Compared to the humid La Niña (2012) and the moderate El Niño (2016) episodes, species richness was lower in the east of the study area during the very humid neutral year(2017), however, higher within a distance of 20-25 km to the coast (left panel of Fig.
 \@ref(fig:spri)).
 Species turnover (beta diversity), expressed as the range of the first DCA axis (see also Fig.
-\@ref(fig:dca)), was also lowest during the dry La Niña (2011), slightly increased in 2012 (humid La Niña) and 2016 (moderate El Niño) and reached its highest value in 2017 (very humid neutral year).
-Gamma diversity increased from 2011 (dry La Niña) to 2012 (humid La Niña), remained almost the same in 2016 (moderate El Niño), and decreased in 2017 (very humid neutral year; Table \@ref(tab:desc-stats)).
+\@ref(fig:dca)), was also lowest during the dry La Niña episode (2011), slightly increased in 2012 (humid La Niña episode) and 2016 (moderate El Niño episode) and reached its highest value in 2017 (very humid neutral year).
+Still, there were no significant interannual differences in beta diversity in accordance with the test for multivariate homogeneity of groups dispersions (ANOVA *F*-test: 0.05069).
+Gamma diversity increased from 2011 (dry La Niña episode) to 2012 (humid La Niña episode), remained almost the same in 2016 (moderate El Niño episode), and decreased in 2017 (very humid neutral year; Table \@ref(tab:desc-stats)).
 By contrast, the latter showed by far the highest plant coverage along the entire gradient (right panel of Fig. \@ref(fig:spri) & section \@ref(spatio-temporal-mapping)).
-
-```{r dca, echo=FALSE, fig.cap="Plotting the scores of the second DCA axis against the scores of the first DCA axis.
-To better visualize the cross-annual changes, the scores of all years were added in light gray to the background of each panel.
-The black dots refer to the scores of the year given in the strip of each panel.
-Numbers above the dots are the plot IDs which were ordered in accordance with the distance to the sea, i.e.
-plot number 1 is closest and plot number 50 farthest away from the sea.", out.width="100%", fig.align="center"}
-
-# include_graphics(here("figures/14_londo_dca.png"))
-library("lattice")
-load(file = here("figures/14_londo_dca.rda"))
-print(p_1)
-```
-
-The first two DCA axes explained 68% of the observed variance with the first axis contributing alone 48%. 
-The first axis is mainly associated with precipitation while the second axis is related to topographic variables such as vertical curvature and catchment area.
-Overall, the DCA scatter cloud becomes more compressed over the years, and the plots become increasingly positioned along the first axis (Fig. \@ref(fig:dca)). 
-Put differently, the importance of the second axis steadily decreased from 3.22 during the dry La Niña (2011) to 1.35 units of standard deviation during the very humid neutral year (2017).
-By contrast, the first axis was more important in all phases which were wetter than the dry La Niña (Table \@ref(tab:desc-stats)).
 
@@
 
+The first two DCA axes explained 68% of the observed variance with the first axis alone contributing 48%. 
+The first axis was mainly associated with precipitation while the second axis was related to topographic variables such as vertical curvature and catchment area.
+Overall, the DCA scatter cloud became more compressed over the years, and the plots became increasingly positioned along the first axis (Fig. \@ref(fig:dca)). 
+Put differently, the importance of the second axis steadily decreased from 3.22 during the dry La Niña episode (2011) to 1.35 units of standard deviation during the very humid neutral year (2017).
+By contrast, the first axis was more important during all episodes which were wetter than the dry La Niña episode (Table \@ref(tab:desc-stats)).
+
+```{r dca, echo=FALSE, fig.cap="Plotting the scores of the second DCA axis against the scores of the first DCA axis.
+To better visualize the cross-annual changes, the scores of all years were added in light gray to the background of each panel.
+The black dots refer to the scores of the year given in the strip of each panel.
+Numbers above the dots are the plot IDs which were ordered in accordance with the distance to the sea, i.e.
+plot number 1 is closest and plot number 50 farthest away from the sea.
+To avoid cluttering the figure with too many numbers, we have only added a few selected plot numbers.", out.width="100%", fig.align="center"}
+
+# include_graphics(here("figures/14_londo_dca.png"))
+library("lattice")
+load(file = here("figures/14_londo_dca.rda"))
+print(p_1)
+```
+
 ## Spatio-temporal mapping
@@
 
-```{r pred-map, echo=FALSE, out.width="100%", fig.align="center", message=FALSE, fig.cap="Spatial prediction of the floristic composition along the cross-section for different ENSO episodes The urban boundaries of the cities Paita, Piura and Chulucanas are displayed in black.
+```{r pred-map, echo=FALSE, out.width="100%", fig.align="center", message=FALSE, warning=FALSE, fig.cap="Spatial prediction of the floristic composition along the cross-section for different ENSO episodes The urban boundaries of the cities Paita, Piura and Chulucanas are displayed in black.
 The Piura River and its tributaries are represented by light blue lines with a black contour."}
@@
 
+Since the predicted DCA scores represent the floristic composition of a pixel, clusters of pixels sharing similar values (visible as similar color tones in Fig.
+\@ref(fig:pred-map)) can be understood as belonging to the same vegetation formation.
 A clear trend towards more pronounced differences in vegetation formations became apparent between the studied years (Fig. \@ref(fig:pred-map)).
-The dry La Niña (2011) showed a floristic composition that was almost uniform across great parts of the study area (Panel 2011 of Fig. \@ref(fig:pred-map)).
-This changed during the wet La Niña (2012) where three distinct vegetation formations developed (Panel 2012 of Fig. \@ref(fig:pred-map)). 
-In the west, we observed a semi- to full desert with a sparse herb and grass cover, which at times was accompanied by isolated trees and bushes.
+The dry La Niña episode (2011) showed a floristic composition that was almost uniform across great parts of the study area (Panel 2011 of Fig. \@ref(fig:pred-map)).
+This changed during the wet La Niña episode (2012) where three distinct vegetation formations developed (Panel 2012 of Fig. \@ref(fig:pred-map)). 
+In the west, we observed a desert-like formation with a sparse herb and grass cover, which at times was accompanied by isolated trees and bushes (blue pixels in Panel 2012 of Fig.
+\@ref(fig:pred-map)) .
 This formation developed into an open xeric shrubland (green/yellow pixels in Fig. \@ref(fig:pred-map)) where *Acacia macracantha* and *Encelia canescens* were the dominating species.
@@
 \@ref(fig:pred-map)) with *Cordia lutea*, *Prosopis pallida*, *Cenchrus echinatus* and *Antephora hermaphrodita* being the most common species.
-The differentiation of these three vegetation formations was even more pronounced during the moderate El Niño (Panel 2016 of Fig. \@ref(fig:pred-map)). 
+The differentiation of these three vegetation formations was even more pronounced during the moderate El Niño episode (Panel 2016 of Fig. \@ref(fig:pred-map)). 
 This trend was both amplified and disrupted during the very humid neutral year (2017) during which a dominant grass-herb formation stretched from the coast until far behind the city of Piura (dark blue pixels in Panel 2017 of Fig.
@@
 TDF formations were also less dominant (orange to red pixels).
-Though species composition remained roughly the same during the very humid neutral year (2017) compared to previous years, plant cover has increased drastically, and particularly in the western part of the study area where highest coverage values were observed for the grasses *Aristida adscensionis* and *Eragrostis cilianensis* and the seedlings of the shrub *Encelia canescens*.
+Though species composition remained roughly the same during the very humid neutral year (2017) compared to previous years, plant cover increased drastically, and particularly in the western part of the study area where highest coverage values were observed for the grasses *Aristida adscensionis* and *Eragrostis cilianensis* and the seedlings of the shrub *Encelia canescens*.
 Among herbal species, *Crotalaria incana*, *Exodeconus maritimus* and *Tiquilia paronychioides* showed the highest cover values.
@@
        Xnames = "", cex = 1)  
-  text(x = 0.5, y = 1.1, labels = years[i], font = 2, cex = 1)
+  text(x = 0.5, y = 0.8, labels = years[i], font = 2, cex = 1)
   text(x = c(-0.4, 1.1), y = c(0.4, 0.4), labels = c("soil", "topography"),
@@
 
-The experiment revealed that edaphic variables are conducive to vegetation development (see also Appendix 3). 
-The fertilized plots representing moderate El Niño years exhibited a slightly greater plant coverage than in their unfertilized counterparts (middle panel of Fig. \@ref(fig:irrigation)). 
+The experiment revealed that edaphic variables are conducive to vegetation development. 
+The fertilized plots representing moderate El Niño episodes exhibited a slightly greater plant coverage than in their unfertilized counterparts (middle panel of Fig. \@ref(fig:irrigation)). 
 Adding even more water revealed the true beneficial effects of fertilizing.
-The fertilized plots simulating Super El Niño conditions showed a threefold increase in plant cover compared to their unfertilized counterparts (right panel of Fig. \@ref(fig:irrigation)).
+The fertilized plots simulating Super El Niño episodes showed a threefold increase in plant cover compared to their unfertilized counterparts (right panel of Fig. \@ref(fig:irrigation)).
 Nevertheless, there was an upper limit regarding the beneficial effects of nutrient addition. 
-Biomass production stopped increasing or even decreased in the fertilized Super-El Niño plot after 8 March 2013, which corresponded to 1487 mm of irrigation and 200 kg ha^-1^ of NH~4~NO~3~.
+Plant cover, and thus biomass production, stopped increasing or even decreased in the fertilized Super-El Niño plots after 8 March 2013, which corresponded to 1487 mm of irrigation and 200 kg ha^-1^ of NH~4~NO~3~.
+Species richness followed a similar pattern -- highest numbers in species were recorded in the the fertilized Super-El Niño plots, though with just seven observed species the overall number remained low (Fig.
+\@ref(fig:spri-exp)).
 
-```{r irrigation, echo=FALSE, out.width="100%", fig.align="center", warning=FALSE, message=FALSE, fig.cap="Results of the irrigation-nutrient experiment.
-The dashed line refers to the water input, note that the left y-axis was log-transformed.
-The solid lines represent the observed plant coverage in \\% (right y-axis)."}
+```{r irrigation, echo=FALSE, out.width="100%", fig.align="center", warning=FALSE, message=FALSE, fig.cap="Results of the irrigation-ferilization experiment.
+The blue line refers to the water input, note that the left y-axis was log-transformed.
+The solid salmon and the gray dashed lines represent the observed plant coverage in \\% (right y-axis)."}
 # include_graphics(here("figures/17_nutrient_irrigation_experiment.png"))
 load(here("figures/17_nutrient_irrigation_experiment.rda"))
+
+# to plot manually axes labels outside of the panel, we have to disable the
+# clipping beforehand
+my.setting = trellis.par.get("clip")
+my.setting$panel = "off"
+trellis.par.set("clip", my.setting)
+# trellis.par.get("clip")
+
+# plot
 print(p_1)
+```
+
+```{r spri-exp, echo=FALSE, fig.cap="Mean number of species per treatment level plotted for the duration of the experiment.
+Each dot represents a sampling visit during which the number of species was recorded.", fig.align="center"}
+spri_exp = readRDS(here("images/17_irrigation_spri.rds"))
+print(spri_exp)
 ```
@@
 Beta diversity (expressed as the range of the first DCA axis), increased with wetter conditions (Fig. \@ref(fig:dca) & Table \@ref(tab:desc-stats)).
-Naturally, higher species turnover was mainly driven by short-lived, less abundant species.
-Concordantly, @cleland_sensitivity_2013 reports similar findings from temperate grasslands.
-The change of the shape of the DCA point scatter between the relatively humid years (panels 2012 & 2016 of Fig.
+However, another measure of beta diversity, the multivariate dispersion, did not change significantly between years (section \@ref(biodiversity-and-plant-cover)) though the spatial prediction yielded drastic changes in the spatial distribution of the floristic composition and beta diversity (see next section).
+Concordantly, the change of the shape of the DCA point scatter between the relatively humid years (panels 2012 & 2016 of Fig.
 \@ref(fig:dca)) and the very humid neutral year (panel 2017 of Fig.
 \@ref(fig:dca)) can be most likely attributed to the increase of plant cover (see also next paragraph).
+Overall, higher species turnover was mainly driven by short-lived, less abundant species.
+@cleland_sensitivity_2013 reports similar findings from temperate grasslands.
 
 Alpha and gamma diversity increased with higher seasonal precipitation amounts. 
-The relationship, however, is non-linear, i.e., the increase in diversity reached a plateau with the rainfall values recorded under humid La Niña (2012) and under moderate El Niño (2016) conditions. 
+The relationship, however, was non-linear, i.e., the increase in diversity reached a plateau with the rainfall values recorded during the humid La Niña (2012) and moderate El Niño (2016) episodes. 
 Then even more water input as observed during the very humid neutral year (2017) resulted in a slightly decreased diversity but also in a massive and sudden increase of overall plant cover due to higher biomass production.
-Interestingly, only a few grass and forb species were in large parts responsible for this increase.
+Interestingly, only a few grass (e.g., *Aristida adscensionis*, *Eragrostis cilianensis*) and forb species (e.g., *Tiquilia paronychioides*) were in large parts responsible for this increase (see section \@ref(spatio-temporal-mapping)).
+Our irrigation-fertilization experiment confirms the findings of the landscape scale, i.e., the increase in plant cover was much more pronounced than the increase in species richness (see Figs.
+\@ref(fig:irrigation) & \@ref(fig:spri-exp)).
+One explanation might be that under favorable environmental conditions a limited amount of species become the dominant actors of an ecosystem, whereas under intermediate levels of stress and disturbance higher species richness is supported [@brown_why_2014].
+A few dominant species outcompeted other species in 2017 by following a ruderal strategy, i.e., by growing as fast and producing as much offspring as possible in the short period of favorable conditions [@schmidtlein_mapping_2012].
 This is in contrast to other dryland studies which showed that dormant and annual species increased biodiversity under more humid conditions [@gutierrez_variation_2000; @holmgren_extreme_2006].
-On the other hand, this finding is in line with the theory that under favorable environmental conditions a limited amount of species become the dominant actors of an ecosystem, whereas under intermediate levels of stress and disturbance a higher species richness is supported [@brown_why_2014].
-Though obviously a few dominant species outcompeted other species in 2017, they still followed a ruderal strategy by growing as fast and producing as much offspring as possible in the short period of favorable conditions [@schmidtlein_mapping_2012].
 Finally, free-ranging livestock might also have an impact on vegetation dynamics in the study area, especially in its eastern parts with more than 100mm of annual precipitation.
@@
 Still, rainy periods are very conducive to the regeneration of woody vegetation in (semi-)arid areas [@bowers_demographic_1997; @brown_reorganization_1997; @sitters_rainfalltuned_2012].
-For example, the growth and recruitment rate of *Prosopis pallida* is almost twice as high during wet El Niño years than in neutral or dry years in northern Peru [@lopez_climatic_2006].
+For example, the growth and recruitment rate of *Prosopis pallida* is almost twice as high during wet El Niño episodes than during neutral or dry episodes in northern Peru [@lopez_climatic_2006].
 
@@
 
-Though we largely focussed on how current-year precipitation affects biodiversity dynamics, previous year rainfall might also have an effect [@tenhumberg_timelagged_2018].
+The ordination and modeling results of our study provide ample evidence how current-year precipitation affects biodiversity dynamics.
+Still, it might be possible that to a much lesser extent previous year rainfall might also have an effect in the study area [@tenhumberg_timelagged_2018].
 For instance, @dudney_lagging_2017 observed an increase in grass and a decrease in forb abundance when rainfall was high one year earlier.
 Similarly, @walter_plants_2011 suggest that grasses might have a 'drought-memory'.
-Time-lagged responses can be also observed in woody species.
-For example, tree ring growth of *Prosopis pallida* is highly correlated with rainfall of the previous year [@salazar_effect_2018].
 
@@
 
-The spatial prediction map highlights that specific vegetation formations become more pronounced with increasing water input. 
+The spatial prediction maps highlight that specific vegetation formations become more pronounced with increasing water input. 
 In the driest year (2011), a rather uniform distribution of the same floristic composition (green pixels in Fig. \@ref(fig:pred-map)) developed along large parts of the study area. 
@@
 This is because in 2011 rainfall barely occurred along large parts of the study area, and therefore especially perennial species were recorded which can and must endure drought years [@salazar_effect_2018].
-By contrast, during the humid La Niña (2012) and the moderate El Niño (2016) the same three distinct vegetation formations developed along the gradient. 
-This can be explained by similar rainfall patterns though 2012 was a humid La Niña and 2016 a moderate El Niño year. 
-Finally, a large water surplus during the very humid neutral year (2017) led to the disruption of the strictly ordered formations along the humidity gradient. 
-On the one hand, dense grass and herb cover dominated large parts of the study area. The increased cover dominance of annual species is visible in larger negative DCA score values. 
-On the other hand, vegetation patches (yellow, green, and orange pixels) can be found near the coast which are otherwise more typical of the eastern part of the gradient (section \@ref(spatio-temporal-mapping) & Fig.
+By contrast, during the humid La Niña (2012) and the moderate El Niño episodes (2016) the same three distinct vegetation formations developed along the gradient. 
+This can be explained by similar rainfall patterns though 2012 was a humid La Niña and 2016 a moderate El Niño episode. 
+A large water surplus during the very humid neutral episode (2017) led to the disruption of the strictly ordered formations along the humidity gradient. 
+Dense grass and herb cover dominated large parts of the study area. 
+The increased cover dominance of annual species is visible in larger negative DCA score values. 
+Interestingly, vegetation patches (yellow, green, and orange pixels) can be found near the coast which are otherwise more typical of the eastern part of the gradient (section \@ref(spatio-temporal-mapping) & Fig.
 \@ref(fig:pred-map)).
@@
 Another interpretation could be also that the NDVI values, our main predictor for the floristic composition, is similar for the dense cover of annual plants and perennial woody plants in 2017 [@pena_assessing_2015; @salazar_intraspecific_2018].
+
+Overall, our spatial mapping approach concedes detailed insights into how floristic composition and beta diversity distribution patterns change in response to varying rainfall patterns caused by different ENSO episodes.
+Similarly, @lovelace_geocomputation_2019a (chapter 13) predicted floristic composition on a Peruvian fog oasis which revealed that the vegetation belts along the mountain slope are partly interrupted.
+Consequently, predictive mapping can contribute to a more nuanced understanding of ecosystem dynamics.
+Naturally, such insights cannot be captured by global beta diversity measures such as the analysis of multivariate homogeneity of groups dispersions which in our case did not yield significant results (see sections \@ref(biodiversity-and-plant-cover) & \@ref(biodiversity-and-biomass)).
+
+Finally, the spatial prediction of species richness (Appendix 4) supports the findings of the floristic gradient mapping since it revealed similar spatial diversity patterns and dynamics.
 
@@
 \@ref(fig:varpart)).
-Interestingly, the shared explained variance in the floristic dataset is roughly the same for the years experiencing wetter conditions (humid La Niña (2012), moderate El Niño (2016) and the very humid neutral year (2017; coastal El Niño)) though 2017 was the most humid by far.
+Interestingly, the shared explained variance in the floristic dataset is roughly the same for the years experiencing wetter conditions (humid La Niña episode (2012), moderate El Niño epidose (2016) and the very humid neutral episode (2017; coastal El Niño) though 2017 was the most humid by far.
 This indicates that there is an upper limit to the beneficial effects of the water-soil and water-topography interaction.
 Topographic variables such as curvature and catchment area play some role in structuring the vegetation composition in the study area especially in dry years [@muenchow_coupling_2013; @muenchow_woody_2013].
-Similar findings could be also observed in Mexico where topography also partly explained the composition of dry forests [@mendez-toribio_effects_2016].
+Similar findings were observed in Mexico where topography also partly explained the composition of dry forests [@mendez-toribio_effects_2016].
 However, the most important topographic variable in the study area was altitude since it is highly correlated with mean annual precipitation. 
@@
 
-Naturally, nutrients can play a major role in drylands [@ronnenberg_effects_2011; @whitford_effects_2011] as confirmed by our irrigation-nutrient experiment.
-In fact, the water-nitrogen interaction can result in a biomass production three times as high as the production achieved when adding the same amount of water but without additional nutrients (Fig.
+Nutrients can play a major role for vegetation development in drylands [@ronnenberg_effects_2011; @whitford_effects_2011] as confirmed by our irrigation-fertilization experiment.
+In fact, the water-nitrogen interaction can result in a plant coverage three times as high as achieved when adding the same amount of water but without additional nutrients (Fig.
 \@ref(fig:irrigation)).
-This is of utmost importance for sustainable yet productive agricultural management. 
-In combination with a reliable climate prediction, it supports the decision on which crops (annual herbal and grass species) to grow and how much fertilizer to use.
+This is of utmost importance for sustainable yet productive agricultural management in the study area, where farming activity increases along the gradient.
+Hence, instead of making disappear the last remnants of TDF in the study area, it would be desirable to increase the productivity of already existing farmland.
+The decision on which crops (annual herbal and grass species) to grow and how much fertilizer to use could be further improved by reliable climate predictions on the occurrence of ENSO episodes.
 Other studies have also confirmed the importance of soil for the composition of the vegetation in drylands.
@@
 
+<!--
 ## ENSO, climate change and ENSO indices
@@
 Climate change will most likely increase the frequency and intensity of El Niño events [@cai_increasing_2014; @IPCC.2014; @wang_continued_2017].
-This and climate change in general pose the risk to turn tropical forests - dry and moist - from carbon sinks to sources [@phillips_changes_2009;@alvarez-davila_forest_2017;@salazar_tree_2019]. 
-This is also true for the study area [@salazar_tree_2019] where El Niño is frequently associated with higher rainfall and La Niña with increased aridity.
+This and climate change in general pose the risk to turn tropical forests - dry and moist - from carbon sinks to sources [@phillips_changes_2009;@alvarez-davila_forest_2017;@salazar_tree_2020]. 
+This is also true for the study area [@salazar_tree_2020] where El Niño is frequently associated with higher rainfall and La Niña with increased aridity.
 However, both can also cause the opposite effects. 
@@
 The first is the Índice Costero El Niño (ICEN; see http://www.met.igp.gob.pe/publicaciones/2014/ElNino_v2_27_03_2014_7.pdf), and the second is an ENSO index for the insurance of agricultural products [@mortensen_enso_2018].
+-->
 
@@
 
+Alpha and gamma diversity increased with wetter conditions during the humid La Niña (2012) and moderate El Niño (2016) episodes, though, a few dominant species largely prevailed under the very humid conditions of the coastal coastal El Niño episode (2017), which even resulted in a slightly decreased alpha diversity.
+Still, plant cover was by far greatest during the most humid conditions.
+Beta diversity also increased with wetter conditions but did not change significantly between years in accordance with the test for multivariate homogeneity of groups dispersions.
+As useful as such global measures are in general, they are in essence a single descriptive figure which naturally lack the capability of capturing spatial diversity patterns.
+By contrast, our spatial predictions revealed drastic changes in the distribution of species composition, beta diversity and plant cover.
+In effect, a large water surplus disrupted the strictly ordered vegetation formations which at first became apparent along the climatic gradient with wetter conditions.
+Hence, spatial predictions might provide more detailed insights into ecosystem dynamics, especially in semi-arid ecosystems where changes in plant cover and distribution patterns are more pronounced than changes in alpha and gamma diversity.
+
+The influence of topographic and edaphic variables on plant diversity and productivity also gained in importance with wetter conditions but this effect was not further amplified with very humid conditions.
+Our irrigation-fertilization experiment partly confirms these findings but also revealed the true beneficial effect of fertilizing under wet Super Niño conditions when plant cover was three times as high as in the unfertilized plots.
+
+In conclusion, our predictive mapping approach can identify locations suitable for restoration and conservation, and thus could form the basis for an informed conservation management (forest management plans).
+At the same time, the results of our experiment can equip farmers with information regarding sustainable agrarian management which might increase the productivity on already existing farmland which in turn might help to prevent the destruction of the last remnants of TDF in the study area.
+
+<!--
 Constant monitoring is a prerequisite for a better understanding of ENSO-affected ecosystems. 
@@
 Additionally, our results clearly reveal patterns in the spatial variability of biomass production in response to water availability along the climatic gradient, which might be also an important information for agrarian and restoration activities.
+-->
 

### 2020-08-10 text/ecography/round_3/supplementary_information.Rmd
migrate gitlab repo to github

--- 2020-08-10 text/ecography/round_2/supplementary_information.Rmd
@@
     toc: false
-    fig_caption: no
+    fig_caption: yes
     number_sections: true
@@
   - \linenumbers
+  # Use Appendix instead of Figure in the caption
+  - \renewcommand{\figurename}{Appendix}
 
 title: Supplementary Information
-bibliography: transect_refs.bib
-csl: ecography.csl
+bibliography: ../transect_refs.bib
+csl: ../ecography.csl
 # biblio-style: "natbib"
@@
 
-```{r sampling-zones, echo=FALSE, fig.cap="Study area stratified by distance to sea in 10 classes."}
+```{r sampling-zones, echo=FALSE, fig.align="center",fig.cap="Study area stratified by distance to sea in 10 classes."}
 p_1 = readRDS(here("images/18_sample_stratification.rds"))
@@
 
-We established 12 experimental plots of 3 × 3 m^2 in a sandy area typical of the region.
+We established 12 experimental plots of 3 × 3 m^2^ in a sandy area typical of the region.
 Plots were located on the campus of the Universidad de Piura (Sechura Desert; 5°10'S, 80°38'W) on a sandy plain devoid of woody vegetation.
@@
 
-\subsection*{Appendix 3. Number of species plotted against time during the duration of the experiment.}
+\subsection*{Appendix 3. Relationship between the response and NDVI by year.}
 
-```{r alpha-exp, echo=FALSE, fig.cap="Number of species plotted against time during the duration of the experiment.", message=FALSE}
-p_1 = readRDS(file = here("images/18_irrigation_spri.rds"))
-print(p_1)
+<!-- since we have no Figure in Appendix 2, we have to increase the figure counter by 1, then our spatial prediction will correctly displayed as Appendix 3-->
+\setcounter{figure}{2}
+
+```{r dca-ndvi, echo=FALSE, message=FALSE, fig.align="center", fig.cap="Relationship between the response (DCA1) and NDVI by year.
+To aid the visual interpretation, line smoothers were added.", out.width="70%"}
+data = readRDS(here("images/15_model_input_data.rds"))
+ggplot(data, aes(x = ndvi, y = DCA1, group = year)) +
+  geom_point(aes(color = as.factor(year)), alpha = 0.4) + 
+  labs(color = "year") +
+  xlab(label = "NDVI") +
+  ylab(label = "DCA1") +
+  geom_smooth(aes(color = as.factor(year)), se = FALSE) + 
+  theme_bw() +
+  coord_fixed(ratio = 0.15)
+```
+
+\subsection*{Appendix 4. Predictive mapping of species richness.}
+
+```{r spri-dist, echo=FALSE, out.width="100%", fig.align="center", message=FALSE, warning=FALSE, fig.cap="Spatial prediction of species richness along the cross-section for different ENSO episodes using the NDVI and year interaction as predictors (used model: GAM with a spline smoother with 3 degrees of freedom and a Poisson distribution).
+The urban boundaries of the cities Paita, Piura and Chulucanas are displayed in black.
+The Piura River and its tributaries are represented by light blue lines with a black contour."}
+library("sp")
+library("sf")
+load(here("figures/15_spri_preds.rda"))
+print(spri_ndvi)
 ```

### 2020-08-10 text/ecography/round_3/title_page.Rmd
migrate gitlab repo to github

--- 2020-08-10 text/ecography/round_2/title_page.Rmd
@@
 
-Though tropical dry forests (TDF) are the most endangered tropical ecosystem, they are still understudied.
-Therefore, we studied the spatio-temporal effects of different El Niño Southern Oscillation (ENSO) episodes on TDF biodiversity in the terrestrial core region of ENSO in NW Peru.
-For this, we sampled 50 permanent vegetation plots along a climatic gradient in four years representing different ENSO episodes: dry La Niña (2011), humid La Niña (2012), moderate El Niño (2016) and a very humid neutral year (coastal El Niño; 2017).
-To visually represent the spatio-temporal changes in floristic composition, we used an innovative approach by modeling the ordination scores using a Generalized Additive Model and spatial cross-validation.
-Additionally, we applied variation partitioning to test if topographic or edaphic variables gained in importance as soon as water was no longer the limiting factor.
-Plant species richness increased under humid conditions during the humid La Niña (2012) and moderate El Niño (2016) year, and slightly decreased under the very humid conditions of the coastal El Niño year (2017).
-Vegetation composition changed also under more humid conditions.
-Vegetation cover was manifold higher in 2017 compared to all previous years.
-Edaphic and topographic variables gained in importance with increased water availability in 2012 and 2016, however, this effect was not further amplified under very wet conditions (2017).
-As nutrients can be equally important for primary production as water, we additionally executed an irrigation-nutrient experiment.
-The experiment showed that biomass production under wet Super Niño conditions was three times higher when a fertilizer treatment was added.
+The tropical dry forests of NW Peru are heavily shaped by the El Niño Southern Oscillation (ENSO), where especially El Niño brings rain to arid to semi-arid areas.
+However, the resulting effects on biodiversity patterns remain largely unknown as well as the effect of environmental variables on the floristic composition under varying rainfall patterns.
+Therefore, we studied the spatio-temporal effects of different ENSO episodes on floristic biodiversity along a climatic gradient ranging from the coastal desert to the Andean foothills.
+We sampled 50 vegetation plots in four years representing different ENSO episodes.
+To highlight the spatio-temporal changes in floristic composition and beta diversity across ENSO episodes, we predicted ordination scores with a Generalized Additive Model.
+We applied variation partitioning to test if topographic or edaphic variables gained in importance during more humid ENSO episodes.
+Additionally, we executed an irrigation-fertilization experiment to quantify the beneficial effects of the water-nutrient interaction under different simulated ENSO rainfall scenarios. 
+Plant species richness increased under humid conditions during the humid La Niña (2012) and the moderate El Niño (2016), and slightly decreased under the very humid conditions during the coastal El Niño (2017).
+The spatial prediction revealed that specific vegetation formations became more pronounced with increasing water input, but that a large water surplus led to the disruption of the strict order along the climatic gradient.
+Edaphic and topographic variables gained in importance with increased water availability (2012 & 2016), however, this effect was not further amplified under very wet conditions (2017).
+The experiment showed that plant cover under Super Niño conditions was three times higher when fertilized.
+Overall, our spatial predictions concede detailed insights into spatio-temporal ecosystem dynamics in response to varying rainfall caused by different ENSO episodes while the results of the experiment can support farmers regarding a sustainable agrarian management.
+
+<!--
 Overall, our results can help to 1) protect local human populations from devastating ENSO effects through informed conservation management and the restoration of degraded ecosystems; 2) equip farmers with information regarding sustainable agrarian management.
+-->
 
@@
 
-*Funding* - We are also most thankful to the Deutsche Forschungsgemeinschaft (project Ri 370/19-1) and a Friedrich Schiller University of Jena program to support junior researchers (project DRM/2017-05) for funding this work.
+*Funding* - We are also most thankful to the Deutsche Forschungsgemeinschaft (project Ri 370/19-1), a Friedrich Schiller University of Jena program to support junior researchers (project DRM/2017-05) and the project "Ecología Funcional de los Bosques Andaluces y Predicciones Sobre Sus Cambios Futuros (For-Change)" (UCO-27943) by Junta de Andalucía (Spain) and European FEDER funds for funding this work.
 
 *Author contributions* - JM, MR and AJ conceived the ideas.
-JM, TB, DJ, RAR and PSZ conducted the ﬁeldwork and collected the data with the help of collaborators.
+JM, TB, DJ, RAR and PSZ conducted the ﬁeld work and collected the data with the help of collaborators.
 JM, PD, AB, TF, JB, GD, RR, DJ, PS, DN and EFR analyzed the data.

### 2020-08-10 code/analysis/00_modis_download.R

@@ -1,0 +1,201 @@
+# Filename: 03_modis_download.R (2018-02-22)
+#
+#
+# Author(s): Jannes Muenchow, Gregor Didenko
+#
+#**********************************************************
+# CONTENTS-------------------------------------------------
+#**********************************************************
+#
+# 1. (INSTALLATION) SET UP
+# 2. DOWNLOAD AND PREPROCESS MODIS
+# 3. LOAD MODIS FILES FROM DISK AND CREATE MEAN VALUE RASTERS
+# 4. CREATE NDVI AT Plots.RDS
+
+#**********************************************************
+# 1 ATTACH PACKAGES AND (INSTALLATION) SET UP--------------
+#**********************************************************
+
+# create an Earthdata account:
+# https://urs.earthdata.nasa.gov/home
+
+# optional: download MODIS Reprojection Tool (MRT) from
+# https://lpdaac.usgs.gov/tools/modis_reprojection_tool
+
+library("MODIS")
+library("dplyr")
+library("sf")
+library("raster")
+library("RSQLite")
+
+# Note: Download GDAL from http://www.gisinternals.com/release.php
+# specify path to GDAL and paths
+# to enable hdf4 support on Arch, run trizen -s gdal-hdf4
+
+MODISoptions(
+  # path to GDAL on your computer, only necessary for Windows
+  # gdalPath = "C:/Program Files/GDAL",
+  # gdalPath = "C:/OSGeo4W64/bin",
+  # where to save downloaded MODIS data
+  localArcPath = "data/raw_data/modis",
+  # where to save processed (projected, cropped and resampled) data
+  outDirPath = "data/raw_data/modis/processed",
+  checkTools = TRUE,
+  MODISserverOrder = c("LPDAAC", "LAADS"),
+  dlmethod = "auto")
+# save options for later on
+opts = MODISoptions()
+
+# optional:
+# assuming the MRT tools are installed in C:/modis_mrt
+# path = Sys.getenv("PATH")
+# Sys.setenv("PATH" = paste(path, "C:/modis_mrt/bin", sep = ";"))
+# Sys.setenv("MRT_HOME" = "C:/modis_mrt")
+# Sys.setenv("MRT_DATA_DIR" = "C:/modist_mrt/data")
+# system("mrtmosaic")  # ok, should work
+# # system("C:/modis_mrt/bin/mrtmosaic")  
+# # check
+# MODIS:::checkTools('MRT')
+
[…]

### 2020-08-10 code/analysis/11_prep_data.R

@@ -1,0 +1,406 @@
+# Filename: 11_preparation.R (2018-06-07)
+#
+# TO DO: bring data in a form ready for analysis
+#
+# Author(s): Jannes Muenchow, Jonas Brock
+#
+#**********************************************************
+# CONTENTS-------------------------------------------------
+#**********************************************************
+#
+# 1. ATTACH PACKAGES
+# 2. VEGETATION DATA PREPROCESSING
+# 3. CREATE CROSS-TABLES
+# 4. LONDO
+# 5. ENIRONMENTAL DATA
+#
+#**********************************************************
+# 1 ATTACH PACKAGES AND DATA-------------------------------
+#**********************************************************
+
+# attach packages
+library("data.table")
+library("tidyverse")
+library("magrittr")
+library("gdata")
+library("reshape2")
+library("RSQLite")
+library("forcats")
+library("here")
+
+# attach data
+con = dbConnect(RSQLite::SQLite(), here("data/tables.gpkg"))
+dbListTables(con)
+lf = dbReadTable(con, "lifeform", as.is = TRUE)
+d_11 = dbReadTable(con, "plot_species_matrix_2011", as.is = TRUE)
+d_12 = dbReadTable(con, "plot_species_matrix_2012", as.is = TRUE)
+d_16 = dbReadTable(con, "plot_species_matrix_2016", as.is = TRUE)
+d_17 = dbReadTable(con, "plot_species_matrix_2017", as.is = TRUE)
+soil = dbReadTable(con, "soil", as.is = TRUE)
+topo = dbReadTable(con, "topography", as.is = TRUE)
+plova = dbReadTable(con, "plot_variables", as.is = TRUE)
+dbDisconnect(con)
+
+#**********************************************************
+# 2 DATA PREPROCESSING-------------------------------------
+#**********************************************************
+
+# 2.1 Prepare 2011 data====================================
+#**********************************************************
+
+setdiff(1:50, d_11$pnr)
+# missing plots (no vegetation found): 2, 18, 20
+
+# classify manually so far uncategorized species
+d_11 = as.data.table(d_11)
+d_11[art == "Acac_maca", type := "woody"]
[…]

### 2020-08-10 code/analysis/12_study_area_ndvi.R

@@ -1,0 +1,393 @@
+# Filename: 12_study_area_ndvi.R (2019-10-15)
+
+# TO DO: Create study area map including an inset map, a ternary plot and
+#        precipitation rain bars
+
+# Author(s): Gregor Didenko, Jannes Muenchow, Patrick Schratz
+
+#**********************************************************
+# CONTENTS-------------------------------------------------
+#**********************************************************
+
+# 1. ATTACH PACKAGES AND DATA
+# 2. STUDY AREA (NDVI BACKGROUND)
+# 3. OVERVIEW MAP
+# 4. TERNARY PLOT
+# 5. PRECIPITATION BARPLOT
+# 6. GRID OF PLOTS
+
+#**********************************************************
+# 1 ATTACH PACKAGES AND DATA-------------------------------
+#**********************************************************
+
+# attach packgaes
+library("cowplot")
+library("data.table")
+library("dplyr")
+library("ggpubr")
+library("gridGraphics")
+library("here")
+library("osmdata")
+library("lattice")
+library("png")
+library("raster")
+library("RCurl")
+library("RSQLite")
+library("sf")
+library("Ternary")
+library("magrittr")
+
+# attach data
+# get NDVI Rasters
+ndvi_rasters = readRDS("images/00_ndvi_rasters.rds")
+## get streets
+con = dbConnect(SQLite(), "data/tables.gpkg")
+dbListTables(con)
+# set all extents to raster-extent
+streets = st_read(con, "streets") %>%
+  st_transform(crs = st_crs(ndvi_rasters)) %>%
+  st_crop(st_bbox(ndvi_rasters))
+street = streets[1, ]
+
+## get town polygons
+towns = st_read(con, "towns_jvs") %>%
+  st_transform(crs = st_crs(ndvi_rasters)) %>%
+  st_crop(st_bbox(ndvi_rasters))
+# get center of towns-polygon
+towns_label = st_centroid(towns) %>%
+  st_coordinates() %>%
+  as.data.frame()
+# adjust for better visualibility
+towns_label[1, 2] = towns_label[1, 2] + 0.024
+towns_label[3, 2] = towns_label[3, 2] - 0.005
[…]

### 2020-08-10 code/analysis/13_descriptive_stats.R

@@ -1,0 +1,195 @@
+# Filename: 13_descriptive_stats.R (2018-06-21)
+#
+# TO DO: Descriptive stats of biodiversity and plant cover
+#
+# Author(s): Jannes Muenchow, Jonas Brock
+#
+#**********************************************************
+# CONTENTS-------------------------------------------------
+#**********************************************************
+#
+# 1. ATTACH PACKAGES AND DATA
+# 2. VISUALIZE CHANGE IN RICHNESS & COVER
+# 3. ALPHA, BETA, GAMMA-DIVERSITY
+#
+#**********************************************************
+# 1 ATTACH PACKAGES AND DATA-------------------------------
+#**********************************************************
+
+# attach packages
+library("data.table")
+library("dplyr")
+library("ggplot2")
+library("gridExtra")
+library("lattice")
+library("latticeExtra")
+library("here")
+library("vegan")
+library("magrittr")
+# source your own functions
+source(here("code/funs/ordi_helper_funs.R"))
+
+# attach cover datasets (d = Deckung, i.e., cover)
+d_11 = readRDS(here("images/11_d_11.rds")) %>%
+  mutate(year = 2011)
+d_12 = readRDS(here("images/11_d_12.rds")) %>%
+  mutate(year = 2012)
+d_16 = readRDS(here("images/11_d_16.rds")) %>%
+  mutate(year = 2016)
+d_17 = readRDS(here("images/11_d_17.rds")) %>%
+  mutate(year = 2017)
+# attach londo cover data
+londo = readRDS(here("images/11_ct_londo.rds"))
+# envdata
+env = readRDS(here("images/11_env_data.rds"))
+
+#**********************************************************
+# 2 VISUALIZE CHANGE IN RICHNESS & COVER-------------------
+#**********************************************************
+
+# 2.1 Data preparation=====================================
+#**********************************************************
+d = data.table::rbindlist(list(d_11, d_12, d_16, d_17), use.names = TRUE)
+d[, cover := as.numeric(as.character(cover))]
+d_2 = d[, .(cover = sum(cover),
+            spri = .N),
+        by = .(pnr, year)]
+# find out about missing ids in groups
+tmp = tibble(pnr = rep(1:50, 4),
+             year = rep(c(2011, 2012, 2016, 2017), each = 50))
+anti_join(tmp, d_2)
+d_2 = full_join(d_2, tmp) %>%
[…]

### 2020-08-10 code/analysis/14_ordination.R

@@ -1,0 +1,288 @@
+# Filename: 14_ordination.R (2018-06-08)
+#
+# TO DO: Find out which ordination yields the "best" and most interpretable 
+#        results
+#
+# Author(s): Jannes Muenchow, Jonas Brock
+#
+#**********************************************************
+# CONTENTS-------------------------------------------------
+#**********************************************************
+#
+# 1. ATTACH PACKAGES AND DATA
+# 2. DCA
+# 3. NMDS
+# 4. ISOMAP
+# 5. WINNER: DCA LONDO
+# 6. DCA VIZ
+#
+#**********************************************************
+# 1 ATTACH PACKAGES AND DATA-------------------------------
+#**********************************************************
+
+# attach packages
+library("vegan")
+library("dplyr")
+library("lattice")
+library("magrittr")
+library("ggplot2")
+library("here")
+
+# source own functions
+source("code/funs/ordi_helper_funs.R")
+source("code/funs/bestisomap.R")
+# attach data
+pa = readRDS("images/11_ct_pa.rds")
+# Remember, there are 196 obs instead of 2000 because:
+# 1: in 2011 there were 3 plots (2, 18, 20) without any vegetation cover
+# 2: in 2017 plot 25 was inundated and could not be visited
+cover = readRDS("images/11_ct_cover.rds")
+londo = readRDS("images/11_ct_londo.rds")
+env_data = readRDS("images/11_env_data.rds")
+
+#**********************************************************
+# 2 DCA----------------------------------------------------
+#**********************************************************
+
+# 2.1 presence-absence=====================================
+#**********************************************************
+
+dca = ord_fun(pa, env_data, method = "dca")
+# visualize for all years
+sc = as.data.frame(scores(dca))
+sc %<>% 
+  mutate(year = gsub(".*_", "", rownames(sc)),
+         id = gsub("_.*", "", rownames(sc)))
+           
+multi_panel_plot(DCA1, DCA2, by = year, data = sc)
+
+# just checking
+# fil = filter(sc, year == 2016)
+# plot(sc$DCA1, sc$DCA2)
+# points(fil$DCA1, fil$DCA2, pch = 16, col = "black")
[…]

### 2020-08-10 code/analysis/15_modeling.R

@@ -1,0 +1,446 @@
+# Filename: 15_modeling.R (2019-15-10)
+#
+# TO DO: model DCA scores as a function of NDVI and year, spatial cv, 
+#        predictive mapping
+#
+# Author(s): Jannes Muenchow, Jonas Brock
+#
+#**********************************************************
+# CONTENTS-------------------------------------------------
+#**********************************************************
+#
+# 1. ATTACH PACKAGES AND DATA
+# 2. DATA PREPARATION
+# 3. DATA EXPLORATION
+# 4. MODELING AND SPATIAL CV
+# 5. PREDICTIVE MAPPING
+#
+#**********************************************************
+# 1 ATTACH PACKAGES AND DATA-------------------------------
+#**********************************************************
+
+# attach packages
+library("dplyr")
+library("gdata")
+library("lattice")
+library("latticeExtra")
+library("mgcv")
+library("raster")
+library("rasterVis")
+library("gridExtra")
+library("sperrorest")
+library("sf")
+library("RSQLite")
+library("here")
+
+# source your own funs
+source("code/funs/spcv_helper_funs.R")
+
+# attach raster stack
+ndvi = readRDS("images/00_ndvi_rasters.rds")
+# attach response (DCA scores)
+data = readRDS("images/14_dca_scores.rds")
+data$id = as.numeric(data$id)
+# attach env_data
+env_data = readRDS("images/11_env_data.rds")
+env_data$year = gsub(".*_", "", env_data$pnr)
+# attaching spatial data
+con = dbConnect(SQLite(), "data/tables.gpkg")
+rivers = st_read(con, "rivers")
+towns = st_read(con, "towns_jvs")
+dbDisconnect(con)
+
+#**********************************************************
+# 2 DATA PREPARATION---------------------------------------
+#**********************************************************
+
+# add xy-coordinates to dca scores
+data = inner_join(data, 
+                  dplyr::select(env_data, id, year, x, y, lon, lat, slope, 
+                                vcurv),
+                  by = c("id", "year"))
+# extract NDVI to points
+data$ndvi = NA
+for (i in c(11, 12, 16, 17)) {
+  ind = grepl(i, data[, "year"]) 
+  data[ind, "ndvi"] = 
[…]

### 2020-08-10 code/analysis/16_varpart.R

@@ -1,0 +1,161 @@
+# Filename: 16_varpart.R (2019-10-15)
+#
+# TO DO: variation partitioning of DCA axes
+#
+# Author(s): Jannes Muenchow
+#
+#**********************************************************
+# CONTENTS-------------------------------------------------
+#**********************************************************
+#
+# 1. ATTACH PACKAGES AND DATA
+# 2. ENV DATA EXPLORATION & TRANSFORMATION
+# 3. VARIATION PARTITIONING
+#
+#**********************************************************
+# 1 ATTACH PACKAGES AND DATA-------------------------------
+#**********************************************************
+
+# attach packages
+library("vegan")
+library("dplyr")
+library("lattice")
+
+# read in functions
+source("code/funs/HighstatLib.R")  # corvif
+source("code/funs/varpart_helper_funs.R")
+
+# attach data
+ct_londo_11 = readRDS("images/11_ct_londo_11.rds")
+ct_londo_12 = readRDS("images/11_ct_londo_12.rds")
+ct_londo_16 = readRDS("images/11_ct_londo_16.rds")
+ct_londo_17 = readRDS("images/11_ct_londo_17.rds")
+# dca = readRDS("images/15_scores_dca.rds")
+# env_data = readRDS("images/11_env_data.rds")
+soil = readRDS("images/11_soil.rds")
+topo = readRDS("images/11_topo.rds")
+
+#**********************************************************
+# 2 ENV DATA EXPLORATION & TRANSFORMATION------------------
+#**********************************************************
+
+# have a look at the soil variables
+tmp = reshape2::melt(soil, id.var = "pnr")
+xyplot(value ~ pnr | variable, data = tmp, pch = 16,
+       col = "salmon",
+       scales = list(relation = "free"),
+       panel = function(x, y, ...) {
+         panel.xyplot(x, y, ...)
+         panel.loess(x, y, lwd = 2.5, col = "gray")
+       })
+
+soil = mutate(soil,
+              log_lf = log(lf_s),
+              log_cn = log(cn + 1)) %>%
+  dplyr::select(-lf_s, -cn)
+# remove collinear variables
+corvif(dplyr::select(soil, -pnr, -sand, -log_lf, -p))
+soil = dplyr::select(soil, -sand, -log_lf, -p)
+
[…]

### 2020-08-10 code/analysis/17_irrigation_nutrient_experiment.R

@@ -1,0 +1,213 @@
+# Filename: 17_irrigation_nutrient_experiment.R (2019-10-15)
+
+# TO DO: Analyze irrigation-nutrient experiment data
+
+# Author(s): Jannes Muenchow
+
+#**********************************************************
+# CONTENTS-------------------------------------------------
+#**********************************************************
+
+# 1. ATTACH PACKAGES AND DATA
+# 2. DATA PREPARATION
+# 3. IRRIGATION PLOT
+
+#**********************************************************
+# 1 ATTACH PACKAGES AND DATA-------------------------------
+#**********************************************************
+
+# attach packages
+library("RSQLite")
+library("lattice")
+library("dplyr")
+library("here")
+
+# attach data
+con = dbConnect(SQLite(), "data/tables.gpkg") 
+dbListTables(con)
+# species counted per observation
+count = dbReadTable(con, "experiment_count")  # d
+# irrigation/nutrient input data
+irr = dbReadTable(con, "experiment_irrigation")  # rain
+# cover per plot in %
+cover = dbReadTable(con, "experiment_cover")  # pcov
+dbDisconnect(con)
+
+#**********************************************************
+# 2 DATA PREPARATION---------------------------------------
+#**********************************************************
+
+count = count %>%
+  mutate_at(c("plot_name", "abb"), as.factor) %>%
+  mutate(obs_date = as.Date(obs_date))
+
+cover = cover %>%
+  mutate(obs_date = as.Date(obs_date),
+         plot_name = as.factor(plot_name))
+
+names(irr) = tolower(names(irr))
+
+irr = irr %>%
+  mutate_at(c("nn", "sn", "prec"), as.numeric) %>%
+  mutate(id = as.Date(id))
+
+#**********************************************************
+# 3 IRRIGATION PLOT----------------------------------------
+#**********************************************************
+
+# monthly precipitation in Piura from December 2012 to May 2013
+# group_by(irr, as.factor(format(irr$id, "%b"))) %>%
+#   summarize(sum(prec))
+
+# irrigation dataframe (natural precipitation, normal Nino water input and 
+# super Nino water input)
+prec = data.frame(prec = c(cumsum(irr$prec), cumsum(irr$nn), cumsum(irr$sn)))
+prec$date = rep(irr$id, 3) 
+prec$enso = as.factor(rep(c("prec", "nn", "sn"), each = nrow(irr)))
[…]

### 2020-08-10 code/analysis/18_appendix.R

@@ -1,0 +1,74 @@
+# Filename: 18_appendix.R (2019-11-20)
+
+# TO DO: Figures found in the Supplementary Information
+
+# Author(s): Jannes Muenchow  
+
+#**********************************************************
+# CONTENTS-------------------------------------------------
+#**********************************************************
+
+# 1. ATTACH PACKAGES AND DATA
+# 2. 
+
+#**********************************************************
+# 1 ATTACH PACKAGES AND DATA-------------------------------
+#**********************************************************
+
+# attach packages
+library("lattice")
+library("here")
+library("sf")
+library("sp")
+library("RSQLite")
+library("tidyverse")
+library("tidyr")
+
+# attach data
+alpha = readRDS(here("images/17_alpha.rds"))
+# get NDVI Rasters
+ndvi_rasters = readRDS(here("images/00_ndvi_rasters.rds"))
+# get vector data
+con = dbConnect(SQLite(), here("data/tables.gpkg"))
+coast = st_read(con, "coast")
+study_area = st_read(con, "study_area")
+dbDisconnect(con)
+
+#**********************************************************
+# APPENDIX 2-----------------------------------------------
+#**********************************************************
+
+study_area = st_transform(study_area, 32717)
+r = raster(as(study_area, "Spatial"), res = 100, 
+           vals = 1)
+dist = distanceFromPoints(r, st_coordinates(coast)[, 1:2])
+# plot(mask(dist, study_area))
+r_range = cellStats(dist, range)
+classes = cut(dist, seq(r_range[1], r_range[2], length.out = 11))
+# make sure that object classes has the same prj, extent, origin
+classes = projectRaster(classes, crs = proj4string(ndvi_rasters))
+classes = resample(classes, ndvi_rasters)
+# set Pacific Ocean values to NA
+classes[is.na(ndvi_rasters$ndvi_12)] = NA
+coast = st_transform(coast, 4326)
+# plot(classes)
+# plot(crop(as(coast, "Spatial"), classes), add = TRUE)
+sp_1 = spplot(classes, scales = list(draw = TRUE, tck = c(1, 0)))
+
+saveRDS(sp_1, file = here("images/18_sample_stratification.rds"))
+saveRDS(dist, file = here("images/18_dist_sea_raster.rds"))
+
+#**********************************************************
+# APPENDIX 3-----------------------------------------------
+#**********************************************************
+
+invisible(Sys.setlocale("LC_TIME", "C"))
[…]

### 2020-08-10 code/funs/HighstatLib.R

@@ -1,0 +1,167 @@
+
+##################################################################
+#Below are some functions that we took from the pairs help file and
+#modified, or wrote ourselves. You need to copy and paste all these commands
+#into R. It is perhaps better not to try and understand what it all does.
+
+
+
+## put correlations on the panels,
+## with size proportional to the correlations.
+panel.cor <- function(x, y, digits=1, prefix="", cex.cor)
+{
+    usr <- par("usr"); on.exit(par(usr))
+    par(usr = c(0, 1, 0, 1))
+    r1=cor(x,y,use="pairwise.complete.obs")
+    r <- abs(cor(x, y,use="pairwise.complete.obs"))
+
+    txt <- format(c(r1, 0.123456789), digits=digits)[1]
+    txt <- paste(prefix, txt, sep="")
+    if(missing(cex.cor)) cex <- 0.9/strwidth(txt)
+    text(0.5, 0.5, txt, cex = cex * r)
+}
+
+panel.smooth2=function (x, y, col = par("col"), bg = NA, pch = par("pch"),
+    cex = 1, col.smooth = "red", span = 2/3, iter = 3, ...)
+{
+    points(x, y, pch = pch, col = col, bg = bg, cex = cex)
+    ok <- is.finite(x) & is.finite(y)
+    if (any(ok))
+        lines(stats::lowess(x[ok], y[ok], f = span, iter = iter),
+            col = 1, ...)
+}
+
+
+panel.lines2=function (x, y, col = par("col"), bg = NA, pch = par("pch"),
+    cex = 1, ...)
+{
+    points(x, y, pch = pch, col = col, bg = bg, cex = cex)
+    ok <- is.finite(x) & is.finite(y)
+    if (any(ok)){
+        tmp=lm(y[ok]~x[ok])
+        abline(tmp)}
+
+}
+
+
+
+
+panel.hist <- function(x, ...)
+{
+    usr <- par("usr"); on.exit(par(usr))
+    par(usr = c(usr[1:2], 0, 1.5) )
+    h <- hist(x, plot = FALSE)
+    breaks <- h$breaks; nB <- length(breaks)
+    y <- h$counts; y <- y/max(y)
+    rect(breaks[-nB], 0, breaks[-1], y, col="white", ...)
+}
+
+
+
+#VIF
+myvif <- function(mod) {
+  v <- vcov(mod)
+  assign <- attributes(model.matrix(mod))$assign
[…]

### 2020-08-10 code/funs/bestisomap.R

@@ -1,0 +1,154 @@
+bestisomap <- function (distance, kmin=3, ndim=3, criterion="overall") {
+
+################################################################################
+##                                                                            ##
+## "bestisomap" finds the Isomap-solution offering a maximum of explained     ##
+## variance on all or the first dimsension. Isomap is calculated using the    ##
+## vegan-package (Oksanen, Kindt, Legendre, O'Hara, Simpson, Solymos,         ##
+## Stevens, & Wagner, 2009, http://vegan.r-forge.r-project.org/)              ##
+##                                                                            ##
+##                          VERSION 1.0                                       ##
+##                                                                            ##
+## Author: Hannes Feilhauer, Vegetation Geography, University of Bonn         ##
+##                                                                            ##
+## ARGUMENTS                                                                  ##
+##                                                                            ##
+## distance   object containing inter-sample dissimilarities, e.g.,           ##
+##            output from vegdist ()                                          ##
+## kmin       smalles k-value as starting point; k determines the number of   ##
+##            nearest neightbors considered in Isomap. Small k-values cause   ##
+##            a non-linear solution, growing k-values increase linearity      ##
+## ndim       dimensionality of the Isomap-space                              ##
+## criterion  "overall" finds the solution with maximized explained variance  ##
+##            across all dimensions defined by ndim (recommended), "first"    ##
+##            finds a solution with maximum explained variance on the first   ##
+##            dimension                                                       ##
+##                                                                            ##
+################################################################################
+
+
+  require (vegan) || stop ("Requires the vegan package")
+
+## minimum dimensionality of the Isomap-space = 2
+
+  if (ndim < 2)
+    ndim <- 2
+
[…]

### 2020-08-10 code/funs/ordi_helper_funs.R

@@ -1,0 +1,109 @@
+#' @title Function to apply several ordination methods
+#' @details This function lets you use DCA, NMDS and Isomap while also returning
+#'   the variance explained by the ordination axes.
+#' @param data Plot-species matrix with an id column named \code{pnr} which will
+#'   be used to rename the rownames, and then will be removed prior to executing
+#'   the ordination.
+#' @param env_data Dataframe containing environmental variables. The row order
+#'   should correspond to the row order of the \code{data} dataframe.
+#' @param method Ordination method to apply. Supported arguments are:
+#'   \code{"dca"}, \code{"nmds"} and \code{"isomap"}.
+#' @return Output of the ordination will be returned.
+#' @author Jannes Muenchow
+
+ord_fun = function(data, env_data, method) {
+  row.names(data) = data$pnr
+  data$pnr = NULL
+  if (method == "dca") {
+    ord = decorana(data, iweigh = TRUE)
+    cat("single axis contribution: ", 
+        round(ord$evals / sum(ord$evals), 2), "\n")
+    cat("cumulative axes contribution: ",
+        round(cumsum(ord$evals / sum(ord$evals)), 2))  
+  }
+  if (method == "nmds") {
+    ord = metaMDS(data, k = 4, try = 10)
+    cat("R^2 first two axes: ", 
+        round(cor(vegdist(data), dist(scores(ord)[, 1:2]))^2, 3), "\n")
+    cat("R^2 first axis: ", 
+        round(cor(vegdist(data), dist(scores(ord)[, 1]))^2, 3))
+  }
+  
+  if (method == "isomap") {
+    ord = bestisomap(vegdist(data, "bray"), k = 2)
+    cat("Cumulative explained variance (%): ", 
+        ord$`Cumulative explained variance in %`)
+    ord = ord$Scores
+    x11()
+  }
+  
+  fit = envfit(ord, dplyr::select(env_data, -pnr, -id), perm = 999)
+  plot(ord)
+  plot(fit, p.max = 0.05)
+  return(ord)
+}
+
+# from facet_wrap documentation
+# ggplot(sc, aes(DCA1, DCA2)) +
+#   geom_point(data = transform(sc, year = NULL), colour = "grey85") +
+#   geom_point() +
+#   facet_wrap(~year)
+
+#  applying the same idea to lattice
+#' @title Multipanel plot grouped by a factor variable
+#' @details Function creates a multipanel lattice plot conditioned by a factor
+#'   variable.
[…]

### 2020-08-10 code/funs/spcv_helper_funs.R

@@ -1,0 +1,85 @@
+# define own functions
+err_nrmse = function(obs, pred, DIFF = diff(range(obs))) {
+  err = list(
+    bias = mean(obs - pred), 
+    stddev = sd(obs - pred), 
+    mse = mean((obs - pred)^2), 
+    rmse = sqrt(mean((obs - pred)^2)),
+    mad = mad(obs - pred), 
+    median = median(obs - pred), 
+    iqr = IQR(obs - pred))
+  err$nrmse = sqrt(err$mse) / DIFF * 100
+  err$count = length(obs)
+  return(err)
+}
+
+# Adjust partition_means: select a location and keep all years
+# of this location in the test/training dataset
+partition_kmeans_factor = 
+  function(data, coords = c("x", "y"), nfold = 5, repetition = 1, fac, 
+           seed1 = NULL, return_factor = FALSE, balancing_steps = 1, 
+           order_clusters = TRUE, ...) {
+    if (length(repetition) < 2) {
+      repetition = seq(1, repetition, by = 1)
+    }
+    balancing_steps = max(1, balancing_steps)
+    resampling = list()
+    for (cnt in repetition) {
+      if (!is.null(seed1)) {
+        set.seed(seed1 + cnt)
+      }
+      kms = list()
+      # just keep unique coordinates
+      fil = data[!duplicated(data[, fac]), ]
+      for (i in 1:balancing_steps) {
+        kms[[i]] = kmeans(fil[, coords], centers = nfold, ...)
+      }
+      kmgini = function(x) {
+        p = x$size / sum(x$size)
+        return(1 - sum(p^2))
+      }
+      km = kms[[which.max(sapply(kms, kmgini))]]
+      if (order_clusters) {
+        o = rank(km$centers[, 1], ties.method = "first")
+        km$cluster = o[km$cluster]
+      }
+      data[!duplicated(data[, fac]), "cluster"] = km$cluster
+      tile = group_by_at(data, fac) %>%
+        mutate(cluster = cluster[1]) %>%
+        pull(cluster)
+      tile = factor(tile)
+      if (!return_factor) {
+        tile = as.resampling(tile)
+      }
+      resampling[[as.character(cnt)]] = tile
+    }
+    if (!return_factor) {
+      resampling = as.represampling(resampling)
+    }
+    return(resampling)
+  }
+
+#' @title Apply a model to response variables for a spatial prediction
+#' @param data A dataframe containing the predictor and response variables.
[…]

### 2020-08-10 code/funs/varpart_helper_funs.R

@@ -1,0 +1,125 @@
+#' @title Function retrieves the DCA scores of the first two axes and plots them
+#'   against two environmental tables
+#' @details Function applies a DCA, retrieves the scores of the first axes and
+#'   plots the scores against the environmental variables.
+#' @param londo Plot-species matrix with an id column named \code{pnr} which
+#'   will be used to rename the rownames, and then will be removed prior to
+#'   executing the ordination.
+#' @param expl Dataframe containing environmental variables. The dataframe
+#'   should have an id column which should have a correspondence in the
+#'   species-plot-matrix with the same name.
+#' @param choices Number of DCA axes to be retained. Only \code{1:2} is
+#'   currently supported.
+#' @return A \code{list} containing two [lattice::xyplot()]s.
+#' @author Jannes Muenchow
+explore_dca = function(londo, id = "pnr", choices = 1:2, expl) {
+  # just keep observations found in response and explanatory tables
+  expl = inner_join(dplyr::select(londo, id), expl, by = id)
+  # get rid off id but keep the id as rownames
+  rownames(londo) = londo[, id]
+  londo = dplyr::select(londo, -id)  
+  if (any(colSums(londo) == 0)) {
+    stop("There are empty columns")
+  }
+  # check if each plot has at least one observation
+  if (!all(rowSums(londo) > 0)) {
+    stop("There are empty rows!")
+  }
+  
+  dca = decorana(londo, iweigh = TRUE)
+  # extract scores and add id
+  dca = dplyr::select(as.data.frame(scores(dca, display = "sites")), choices)
+  dca = cbind(as.integer(row.names(dca)), dca)
+  names(dca)[1] = id
+  # join environmental data
+  tmp = inner_join(dca, expl, by = id)
+  tmp = select(tmp, -id)
+  dca1 = reshape2::melt(select(tmp, -DCA2), id.var = "DCA1")
+  dca2 = reshape2::melt(select(tmp, -DCA1), id.var = "DCA2")
+  p_1 = xyplot(DCA1 ~ value | variable, data = dca1,
+               col = "salmon", pch = 16,
+               scales = list(relation = "free"),
+               panel = function(x, y, ...) {
+                 panel.xyplot(x, y, ...)
+                 panel.loess(x, y, span = 0.9, lwd = 2.5, col = "gray")
+               })
+  p_2 = xyplot(DCA2 ~ value | variable, data = dca2,
[…]

### 2020-08-10 code/misc/11_consistency.R

@@ -1,0 +1,359 @@
+# Filename: 11_consistency.R
+#
+# TO DO: test for consistency
+#
+# Author(s): Jonas Brock & Jannes Muenchow
+
+
+#Packages and Data-------------------------------------------------------
+
+setwd("C:/Users/Jonas Brock/Desktop/papers/ENSO")
+
+
+library("RODBC")
+library("Hmisc")
+library("lattice")
+library("dplyr")
+library("reshape")
+library("data.table")
+library("RODBC")
+
+path = "data/db/Transekt_0417.accdb"
+con = odbcConnectAccess2007(path)
+sqlTables(con)
+db_odbc = sqlFetch(con, "qurArtDeckungPlot_Kreuztabelle", as.is = TRUE)
+close(con)
+
+# JONAS:
+# load data for 2017
+data17 <- fread("data/data2017.csv")
+all.equal(data17, as.data.table(db_odbc))
+setdiff(names(data17), names(db_odbc))  # Euph_hype
+setdiff(names(db_odbc), names(data17))  # Euph_serp
+
+# use another db saved in our folder structure
+path = "_raw/data_2017/Transekt_0417.accdb"
+con = odbcConnectAccess2007(path)
+sqlTables(con)
+db_odbc = sqlFetch(con, "qurArtDeckungPlot_Kreuztabelle", as.is = TRUE)
+close(con)
+all.equal(data17, as.data.table(db_odbc))
+str(data17[, 1:3])
+str(d_17[, 1:3])
+# ok, this was the one used, clean up accordingly
+
+
+# delete "<>" column
+data17[, "<>" := NULL]
+
+# JONAS:
+#load data for 2016
+data16 <- fread("data/data2016.csv")
+
+path = "data/db/Transekt_0316.accdb"
+con = odbcConnectAccess2007(path)
+sqlTables(con)
+db_odbc = sqlFetch(con, "tblArten", as.is = TRUE)
+close(con)
+
+all.equal(data16, as.data.table(db_odbc))
+setdiff(names(data16), names(db_odbc))
+setdiff(names(db_odbc), names(data16))
+# ok, family was added to tblArten
+
+path = "data/db/Transekt_0316.accdb"
+con = odbcConnectAccess2007(path)
+query = paste("SELECT tblArten.*, tblFamilie.Familie",
[…]

### 2020-08-10 code/misc/12_rda.R

@@ -1,0 +1,352 @@
+# Filename: 12_rda.R (2018-06-01)
+#
+# Author(s):  Gregor Didenko
+#
+# Source(s): Jannes Muenchow & "Chapter 6 - Numerical Ecology with R"
+#             
+#
+#**********************************************************
+# CONTENTS-------------------------------------------------
+#**********************************************************
+#
+# - ATTACH PACKAGES AND DATA
+# - PREPROCESSING
+# - RDA WITH env_chem
+# - RDA WITH TOPO CHARACTERISTICS
+#
+#**********************************************************
+# ATTACH PACKAGES AND READ DATA--------------------------
+#**********************************************************
+
+# load packages
+pacman::p_load(vegan, packfor, dplyr, magrittr)
+
+# load data
+# Environment data
+env_all <- as.data.frame(readRDS("images/06_env_data.rda"))
+
+# Crosstable Londo-scale vegation cover
+CTlondo = readRDS("images/02_CTlondo.rda")
+
+# Crosstable Presence/Absence
+CTpa = readRDS("images/02_CTpa.rda")
+
+
+#**********************************************************
+# PREPROCESSING--------------------------
+#**********************************************************
+
+## CT LONDO:
+# set na values to 0
+CTlondo[is.na(CTlondo)] = 0
+
+# remove plot nr 25
+CTlondo %<>% filter(!grepl("25", pnr))
+
+# Divide datasets into years
+CTlondo_2011 = CTlondo %>% filter(grepl("2011", pnr))
+CTlondo_2012 = CTlondo %>% filter(grepl("2012", pnr))
+CTlondo_2016 = CTlondo %>% filter(grepl("2016", pnr))
+CTlondo_2017 = CTlondo %>% filter(grepl("2017", pnr))
+
+#Helling-transformation (and remove Plotnr. column)
+vege_hel_2011 <- decostand(CTlondo_2011[, -1], "hellinger")
+vege_hel_2012 <- decostand(CTlondo_2012[, -1], "hellinger")
+vege_hel_2016 <- decostand(CTlondo_2016[, -1], "hellinger")
+vege_hel_2017 <- decostand(CTlondo_2017[, -1], "hellinger")
+
+## ENV-DATA
+
+# Correct Error in CN value and transform to numerical
+env_all$cn[env_all$cn == "1.#INF"] = 0
[…]

### 2020-08-10 code/misc/21_greg_rain_expl.R

@@ -1,0 +1,66 @@
+library("tidyverse")
+
+precip = readRDS("images/21_precip.rds")
+d_pred_stack = readRDS("images/25_d_pred_stack.rds")
+
+
+#**********************************************************
+# CREATE DF WITH PRECIP. SEASON SEP-AUG ---------------
+#****************************************<******************
+
+# Heavy Rain can start in December goes to April
+# Add rainy season data variable
+precip$season = precip$year
+# sep-dec from last year, count to season this year
+precip[precip$month %in% char_months[9:12], "season"] = precip[precip$month %in% char_months[9:12], "season"] + 1
+
+precip_season = precip %>% group_by(station, season) %>% summarise(precip = sum(precip, na.rm = TRUE))
+
+summary(precip_season)
+
+## Save precipitation object
+#saveRDS(precip_season, file = "images/21_precip_seasons.rds")
+
+#**********************************************************
+# PLOTTING
+#**********************************************************
+
+precip_season %>% 
+  filter(season > 2007, season < 2018) %>%
+  ggplot(aes(season, precip, color = station)) +
+  geom_point(size = 2) +
+  scale_x_continuous(breaks = 2008:2017)
+
+
+location_weather_stations =
+  readr::read_csv("data/spatial/location_weather_stations.csv")
+
+predic_rgb = RStoolbox::ggRGB(d_pred_stack * coef, 3, 1, 2) + 
+  geom_point(aes(y = location_weather_stations$Lat,
+                 x = location_weather_stations$Long), colour = "white")
+
+
+rain_chulu = precip_season %>% 
+  filter(season %in% c(2011, 2012, 2016, 2017), station == "chulu") %>%
+  ggplot(aes(as.factor(season), precip)) +
+  geom_col(fill = "blue") + 
+  scale_x_discrete(name = "") + 
+  scale_y_log10(breaks = c(10, 50, 100, 500, 1000), limits = c(1, 1700))
+rain_paita = precip_season %>% 
+  filter(season %in% c(2011, 2012, 2016, 2017), station == "paita") %>%
+  ggplot(aes(as.factor(season), precip)) +
+  geom_col(fill = "blue") + 
+  scale_x_discrete(name = "")  + 
+  scale_y_log10(breaks = c(10, 50, 100, 500, 1000), limits = c(1, 1700))
+rain_piura = precip_season %>% 
[…]

### 2020-08-10 code/misc/23_gregor_dca_expl.R

@@ -1,0 +1,148 @@
+#**********************************************************
+# 6 DCA VIZ------------------------------------------------
+#**********************************************************
+
+library("stringr")
+library("vegan")
+library("dplyr")
+library("lattice")
+library("magrittr")
+library("ggplot2")
+
+# source own functions
+source("code/funs/ordi_helper_funs.R")
+
+londo = readRDS("images/11_ct_londo.rds")
+env_data = readRDS("images/11_env_data.rds")
+sc = readRDS("images/15_dca_scores.rds")
+
+dca = ord_fun(londo, env_data, method = "dca")
+# create dataframe with DCA scores and metadata
+scores_df = data.frame(pnr = londo$pnr,
+                       # get year of survey
+                       year = str_sub(londo$pnr, start = -4),
+                       # get plotID from plot number
+                       plotID = str_sub(londo$pnr, end = -6),
+                       # create grouping variables
+                       # group_west = 1:18, group_middle = 19:38, 
+                       # group_east = 39:49
+                       group = factor(rep(c(1, 2, 3), c(18, 20, 11)),
+                                      labels = c("west", "middle", "east")),
+                       sc)
+
+# source("code/funs/ordi_helper_funs.R")
+# multi_panel_plot(DCA1, DCA2, year, id = plotID, data = scores_df)
+
+
+# 6.1 Posthoc env fitting==================================
+#**********************************************************
+
+# simple test
+test = envfit(dca, env_data$x)
+test$vectors$r
+scores(test, display = "vectors")
+
+# check for compatibilty
+str(env_data) # 1, 25, 39 are character type
+summary(env_data)
+# remove character-type colums (pnr, bodenart, cn) from calculation
+sel_columns = unlist(lapply(env_data, class)) != "character"
+# calculate envfit for all variables
+set.seed(123)
+envfit_vectors = sapply(env_data[, sel_columns],
+                        function(x){
+                          # calculate envfit
+                          temp_envfit = envfit(dca, x, perm = 1000)
+                          # get correlation and significance with ordination
[…]

### 2020-08-10 code/misc/25_plotting_deltas.R

@@ -1,0 +1,448 @@
+# Filename: 26_plotting_deltas.R (2018-03-11)
+#
+# Author(s): Gregor Didenko
+#
+#**********************************************************
+# CONTENTS:-------------------------------------------------
+#
+# LOAD AND PREPARE DATA
+# Calculate deltas 12-11, 16-12, 17-16
+# Plotting deltas
+# DCA RGB COLOR PLOTS
+# NDVI RGB COLOR PLOTS
+#
+#**********************************************************
+
+#************************************************
+# LOAD AND PREPARE DATA ----
+#************************************************
+
+pacman::p_load(raster, ggtern, dplyr, reshape2, rasterVis, ggplot2, Ternary)
+
+ndvi_rasters = readRDS("images/14_ndvi_rasters.rds")
+
+pred_rasters = readRDS("images/24_pred_rasters.rds")
+
+# create raster stack of selected years
+ndvi_stack = stack(ndvi_rasters$ndvi_11, ndvi_rasters$ndvi_12, ndvi_rasters$ndvi_16, ndvi_rasters$ndvi_17)
+names(ndvi_stack) = c("ndvi_2011", "ndvi_2012", "ndvi_2016", "ndvi_2017")
+
+
+
+#************************************************
+# Calculate deltas 12-11, 16-12, 17-16 ----
+#************************************************
+
+# NDVI deltas
+ndvi_delta_12_11 = ndvi_rasters$ndvi_12 - ndvi_rasters$ndvi_11
+ndvi_delta_16_12 = ndvi_rasters$ndvi_16 - ndvi_rasters$ndvi_12
+ndvi_delta_17_16 = ndvi_rasters$ndvi_17 - ndvi_rasters$ndvi_16
+d_ndvi_stack = stack(ndvi_delta_12_11, ndvi_delta_16_12, ndvi_delta_17_16)
+names(d_ndvi_stack) = c("d_12_11", "d_16_12", "d_17_16")
+
+## Calculate delta with direction
+# get min value
+min_val = min(values(pred_rasters), na.rm = TRUE)*-1 # -1.730218
+
+# deltas predicted scores with added minimal value to get only change > 0
+scores_delta_12_11 = (pred_rasters$pred_12 + min_val) - (pred_rasters$pred_11 + min_val)
+scores_delta_16_12 = (pred_rasters$pred_16 + min_val) - (pred_rasters$pred_12 + min_val)
[…]

### 2020-08-10 code/misc/99_testingtriangle.R

@@ -1,0 +1,63 @@
+library(ade4)
+
+data2= data.frame(R = 0:10) 
+head(RGB0)  
+
+    
+data2$G <- 10:20
+data2$B <- 20:30
+head(data2)
+
+
+triangle.plot <- triangle.plot(data2, clab = 0, cpoi = 2, 
+                               addmean = F, 
+                               show = F, addaxes = F, scale = T)
+
+
+
+#plot(NA,NA,xlim=c(0,1),ylim=c(0,1),asp=1,bty="n",axes=F,xlab="",ylab="")
+# bigger number makes it smoother
+sm <- 150
+x <- do.call(c, sapply(1:(sm*sqrt(3)/2)/sm,
+                       function(i) (i*sm/sqrt(3)):(sm-i*sm/sqrt(3))/sm))
+y <- do.call(c, sapply(1:(sm*sqrt(3)/2)/sm,
+                       function(i) rep(i, length((i*sm/sqrt(3)):(sm-i*sm/sqrt(3))))))
+d.red = y
+d.green = abs(sqrt(3) * x - y) / sqrt(3 + 1)
+d.blue = abs(- sqrt(3) * x - y + sqrt(3)) / sqrt(3 + 1)
+points(x, y, col=rgb(1-d.red,1 - d.green,1 - d.blue), pch=19)
+points(x, y, col=rgb(d.red,d.green,d.blue), pch=19)
+
+
+
+points(triangle.plot, pch = 19, cex = 1, col = nacreHEX)  
+
+
+
+
+
+
+
+
+
+library(colorscience)
+Maxwell.triangle()
+xl<-yl<-0:1
+Maxwell.triangle(xlim=xl,ylim=yl)
+Maxwell.triangle(conversionFunction=CIE1931XYZ2CIE1976uv, 
+                 xlim=xl,ylim=yl,xlab="u'",ylab="v'")
+
+
+library(trifield)
+
+grid.size = 128
+par(mar = rep(2, 4), oma = rep(0, 4))
+tg = ternary.grid(grid.size)
+f = function(x)
+       sin(2 * pi * x[1]) +
+       sin(3 * pi * x[2]) +
+       sin(4 * pi * x[3])
+z = ternary.apply(tg, f)
+tf = ternary.field(tg, z)
+plot(tf)
+ternary.legend()

### 2020-08-10 code/misc/convert_docx_to_rmd.sh

@@ -1,0 +1,6 @@
+# convert docx into Rmd
+# see: https://gist.github.com/petzi53/8120df519b86578fd7c83e4d74db273f
+cd text/ecography
+pandoc -f docx -t markdown+auto_identifiers -s round_1/enso_comp.docx -o round_2/enso_comp_test.Rmd --atx-headers --wrap=none --extract-media=""
+pandoc -f docx -t markdown+auto_identifiers -s round_1/title_page.docx -o round_2/title_page_test.Rmd --atx-headers --wrap=none --extract-media=""
+pandoc -f docx -t markdown+auto_identifiers -s round_1/supplementary_information.docx -o round_2/supplementary_information_test.Rmd --atx-headers --wrap=none --extract-media=""

### 2020-08-10 code/misc/explore_icen_index.R

@@ -1,0 +1,126 @@
+# Filename: 11_precipitation.R (2018-04-27)
+#
+# TO DO: Analyze precipitation data
+#
+# Author(s):  Jannes Muenchow, Gregor Didenko
+#
+#**********************************************************
+# CONTENTS-------------------------------------------------
+#**********************************************************
+#
+# 1. ATTACH PACKAGES AND DATA
+# 2. PREPROCESSING
+# 3. PREC BARPLOT
+# 4. CREATE DF WITH PRECIP. SEASON SEP-AUG
+# 5. PLOTTING
+#
+#**********************************************************
+# 1 ATTACH PACKAGES AND READ DATA--------------------------
+#**********************************************************
+
+# attach packages
+library("RSQLite")
+library("dplyr")
+library("magrittr")
+library("sf")
+library("raster")
+library("ggplot2")
+library("lattice")
+library("data.table")
+library("RCurl")
+
+# read in the individual sheets with specified range
+con = dbConnect(RSQLite::SQLite(), "../../../../data/tables.gpkg")
+dbListTables(con)
+precip = RSQLite::dbReadTable(con, "precipitation")
+dbDisconnect(con)
+
+#**********************************************************
+# 3 PREC BARPLOT-------------------------------------------
+#**********************************************************
+
+# remove years 1991-1996, they only exist for piura 
+# precip %<>% filter(!year %in% 1991:1996)
+
+# 3.1 Download and prepare ICEN index======================
+#**********************************************************
+
+# ICEN phases
+# ic = readr::read_table("http://www.met.igp.gob.pe/datos/icen.txt", skip = 11)
+# names(ic) = c("year", "month", "icen")
+# ic = mutate(ic, month = as.numeric(as.character(month)))
+# saveRDS(ic, "images/00_icen_index.rds")
+ic = readRDS("../../../../images/00_icen_index.rds")
+ic$enso = cut(ic$icen, breaks = c(-10, -0.99, 0.41, 10),
+              labels = c("LN", "neutral", "EN"))
+# now aggregate to years!!
+# running mean over three months, hence 1 corresponds to djf and 2 to jfm
+tab_2 = filter(ic, month %in% c(1, 2))
+tab_2 = reshape2::dcast(tab_2, year ~ month)
+tab_2 = as.data.table(tab_2)
+names(tab_2) = c("year", "djf", "jfm")
[…]

### 2020-08-10 code/misc/input_data/01_accdb_tables.R

@@ -1,0 +1,187 @@
+# Filename: 00_rodbc.R (2019-09-05)
+
+# TO DO: Read in raw data from accdb, i.e., only possible under Windows
+
+# Author(s): Jannes Muenchow
+
+#**********************************************************
+# CONTENTS-------------------------------------------------
+#**********************************************************
+
+# 1. ATTACH PACKAGES AND DATA
+# 2. PLOT-SPECIES MATRICES (2011, 2012, 2016, 2017)
+# 3. ENVIRONMENTAL DATA
+# 4. IRRIGATION-NUTRIENT EXPERIMENT
+# 5. SAVE TO GPKG
+
+#**********************************************************
+# 1 ATTACH PACKAGES AND DATA-------------------------------
+#**********************************************************
+
+# attach packages
+library("RODBC")
+library("RSQLite")
+library("dplyr")
+library("data.table")
+library("magrittr")
+library("lubridate")
+library("readxl")
+library("reshape2")
+
+# little helper function because we have to do the same thing 4 times
+adjust_tab = function(d, cover = "Deckung_%") {
+  d %>% 
+    # if Arbeitsname does not exist, this only throws a warning message
+    dplyr::select(-one_of("Arbeitsname")) %>%
+    dplyr::rename(art = "Artkürzel", cover = cover) %>%
+    left_join(., lf, by = "art") %>%
+    # arrange by plot id (plot number)
+    arrange(pnr)
+}
+
+#**********************************************************
+# 2 PLOT-SPECIES MATRICES (2011, 2012, 2016, 2017)---------
+#**********************************************************
+
+# NOTE:
+# use 32-bit version of R/Microsoft Office and make sure that you have installed
+# Microsoft Database Engine Distributable (32-bit)
+# browseURL("https://www.microsoft.com/en-us/download/details.aspx?id=13255")
+# remember also that RStudio no longer supports 32-bit R versions, so you have 
+# to run RODBC code in the R console directly
+
+# life form data (lf)
+con = odbcConnectAccess2007("data/raw_data/Transekt_0316.accdb")
+lf = sqlFetch(con, "qurSpecOrigin", as.is = TRUE)
+close(con)
+lf = dplyr::select(lf, art = "Artkürzel", type = "Life form") %>%
[…]

### 2020-08-10 code/misc/input_data/02_vector_tables.R

@@ -1,0 +1,103 @@
+# Filename: vector.R (2018-07-03)
+#
+# TO DO: putting several needed vector data into gpkg
+#
+# Author(s): Jannes Muenchow
+#
+#**********************************************************
+# CONTENTS-------------------------------------------------
+#**********************************************************
+#
+# 1. ATTACH PACKAGES AND DATA
+# 2. GPKG
+# 3. NOMINATIM CITY BOUNDARIES
+#
+#**********************************************************
+# 1 ATTACH PACKAGES AND DATA-------------------------------
+#**********************************************************
+
+# attach packages
+library("sf")
+library("raster")
+library("RSQLite")
+# define directory
+old_dir = file.path("~/work/acad/science/projects/ecology/latin_america/peru/", 
+                    "transect_piura/old/data/R/R data/shapes")
+# read in data
+towns = st_read(file.path(old_dir, "towns.shp"))
+plot(towns$geometry)  # these were the town outlines used in the Erdkunde paper
+towns_2 = st_read(file.path(old_dir, "towns_hannes.shp"))
+plot(towns_2$geometry, add = TRUE)  # and these were the ones used in the JVS paper
+streets = st_read(file.path(old_dir, "streets.shp"))
+arable = st_read(file.path(old_dir, "arable_land.shp"))
+mountains = st_read(file.path(old_dir, "mountains.shp"))
+sa = st_read(file.path(old_dir, "study_area.shp"))
+coast = st_read(file.path(old_dir, "coastline.shp"))
+# water lines were downloaded from DIVA GIS
+rivers = st_read(file.path(old_dir, "PER_wat/PER_water_lines_dcw.shp"))
+rivers = st_transform(rivers, st_crs(sa))
+# neighboring countries
+tmp = ccodes()
+dplyr::filter(tmp,
+              NAME %in% c("Peru", "Ecuador", "Brazil", "Colombia", "Bolivia", 
+                          "Chile"))
+nbs = lapply(c("BOL", "BRA", "CHL", "COL", "ECU"), function(i) {
+  getData("GADM", country = i, level = 0) %>%
+    st_as_sf
+  })
+nbs = do.call(rbind, nbs)
+plot(nbs$geometry, col = NA)
+per = getData("GADM", country = "PER", level = 1) %>% 
+  st_as_sf
+
+#**********************************************************
+# 2 GPKG---------------------------------------------------
[…]

### 2020-08-10 code/misc/inset_map.R

@@ -1,0 +1,225 @@
+# Filename: study_area.R (2019-06-03)
+#
+# TO DO: Overview figure (South America and prediction map)
+#
+# Author(s): Jannes Muenchow
+#
+#**********************************************************
+# CONTENTS-------------------------------------------------
+#**********************************************************
+#
+# 1. ATTACH PACKAGES AND DATA
+# 2. SOUTH AMERICA STUDY AREA
+# 3. LOCAL PREDICTION MAP
+# 4. GRID FIGURE
+#
+#**********************************************************
+# 1 ATTACH PACKAGES AND DATA-------------------------------
+#**********************************************************
+
+# attach packages
+library("sf")
+library("sp")
+library("raster")
+library("dplyr")
+library("latticeExtra")
+library("grid")
+library("gridExtra")
+library("RSQLite")
+
+# # attach data
+cous = c("PER", "ECU", "BOL", "COL", "BRA", "CHL")
+cous = lapply(cous, function(x) {
+  shp = raster::getData(cou = x, level = 0) %>%
+    st_as_sf
+})
+cous = do.call(rbind, cous)
+b_box = st_bbox(cous[1:2, ])
+b_box["xmin"] = st_bbox(cous[1, ])$xmin
+# add something, otherwise Bolivia will become a gc which cannot be converted
+# into a Spatial object
+b_box["xmax"] = st_bbox(cous[1, ])$xmax + 0.5
+sa = st_crop(cous, b_box)
+# check
+sa = as(sa, "Spatial")  # ok, it works
+# saveRDS(sa, file = "images/sa.rds")
+
+# con = dbConnect(SQLite(), "data/tables.gpkg")
+# dbListTables(con)
+# peru = st_read(con, "peru")
+# nbs = st_read(con, "neighbors")
+# 
+# sa = rbind(nbs, select(peru, GID_0, NAME_0))
+# sa = as(sa, "Spatial")  
+
+#*******************************************************
+# 2 SA inset map----------------------------------------
+#*******************************************************
+
+poly = st_bbox(c(xmin = -81, xmax = -79, ymax = -4.6, ymin = -5.4), 
+               crs = st_crs(4326)) %>% 
+  st_as_sfc %>% 
+  as(., "Spatial")
+
+# Base plot version
+
+# however, you cannot mix base plots and lattice/ggplots with grid
+
+# pin = par("pin")
+# # calculate the x- and y-extent
[…]

### 2020-08-10 code/misc/jonas_desc_analysis.R

@@ -1,0 +1,364 @@
+#**********************************************************
+# JONAS STUFF - CAN BE DELETED AFTER TALKING TO ALEX-------
+#**********************************************************
+
+#************************************************
+# CALCULATE MEAN NUMBER OF PLANTS PER PLOT ----
+#************************************************
+# mean number of plants per plot
+dat11 = base::subset(d11, select = "pnr")
+dat12 = base::subset(d12, select = "pnr")
+dat16 = base::subset(d16, select = "pnr")
+dat17 = base::subset(d17, select = "pnr")
+
+# calculate sum of respective plots
+dat11 = dat11 %>%
+  summarise(no_rows = length(pnr))
+
+dat12 = dat12 %>%
+  summarise(no_rows = length(pnr))
+
+dat16 = dat16 %>%
+  summarise(no_rows = length(pnr))
+
+dat17 = dat17 %>%
+  summarise(no_rows = length(pnr))
+
+# preparation for calculating the mean
+plant_mean = cbind(dat11, dat12, dat16, dat17)
+plant_mean = plant_mean[-c(3, 5, 7)]
+colnames(plant_mean)[2] = "2011"
+colnames(plant_mean)[3] = "2012"
+colnames(plant_mean)[4] = "2016"
+colnames(plant_mean)[5] = "2017"
+
+# calculate the mean per plot
+plant_mean$mean = round((plant_mean$`2011` + plant_mean$`2012` +
+  plant_mean$`2016` + plant_mean$`2017`)/4, 0)
+
+
+# define plot style
+plot_syle = theme_bw() + theme(panel.border = element_rect(colour = "black", size=1.4),
+                                axis.line = element_line(colour = "black", size = 1.4),
+                                axis.text.x = element_text(colour = "black"),
+                                axis.text.y = element_text(colour = "black"),
+                                axis.title=element_text(size=14,face="bold"))
+
+# setting the style of the plot (without grid!) -> alternative style
+# plot_syle = theme_bw() + theme(panel.border = element_rect(colour = "black", size=1.4),
[…]

### 2020-08-10 code/misc/latexdiff.sh

@@ -1,0 +1,2 @@
+latexdiff text/ecography/round_2/enso_comp.tex text/ecography/round_3/enso_comp.tex > text/ecography/round_3/diff_ms.tex
+latexdiff text/ecography/round_2/title_page.tex text/ecography/round_3/title_page.tex > text/ecography/round_3/diff_tp.tex

### 2020-08-10 code/misc/single_models.R

@@ -1,0 +1,128 @@
+# Filename: single_models (2019-09-03)
+
+# TO DO: Run each year as a single model
+
+# Author(s): Jannes Muenchow
+
+#**********************************************************
+# CONTENTS-------------------------------------------------
+#**********************************************************
+
+# 1. ATTACH PACKAGES AND DATA
+# 2. INDVIDUAL MODELS
+
+#**********************************************************
+# 1 ATTACH PACKAGES AND DATA-------------------------------
+#**********************************************************
+
+# attach packages
+# attach packages
+library("dplyr")
+library("gdata")
+library("lattice")
+library("latticeExtra")
+library("mgcv")
+library("raster")
+library("rasterVis")
+library("gridExtra")
+library("sperrorest")
+library("sf")
+
+# source your own funs
+source("code/funs/spcv_helper_funs.R")
+
+# attach raster stack
+ndvi = readRDS("images/14_ndvi_rasters.rds")
+# attach response (DCA scores)
+data = readRDS("images/23_dca_scores.rds")
+data$id = as.numeric(data$id)
+# attach env_data
+env_data = readRDS("images/12_env_data.rds")
+env_data$year = gsub(".*_", "", env_data$pnr)
+# attaching spatial data
+rivers = st_read("data/spatial/vector.gpkg", "rivers")
+towns = st_read("data/spatial/vector.gpkg", "towns_jvs")
+
+#**********************************************************
+# 1 INDIVIDUAL MODELS--------------------------------------
+#**********************************************************
+
+# just for exploratory purposes, not used in the paper!
+# add xy-coordinates to dca scores
+data = inner_join(data, 
+                  dplyr::select(env_data, id, year, x, y, lon, lat, slope, 
+                                vcurv),
+                  by = c("id", "year"))
+data$ndvi = NA
+for (i in c(11, 12, 16, 17)) {
+  ind = grepl(i, data[, "year"]) 
+  data[ind, "ndvi"] = 
+    raster::extract(ndvi[[paste0("ndvi_", i)]], 
+                    data[ind, c("lon", "lat")])
+}
+sum(is.na(data$ndvi))  # 0, very good
+
+
+# have a look at the relationship between DCA1 and NDVI again
+
+# make a plot
[…]

### 2020-08-11 R/ordi_helper_funs.R
putting helper_funs into R folder and fixing pull bug

@@ -10,34 +10,37 @@
 #'   \code{"dca"}, \code{"nmds"} and \code{"isomap"}.
 #' @return Output of the ordination will be returned.
 #' @author Jannes Muenchow
+#' @importFrom vegan decorana metaMDS vegdist scores envfit
+#' @importFrom dplyr select
 
 ord_fun = function(data, env_data, method) {
   row.names(data) = data$pnr
   data$pnr = NULL
   if (method == "dca") {
-    ord = decorana(data, iweigh = TRUE)
+    ord = vegan::decorana(data, iweigh = TRUE)
     cat("single axis contribution: ", 
         round(ord$evals / sum(ord$evals), 2), "\n")
     cat("cumulative axes contribution: ",
         round(cumsum(ord$evals / sum(ord$evals)), 2))  
   }
   if (method == "nmds") {
-    ord = metaMDS(data, k = 4, try = 10)
+    ord = vegan::metaMDS(data, k = 4, try = 10)
     cat("R^2 first two axes: ", 
-        round(cor(vegdist(data), dist(scores(ord)[, 1:2]))^2, 3), "\n")
+        round(cor(vegan::vegdist(data), 
+                  dist(vegan::scores(ord)[, 1:2]))^2, 3), "\n")
     cat("R^2 first axis: ", 
-        round(cor(vegdist(data), dist(scores(ord)[, 1]))^2, 3))
+        round(cor(vegan::vegdist(data), dist(vegan::scores(ord)[, 1]))^2, 3))
   }
   
   if (method == "isomap") {
-    ord = bestisomap(vegdist(data, "bray"), k = 2)
+    ord = bestisomap(vegan::vegdist(data, "bray"), k = 2)
     cat("Cumulative explained variance (%): ", 
         ord$`Cumulative explained variance in %`)
     ord = ord$Scores
     x11()
   }
   
-  fit = envfit(ord, dplyr::select(env_data, -pnr, -id), perm = 999)
+  fit = vegan::envfit(ord, dplyr::select(env_data, -pnr, -id), perm = 999)
   plot(ord)
   plot(fit, p.max = 0.05)
   return(ord)
@@ -62,40 +65,49 @@ ord_fun = function(data, env_data, method) {
 #' @param data The dataframe containing the columns.
 #' @return xyplot
 #' @author Jannes Muenchow
+#' @importFrom rlang as_string enexpr
+#' @importFrom dplyr filter pull
[…]

### 2020-08-11 code/analysis/00_modis_download.R
rename properly

@@ -1,4 +1,4 @@
-# Filename: 03_modis_download.R (2018-02-22)
+# Filename: 00_modis_download.R (2018-02-22)
 #
 #
 # Author(s): Jannes Muenchow, Gregor Didenko

### 2020-08-11 text/ecography/round_3/enso_comp.Rmd
replace source() by load_all()

@@
 library("dplyr")
-# read in functions
-source(here("code/funs/varpart_helper_funs.R"))
+# attach own functions
+devtools::load_all()
 

### 2020-08-11 code/analysis/13_descriptive_stats.R

@@ -26,8 +26,8 @@ library("latticeExtra")
 library("here")
 library("vegan")
 library("magrittr")
-# source your own functions
-source(here("code/funs/ordi_helper_funs.R"))
+# attach your own functions
+devtools::load_all()
 
 # attach cover datasets (d = Deckung, i.e., cover)
 d_11 = readRDS(here("images/11_d_11.rds")) %>%

### 2020-08-11 code/analysis/14_ordination.R

@@ -28,9 +28,9 @@ library("magrittr")
 library("ggplot2")
 library("here")
 
-# source own functions
-source("code/funs/ordi_helper_funs.R")
-source("code/funs/bestisomap.R")
+# attach own functions
+devtools::load_all()
+
 # attach data
 pa = readRDS("images/11_ct_pa.rds")
 # Remember, there are 196 obs instead of 2000 because:

### 2020-08-11 code/analysis/15_modeling.R

@@ -33,8 +33,8 @@ library("sf")
 library("RSQLite")
 library("here")
 
-# source your own funs
-source("code/funs/spcv_helper_funs.R")
+# attach your own funs
+devtools::load_all()
 
 # attach raster stack
 ndvi = readRDS("images/00_ndvi_rasters.rds")

### 2020-08-11 code/analysis/16_varpart.R

@@ -20,10 +20,7 @@
 library("vegan")
 library("dplyr")
 library("lattice")
-
-# read in functions
-source("code/funs/HighstatLib.R")  # corvif
-source("code/funs/varpart_helper_funs.R")
+devtools::load_all()
 
 # attach data
 ct_londo_11 = readRDS("images/11_ct_londo_11.rds")

### 2020-08-11 R/HighstatLib.R
add acknowledgment

@@ -1,3 +1,4 @@
+# Written by Dr. Alain Zuur (www.highstat.com)
 
 ##################################################################
 #Below are some functions that we took from the pairs help file and

### 2020-08-11 R/spcv_helper_funs.R
add importFrom functions

@@ -1,4 +1,4 @@
-# define own functions
+# define own error functions
 err_nrmse = function(obs, pred, DIFF = diff(range(obs))) {
   err = list(
     bias = mean(obs - pred), 
@@ -15,6 +15,8 @@ err_nrmse = function(obs, pred, DIFF = diff(range(obs))) {
 
 # Adjust partition_means: select a location and keep all years
 # of this location in the test/training dataset
+#' @importFrom dplyr group_by_at select pull mutate
+#' @importFrom sperrorest as.represampling 
 partition_kmeans_factor = 
   function(data, coords = c("x", "y"), nfold = 5, repetition = 1, fac, 
            seed1 = NULL, return_factor = FALSE, balancing_steps = 1, 
@@ -44,17 +46,17 @@ partition_kmeans_factor =
         km$cluster = o[km$cluster]
       }
       data[!duplicated(data[, fac]), "cluster"] = km$cluster
-      tile = group_by_at(data, fac) %>%
-        mutate(cluster = cluster[1]) %>%
-        pull(cluster)
+      tile = dplyr::group_by_at(data, fac) %>%
+        dplyr::mutate(cluster = cluster[1]) %>%
+        dplyr::pull(cluster)
       tile = factor(tile)
       if (!return_factor) {
-        tile = as.resampling(tile)
+        tile = sperrorest::as.resampling(tile)
       }
       resampling[[as.character(cnt)]] = tile
     }
     if (!return_factor) {
-      resampling = as.represampling(resampling)
+      resampling = sperrorest::as.represampling(resampling)
     }
     return(resampling)
   }
@@ -65,6 +67,7 @@ partition_kmeans_factor =
 #' @param year The year given as a factor for the prediction should be done.
 #' @param pred_raster The raster containing the predictor raster of the year
 #'   for which the spatial prediction should be done.
+#' @importFrom raster values stack extend predict
 pred_fun = function(data, model, year, ndvi_raster) {
   data = data[data$year == year, ]
   # just keep positive NDVI values
@@ -74,12 +77,12 @@ pred_fun = function(data, model, year, ndvi_raster) {
   # 2016, 2017), if you just type factor(2011) or factor(2012) the predictions
   # will be all made for the first level, i.e., the first year (2011) which is
   # surely not what we want
-  values(y) = data$year[1]
-  s = stack(r, y)
+  raster::values(y) = data$year[1]
+  s = raster::stack(r, y)
[…]

### 2020-08-11 R/varpart_helper_funs.R

@@ -12,9 +12,13 @@
 #'   currently supported.
 #' @return A \code{list} containing two [lattice::xyplot()]s.
 #' @author Jannes Muenchow
+#' @importFrom dplyr select inner_join
+#' @importFrom vegan scores decorana
+#' @importFrom reshape2 melt
+#' @importFrom lattice xyplot panel.xyplot panel.loess
 explore_dca = function(londo, id = "pnr", choices = 1:2, expl) {
   # just keep observations found in response and explanatory tables
-  expl = inner_join(dplyr::select(londo, id), expl, by = id)
+  expl = dplyr::inner_join(dplyr::select(londo, id), expl, by = id)
   # get rid off id but keep the id as rownames
   rownames(londo) = londo[, id]
   londo = dplyr::select(londo, -id)  
@@ -26,31 +30,36 @@ explore_dca = function(londo, id = "pnr", choices = 1:2, expl) {
     stop("There are empty rows!")
   }
   
-  dca = decorana(londo, iweigh = TRUE)
+  dca = vegan::decorana(londo, iweigh = TRUE)
   # extract scores and add id
-  dca = dplyr::select(as.data.frame(scores(dca, display = "sites")), choices)
+  dca = dplyr::select(as.data.frame(vegan::scores(dca, display = "sites")),
+                      choices)
   dca = cbind(as.integer(row.names(dca)), dca)
   names(dca)[1] = id
   # join environmental data
-  tmp = inner_join(dca, expl, by = id)
-  tmp = select(tmp, -id)
-  dca1 = reshape2::melt(select(tmp, -DCA2), id.var = "DCA1")
-  dca2 = reshape2::melt(select(tmp, -DCA1), id.var = "DCA2")
-  p_1 = xyplot(DCA1 ~ value | variable, data = dca1,
-               col = "salmon", pch = 16,
-               scales = list(relation = "free"),
-               panel = function(x, y, ...) {
-                 panel.xyplot(x, y, ...)
-                 panel.loess(x, y, span = 0.9, lwd = 2.5, col = "gray")
-               })
-  p_2 = xyplot(DCA2 ~ value | variable, data = dca2,
-               col = "salmon", pch = 16,
-               scales = list(relation = "free"),
-               panel = function(x, y, ...) {
-                 panel.xyplot(x, y, ...)
-                 panel.loess(x, y, span = 0.9, lwd = 2.5, col = "gray")
-               })
-  return(list(p_1, p_2))
[…]

### 2020-08-11 docs/enso_comp.Rmd
put ms into docs and just keep necessary files

--- 2020-08-11 text/ecography/round_3/enso_comp.Rmd
@@
 title: "Monitoring and predictive mapping of floristic biodiversity along a climatic gradient in ENSO's terrestrial core region, NW Peru"
-bibliography: ../transect_refs.bib
-csl: ../ecography.csl
+bibliography: transect_refs.bib
+csl: ecography.csl
 # biblio-style: "natbib"

### 2020-08-11 docs/title_page.Rmd
put ms into docs and just keep necessary files

--- 2020-08-10 text/ecography/round_3/title_page.Rmd
@@
 title: "Monitoring and predictive mapping of floristic biodiversity along a climatic gradient in ENSO's terrestrial core region, NW Peru"
-#bibliography: transect_refs.bib
-#csl: ecography.csl
+bibliography: transect_refs.bib
+csl: ecography.csl
 # biblio-style: "natbib"

### 2020-08-11 docs/supplementary_information.Rmd
put ms into docs and just keep necessary files
keep also tex in appendix

--- 2020-08-10 text/ecography/round_3/supplementary_information.Rmd
@@
   bookdown::pdf_document2:
-    keep_tex: no
+    keep_tex: yes
     latex_engine: xelatex
@@
 title: Supplementary Information
-bibliography: ../transect_refs.bib
-csl: ../ecography.csl
+bibliography: transect_refs.bib
+csl: ecography.csl
 # biblio-style: "natbib"

### 2020-08-12 code/misc/11_consistency.R
delete unnecessary stuff from github repo

@@ -1,359 +1,0 @@
-# Filename: 11_consistency.R
-#
-# TO DO: test for consistency
-#
-# Author(s): Jonas Brock & Jannes Muenchow
-
-
-#Packages and Data-------------------------------------------------------
-
-setwd("C:/Users/Jonas Brock/Desktop/papers/ENSO")
-
-
-library("RODBC")
-library("Hmisc")
-library("lattice")
-library("dplyr")
-library("reshape")
-library("data.table")
-library("RODBC")
-
-path = "data/db/Transekt_0417.accdb"
-con = odbcConnectAccess2007(path)
-sqlTables(con)
-db_odbc = sqlFetch(con, "qurArtDeckungPlot_Kreuztabelle", as.is = TRUE)
-close(con)
-
-# JONAS:
-# load data for 2017
-data17 <- fread("data/data2017.csv")
-all.equal(data17, as.data.table(db_odbc))
-setdiff(names(data17), names(db_odbc))  # Euph_hype
-setdiff(names(db_odbc), names(data17))  # Euph_serp
-
-# use another db saved in our folder structure
-path = "_raw/data_2017/Transekt_0417.accdb"
-con = odbcConnectAccess2007(path)
-sqlTables(con)
-db_odbc = sqlFetch(con, "qurArtDeckungPlot_Kreuztabelle", as.is = TRUE)
-close(con)
-all.equal(data17, as.data.table(db_odbc))
-str(data17[, 1:3])
-str(d_17[, 1:3])
-# ok, this was the one used, clean up accordingly
-
-
-# delete "<>" column
-data17[, "<>" := NULL]
-
-# JONAS:
-#load data for 2016
-data16 <- fread("data/data2016.csv")
-
-path = "data/db/Transekt_0316.accdb"
-con = odbcConnectAccess2007(path)
-sqlTables(con)
-db_odbc = sqlFetch(con, "tblArten", as.is = TRUE)
-close(con)
-
-all.equal(data16, as.data.table(db_odbc))
-setdiff(names(data16), names(db_odbc))
-setdiff(names(db_odbc), names(data16))
-# ok, family was added to tblArten
-
-path = "data/db/Transekt_0316.accdb"
-con = odbcConnectAccess2007(path)
-query = paste("SELECT tblArten.*, tblFamilie.Familie",
[…]

### 2020-08-12 code/misc/12_rda.R

@@ -1,352 +1,0 @@
-# Filename: 12_rda.R (2018-06-01)
-#
-# Author(s):  Gregor Didenko
-#
-# Source(s): Jannes Muenchow & "Chapter 6 - Numerical Ecology with R"
-#             
-#
-#**********************************************************
-# CONTENTS-------------------------------------------------
-#**********************************************************
-#
-# - ATTACH PACKAGES AND DATA
-# - PREPROCESSING
-# - RDA WITH env_chem
-# - RDA WITH TOPO CHARACTERISTICS
-#
-#**********************************************************
-# ATTACH PACKAGES AND READ DATA--------------------------
-#**********************************************************
-
-# load packages
-pacman::p_load(vegan, packfor, dplyr, magrittr)
-
-# load data
-# Environment data
-env_all <- as.data.frame(readRDS("images/06_env_data.rda"))
-
-# Crosstable Londo-scale vegation cover
-CTlondo = readRDS("images/02_CTlondo.rda")
-
-# Crosstable Presence/Absence
-CTpa = readRDS("images/02_CTpa.rda")
-
-
-#**********************************************************
-# PREPROCESSING--------------------------
-#**********************************************************
-
-## CT LONDO:
-# set na values to 0
-CTlondo[is.na(CTlondo)] = 0
-
-# remove plot nr 25
-CTlondo %<>% filter(!grepl("25", pnr))
-
-# Divide datasets into years
-CTlondo_2011 = CTlondo %>% filter(grepl("2011", pnr))
-CTlondo_2012 = CTlondo %>% filter(grepl("2012", pnr))
-CTlondo_2016 = CTlondo %>% filter(grepl("2016", pnr))
-CTlondo_2017 = CTlondo %>% filter(grepl("2017", pnr))
-
-#Helling-transformation (and remove Plotnr. column)
-vege_hel_2011 <- decostand(CTlondo_2011[, -1], "hellinger")
-vege_hel_2012 <- decostand(CTlondo_2012[, -1], "hellinger")
-vege_hel_2016 <- decostand(CTlondo_2016[, -1], "hellinger")
-vege_hel_2017 <- decostand(CTlondo_2017[, -1], "hellinger")
-
-## ENV-DATA
-
-# Correct Error in CN value and transform to numerical
-env_all$cn[env_all$cn == "1.#INF"] = 0
[…]

### 2020-08-12 code/misc/21_greg_rain_expl.R

@@ -1,66 +1,0 @@
-library("tidyverse")
-
-precip = readRDS("images/21_precip.rds")
-d_pred_stack = readRDS("images/25_d_pred_stack.rds")
-
-
-#**********************************************************
-# CREATE DF WITH PRECIP. SEASON SEP-AUG ---------------
-#****************************************<******************
-
-# Heavy Rain can start in December goes to April
-# Add rainy season data variable
-precip$season = precip$year
-# sep-dec from last year, count to season this year
-precip[precip$month %in% char_months[9:12], "season"] = precip[precip$month %in% char_months[9:12], "season"] + 1
-
-precip_season = precip %>% group_by(station, season) %>% summarise(precip = sum(precip, na.rm = TRUE))
-
-summary(precip_season)
-
-## Save precipitation object
-#saveRDS(precip_season, file = "images/21_precip_seasons.rds")
-
-#**********************************************************
-# PLOTTING
-#**********************************************************
-
-precip_season %>% 
-  filter(season > 2007, season < 2018) %>%
-  ggplot(aes(season, precip, color = station)) +
-  geom_point(size = 2) +
-  scale_x_continuous(breaks = 2008:2017)
-
-
-location_weather_stations =
-  readr::read_csv("data/spatial/location_weather_stations.csv")
-
-predic_rgb = RStoolbox::ggRGB(d_pred_stack * coef, 3, 1, 2) + 
-  geom_point(aes(y = location_weather_stations$Lat,
-                 x = location_weather_stations$Long), colour = "white")
-
-
-rain_chulu = precip_season %>% 
-  filter(season %in% c(2011, 2012, 2016, 2017), station == "chulu") %>%
-  ggplot(aes(as.factor(season), precip)) +
-  geom_col(fill = "blue") + 
-  scale_x_discrete(name = "") + 
-  scale_y_log10(breaks = c(10, 50, 100, 500, 1000), limits = c(1, 1700))
-rain_paita = precip_season %>% 
-  filter(season %in% c(2011, 2012, 2016, 2017), station == "paita") %>%
-  ggplot(aes(as.factor(season), precip)) +
-  geom_col(fill = "blue") + 
-  scale_x_discrete(name = "")  + 
-  scale_y_log10(breaks = c(10, 50, 100, 500, 1000), limits = c(1, 1700))
-rain_piura = precip_season %>% 
[…]

### 2020-08-12 code/misc/23_gregor_dca_expl.R

@@ -1,148 +1,0 @@
-#**********************************************************
-# 6 DCA VIZ------------------------------------------------
-#**********************************************************
-
-library("stringr")
-library("vegan")
-library("dplyr")
-library("lattice")
-library("magrittr")
-library("ggplot2")
-
-# source own functions
-source("code/funs/ordi_helper_funs.R")
-
-londo = readRDS("images/11_ct_londo.rds")
-env_data = readRDS("images/11_env_data.rds")
-sc = readRDS("images/15_dca_scores.rds")
-
-dca = ord_fun(londo, env_data, method = "dca")
-# create dataframe with DCA scores and metadata
-scores_df = data.frame(pnr = londo$pnr,
-                       # get year of survey
-                       year = str_sub(londo$pnr, start = -4),
-                       # get plotID from plot number
-                       plotID = str_sub(londo$pnr, end = -6),
-                       # create grouping variables
-                       # group_west = 1:18, group_middle = 19:38, 
-                       # group_east = 39:49
-                       group = factor(rep(c(1, 2, 3), c(18, 20, 11)),
-                                      labels = c("west", "middle", "east")),
-                       sc)
-
-# source("code/funs/ordi_helper_funs.R")
-# multi_panel_plot(DCA1, DCA2, year, id = plotID, data = scores_df)
-
-
-# 6.1 Posthoc env fitting==================================
-#**********************************************************
-
-# simple test
-test = envfit(dca, env_data$x)
-test$vectors$r
-scores(test, display = "vectors")
-
-# check for compatibilty
-str(env_data) # 1, 25, 39 are character type
-summary(env_data)
-# remove character-type colums (pnr, bodenart, cn) from calculation
-sel_columns = unlist(lapply(env_data, class)) != "character"
-# calculate envfit for all variables
-set.seed(123)
-envfit_vectors = sapply(env_data[, sel_columns],
-                        function(x){
-                          # calculate envfit
-                          temp_envfit = envfit(dca, x, perm = 1000)
-                          # get correlation and significance with ordination
[…]

### 2020-08-12 code/misc/25_plotting_deltas.R

@@ -1,448 +1,0 @@
-# Filename: 26_plotting_deltas.R (2018-03-11)
-#
-# Author(s): Gregor Didenko
-#
-#**********************************************************
-# CONTENTS:-------------------------------------------------
-#
-# LOAD AND PREPARE DATA
-# Calculate deltas 12-11, 16-12, 17-16
-# Plotting deltas
-# DCA RGB COLOR PLOTS
-# NDVI RGB COLOR PLOTS
-#
-#**********************************************************
-
-#************************************************
-# LOAD AND PREPARE DATA ----
-#************************************************
-
-pacman::p_load(raster, ggtern, dplyr, reshape2, rasterVis, ggplot2, Ternary)
-
-ndvi_rasters = readRDS("images/14_ndvi_rasters.rds")
-
-pred_rasters = readRDS("images/24_pred_rasters.rds")
-
-# create raster stack of selected years
-ndvi_stack = stack(ndvi_rasters$ndvi_11, ndvi_rasters$ndvi_12, ndvi_rasters$ndvi_16, ndvi_rasters$ndvi_17)
-names(ndvi_stack) = c("ndvi_2011", "ndvi_2012", "ndvi_2016", "ndvi_2017")
-
-
-
-#************************************************
-# Calculate deltas 12-11, 16-12, 17-16 ----
-#************************************************
-
-# NDVI deltas
-ndvi_delta_12_11 = ndvi_rasters$ndvi_12 - ndvi_rasters$ndvi_11
-ndvi_delta_16_12 = ndvi_rasters$ndvi_16 - ndvi_rasters$ndvi_12
-ndvi_delta_17_16 = ndvi_rasters$ndvi_17 - ndvi_rasters$ndvi_16
-d_ndvi_stack = stack(ndvi_delta_12_11, ndvi_delta_16_12, ndvi_delta_17_16)
-names(d_ndvi_stack) = c("d_12_11", "d_16_12", "d_17_16")
-
-## Calculate delta with direction
-# get min value
-min_val = min(values(pred_rasters), na.rm = TRUE)*-1 # -1.730218
-
-# deltas predicted scores with added minimal value to get only change > 0
-scores_delta_12_11 = (pred_rasters$pred_12 + min_val) - (pred_rasters$pred_11 + min_val)
-scores_delta_16_12 = (pred_rasters$pred_16 + min_val) - (pred_rasters$pred_12 + min_val)
[…]

### 2020-08-12 code/misc/99_testingtriangle.R

@@ -1,63 +1,0 @@
-library(ade4)
-
-data2= data.frame(R = 0:10) 
-head(RGB0)  
-
-    
-data2$G <- 10:20
-data2$B <- 20:30
-head(data2)
-
-
-triangle.plot <- triangle.plot(data2, clab = 0, cpoi = 2, 
-                               addmean = F, 
-                               show = F, addaxes = F, scale = T)
-
-
-
-#plot(NA,NA,xlim=c(0,1),ylim=c(0,1),asp=1,bty="n",axes=F,xlab="",ylab="")
-# bigger number makes it smoother
-sm <- 150
-x <- do.call(c, sapply(1:(sm*sqrt(3)/2)/sm,
-                       function(i) (i*sm/sqrt(3)):(sm-i*sm/sqrt(3))/sm))
-y <- do.call(c, sapply(1:(sm*sqrt(3)/2)/sm,
-                       function(i) rep(i, length((i*sm/sqrt(3)):(sm-i*sm/sqrt(3))))))
-d.red = y
-d.green = abs(sqrt(3) * x - y) / sqrt(3 + 1)
-d.blue = abs(- sqrt(3) * x - y + sqrt(3)) / sqrt(3 + 1)
-points(x, y, col=rgb(1-d.red,1 - d.green,1 - d.blue), pch=19)
-points(x, y, col=rgb(d.red,d.green,d.blue), pch=19)
-
-
-
-points(triangle.plot, pch = 19, cex = 1, col = nacreHEX)  
-
-
-
-
-
-
-
-
-
-library(colorscience)
-Maxwell.triangle()
-xl<-yl<-0:1
-Maxwell.triangle(xlim=xl,ylim=yl)
-Maxwell.triangle(conversionFunction=CIE1931XYZ2CIE1976uv, 
-                 xlim=xl,ylim=yl,xlab="u'",ylab="v'")
-
-
-library(trifield)
-
-grid.size = 128
-par(mar = rep(2, 4), oma = rep(0, 4))
-tg = ternary.grid(grid.size)
-f = function(x)
-       sin(2 * pi * x[1]) +
-       sin(3 * pi * x[2]) +
-       sin(4 * pi * x[3])
-z = ternary.apply(tg, f)
-tf = ternary.field(tg, z)
-plot(tf)
-ternary.legend()

### 2020-08-12 code/misc/convert_docx_to_rmd.sh

@@ -1,6 +1,0 @@
-# convert docx into Rmd
-# see: https://gist.github.com/petzi53/8120df519b86578fd7c83e4d74db273f
-cd text/ecography
-pandoc -f docx -t markdown+auto_identifiers -s round_1/enso_comp.docx -o round_2/enso_comp_test.Rmd --atx-headers --wrap=none --extract-media=""
-pandoc -f docx -t markdown+auto_identifiers -s round_1/title_page.docx -o round_2/title_page_test.Rmd --atx-headers --wrap=none --extract-media=""
-pandoc -f docx -t markdown+auto_identifiers -s round_1/supplementary_information.docx -o round_2/supplementary_information_test.Rmd --atx-headers --wrap=none --extract-media=""

### 2020-08-12 code/misc/explore_icen_index.R

@@ -1,126 +1,0 @@
-# Filename: 11_precipitation.R (2018-04-27)
-#
-# TO DO: Analyze precipitation data
-#
-# Author(s):  Jannes Muenchow, Gregor Didenko
-#
-#**********************************************************
-# CONTENTS-------------------------------------------------
-#**********************************************************
-#
-# 1. ATTACH PACKAGES AND DATA
-# 2. PREPROCESSING
-# 3. PREC BARPLOT
-# 4. CREATE DF WITH PRECIP. SEASON SEP-AUG
-# 5. PLOTTING
-#
-#**********************************************************
-# 1 ATTACH PACKAGES AND READ DATA--------------------------
-#**********************************************************
-
-# attach packages
-library("RSQLite")
-library("dplyr")
-library("magrittr")
-library("sf")
-library("raster")
-library("ggplot2")
-library("lattice")
-library("data.table")
-library("RCurl")
-
-# read in the individual sheets with specified range
-con = dbConnect(RSQLite::SQLite(), "../../../../data/tables.gpkg")
-dbListTables(con)
-precip = RSQLite::dbReadTable(con, "precipitation")
-dbDisconnect(con)
-
-#**********************************************************
-# 3 PREC BARPLOT-------------------------------------------
-#**********************************************************
-
-# remove years 1991-1996, they only exist for piura 
-# precip %<>% filter(!year %in% 1991:1996)
-
-# 3.1 Download and prepare ICEN index======================
-#**********************************************************
-
-# ICEN phases
-# ic = readr::read_table("http://www.met.igp.gob.pe/datos/icen.txt", skip = 11)
-# names(ic) = c("year", "month", "icen")
-# ic = mutate(ic, month = as.numeric(as.character(month)))
-# saveRDS(ic, "images/00_icen_index.rds")
-ic = readRDS("../../../../images/00_icen_index.rds")
-ic$enso = cut(ic$icen, breaks = c(-10, -0.99, 0.41, 10),
-              labels = c("LN", "neutral", "EN"))
-# now aggregate to years!!
-# running mean over three months, hence 1 corresponds to djf and 2 to jfm
-tab_2 = filter(ic, month %in% c(1, 2))
-tab_2 = reshape2::dcast(tab_2, year ~ month)
-tab_2 = as.data.table(tab_2)
-names(tab_2) = c("year", "djf", "jfm")
[…]

### 2020-08-12 code/misc/input_data/01_accdb_tables.R

@@ -1,187 +1,0 @@
-# Filename: 00_rodbc.R (2019-09-05)
-
-# TO DO: Read in raw data from accdb, i.e., only possible under Windows
-
-# Author(s): Jannes Muenchow
-
-#**********************************************************
-# CONTENTS-------------------------------------------------
-#**********************************************************
-
-# 1. ATTACH PACKAGES AND DATA
-# 2. PLOT-SPECIES MATRICES (2011, 2012, 2016, 2017)
-# 3. ENVIRONMENTAL DATA
-# 4. IRRIGATION-NUTRIENT EXPERIMENT
-# 5. SAVE TO GPKG
-
-#**********************************************************
-# 1 ATTACH PACKAGES AND DATA-------------------------------
-#**********************************************************
-
-# attach packages
-library("RODBC")
-library("RSQLite")
-library("dplyr")
-library("data.table")
-library("magrittr")
-library("lubridate")
-library("readxl")
-library("reshape2")
-
-# little helper function because we have to do the same thing 4 times
-adjust_tab = function(d, cover = "Deckung_%") {
-  d %>% 
-    # if Arbeitsname does not exist, this only throws a warning message
-    dplyr::select(-one_of("Arbeitsname")) %>%
-    dplyr::rename(art = "Artkürzel", cover = cover) %>%
-    left_join(., lf, by = "art") %>%
-    # arrange by plot id (plot number)
-    arrange(pnr)
-}
-
-#**********************************************************
-# 2 PLOT-SPECIES MATRICES (2011, 2012, 2016, 2017)---------
-#**********************************************************
-
-# NOTE:
-# use 32-bit version of R/Microsoft Office and make sure that you have installed
-# Microsoft Database Engine Distributable (32-bit)
-# browseURL("https://www.microsoft.com/en-us/download/details.aspx?id=13255")
-# remember also that RStudio no longer supports 32-bit R versions, so you have 
-# to run RODBC code in the R console directly
-
-# life form data (lf)
-con = odbcConnectAccess2007("data/raw_data/Transekt_0316.accdb")
-lf = sqlFetch(con, "qurSpecOrigin", as.is = TRUE)
-close(con)
-lf = dplyr::select(lf, art = "Artkürzel", type = "Life form") %>%
[…]

### 2020-08-12 code/misc/input_data/02_vector_tables.R

@@ -1,103 +1,0 @@
-# Filename: vector.R (2018-07-03)
-#
-# TO DO: putting several needed vector data into gpkg
-#
-# Author(s): Jannes Muenchow
-#
-#**********************************************************
-# CONTENTS-------------------------------------------------
-#**********************************************************
-#
-# 1. ATTACH PACKAGES AND DATA
-# 2. GPKG
-# 3. NOMINATIM CITY BOUNDARIES
-#
-#**********************************************************
-# 1 ATTACH PACKAGES AND DATA-------------------------------
-#**********************************************************
-
-# attach packages
-library("sf")
-library("raster")
-library("RSQLite")
-# define directory
-old_dir = file.path("~/work/acad/science/projects/ecology/latin_america/peru/", 
-                    "transect_piura/old/data/R/R data/shapes")
-# read in data
-towns = st_read(file.path(old_dir, "towns.shp"))
-plot(towns$geometry)  # these were the town outlines used in the Erdkunde paper
-towns_2 = st_read(file.path(old_dir, "towns_hannes.shp"))
-plot(towns_2$geometry, add = TRUE)  # and these were the ones used in the JVS paper
-streets = st_read(file.path(old_dir, "streets.shp"))
-arable = st_read(file.path(old_dir, "arable_land.shp"))
-mountains = st_read(file.path(old_dir, "mountains.shp"))
-sa = st_read(file.path(old_dir, "study_area.shp"))
-coast = st_read(file.path(old_dir, "coastline.shp"))
-# water lines were downloaded from DIVA GIS
-rivers = st_read(file.path(old_dir, "PER_wat/PER_water_lines_dcw.shp"))
-rivers = st_transform(rivers, st_crs(sa))
-# neighboring countries
-tmp = ccodes()
-dplyr::filter(tmp,
-              NAME %in% c("Peru", "Ecuador", "Brazil", "Colombia", "Bolivia", 
-                          "Chile"))
-nbs = lapply(c("BOL", "BRA", "CHL", "COL", "ECU"), function(i) {
-  getData("GADM", country = i, level = 0) %>%
-    st_as_sf
-  })
-nbs = do.call(rbind, nbs)
-plot(nbs$geometry, col = NA)
-per = getData("GADM", country = "PER", level = 1) %>% 
-  st_as_sf
-
-#**********************************************************
-# 2 GPKG---------------------------------------------------
[…]

### 2020-08-12 code/misc/inset_map.R

@@ -1,225 +1,0 @@
-# Filename: study_area.R (2019-06-03)
-#
-# TO DO: Overview figure (South America and prediction map)
-#
-# Author(s): Jannes Muenchow
-#
-#**********************************************************
-# CONTENTS-------------------------------------------------
-#**********************************************************
-#
-# 1. ATTACH PACKAGES AND DATA
-# 2. SOUTH AMERICA STUDY AREA
-# 3. LOCAL PREDICTION MAP
-# 4. GRID FIGURE
-#
-#**********************************************************
-# 1 ATTACH PACKAGES AND DATA-------------------------------
-#**********************************************************
-
-# attach packages
-library("sf")
-library("sp")
-library("raster")
-library("dplyr")
-library("latticeExtra")
-library("grid")
-library("gridExtra")
-library("RSQLite")
-
-# # attach data
-cous = c("PER", "ECU", "BOL", "COL", "BRA", "CHL")
-cous = lapply(cous, function(x) {
-  shp = raster::getData(cou = x, level = 0) %>%
-    st_as_sf
-})
-cous = do.call(rbind, cous)
-b_box = st_bbox(cous[1:2, ])
-b_box["xmin"] = st_bbox(cous[1, ])$xmin
-# add something, otherwise Bolivia will become a gc which cannot be converted
-# into a Spatial object
-b_box["xmax"] = st_bbox(cous[1, ])$xmax + 0.5
-sa = st_crop(cous, b_box)
-# check
-sa = as(sa, "Spatial")  # ok, it works
-# saveRDS(sa, file = "images/sa.rds")
-
-# con = dbConnect(SQLite(), "data/tables.gpkg")
-# dbListTables(con)
-# peru = st_read(con, "peru")
-# nbs = st_read(con, "neighbors")
-# 
-# sa = rbind(nbs, select(peru, GID_0, NAME_0))
-# sa = as(sa, "Spatial")  
-
-#*******************************************************
-# 2 SA inset map----------------------------------------
-#*******************************************************
-
-poly = st_bbox(c(xmin = -81, xmax = -79, ymax = -4.6, ymin = -5.4), 
-               crs = st_crs(4326)) %>% 
-  st_as_sfc %>% 
-  as(., "Spatial")
-
-# Base plot version
-
-# however, you cannot mix base plots and lattice/ggplots with grid
-
-# pin = par("pin")
-# # calculate the x- and y-extent
[…]

### 2020-08-12 code/misc/jonas_desc_analysis.R

@@ -1,364 +1,0 @@
-#**********************************************************
-# JONAS STUFF - CAN BE DELETED AFTER TALKING TO ALEX-------
-#**********************************************************
-
-#************************************************
-# CALCULATE MEAN NUMBER OF PLANTS PER PLOT ----
-#************************************************
-# mean number of plants per plot
-dat11 = base::subset(d11, select = "pnr")
-dat12 = base::subset(d12, select = "pnr")
-dat16 = base::subset(d16, select = "pnr")
-dat17 = base::subset(d17, select = "pnr")
-
-# calculate sum of respective plots
-dat11 = dat11 %>%
-  summarise(no_rows = length(pnr))
-
-dat12 = dat12 %>%
-  summarise(no_rows = length(pnr))
-
-dat16 = dat16 %>%
-  summarise(no_rows = length(pnr))
-
-dat17 = dat17 %>%
-  summarise(no_rows = length(pnr))
-
-# preparation for calculating the mean
-plant_mean = cbind(dat11, dat12, dat16, dat17)
-plant_mean = plant_mean[-c(3, 5, 7)]
-colnames(plant_mean)[2] = "2011"
-colnames(plant_mean)[3] = "2012"
-colnames(plant_mean)[4] = "2016"
-colnames(plant_mean)[5] = "2017"
-
-# calculate the mean per plot
-plant_mean$mean = round((plant_mean$`2011` + plant_mean$`2012` +
-  plant_mean$`2016` + plant_mean$`2017`)/4, 0)
-
-
-# define plot style
-plot_syle = theme_bw() + theme(panel.border = element_rect(colour = "black", size=1.4),
-                                axis.line = element_line(colour = "black", size = 1.4),
-                                axis.text.x = element_text(colour = "black"),
-                                axis.text.y = element_text(colour = "black"),
-                                axis.title=element_text(size=14,face="bold"))
-
-# setting the style of the plot (without grid!) -> alternative style
-# plot_syle = theme_bw() + theme(panel.border = element_rect(colour = "black", size=1.4),
[…]

### 2020-08-12 code/misc/latexdiff.sh

@@ -1,2 +1,0 @@
-latexdiff text/ecography/round_2/enso_comp.tex text/ecography/round_3/enso_comp.tex > text/ecography/round_3/diff_ms.tex
-latexdiff text/ecography/round_2/title_page.tex text/ecography/round_3/title_page.tex > text/ecography/round_3/diff_tp.tex

### 2020-08-12 code/misc/single_models.R

@@ -1,128 +1,0 @@
-# Filename: single_models (2019-09-03)
-
-# TO DO: Run each year as a single model
-
-# Author(s): Jannes Muenchow
-
-#**********************************************************
-# CONTENTS-------------------------------------------------
-#**********************************************************
-
-# 1. ATTACH PACKAGES AND DATA
-# 2. INDVIDUAL MODELS
-
-#**********************************************************
-# 1 ATTACH PACKAGES AND DATA-------------------------------
-#**********************************************************
-
-# attach packages
-# attach packages
-library("dplyr")
-library("gdata")
-library("lattice")
-library("latticeExtra")
-library("mgcv")
-library("raster")
-library("rasterVis")
-library("gridExtra")
-library("sperrorest")
-library("sf")
-
-# source your own funs
-source("code/funs/spcv_helper_funs.R")
-
-# attach raster stack
-ndvi = readRDS("images/14_ndvi_rasters.rds")
-# attach response (DCA scores)
-data = readRDS("images/23_dca_scores.rds")
-data$id = as.numeric(data$id)
-# attach env_data
-env_data = readRDS("images/12_env_data.rds")
-env_data$year = gsub(".*_", "", env_data$pnr)
-# attaching spatial data
-rivers = st_read("data/spatial/vector.gpkg", "rivers")
-towns = st_read("data/spatial/vector.gpkg", "towns_jvs")
-
-#**********************************************************
-# 1 INDIVIDUAL MODELS--------------------------------------
-#**********************************************************
-
-# just for exploratory purposes, not used in the paper!
-# add xy-coordinates to dca scores
-data = inner_join(data, 
-                  dplyr::select(env_data, id, year, x, y, lon, lat, slope, 
-                                vcurv),
-                  by = c("id", "year"))
-data$ndvi = NA
-for (i in c(11, 12, 16, 17)) {
-  ind = grepl(i, data[, "year"]) 
-  data[ind, "ndvi"] = 
-    raster::extract(ndvi[[paste0("ndvi_", i)]], 
-                    data[ind, c("lon", "lat")])
-}
-sum(is.na(data$ndvi))  # 0, very good
-
-
-# have a look at the relationship between DCA1 and NDVI again
-
-# make a plot
[…]

### 2020-08-12 code/12_study_area_ndvi.R
rename towns_jvs to towns

@@ -50,7 +50,7 @@ streets = st_read(con, "streets") %>%
 street = streets[1, ]
 
 ## get town polygons
-towns = st_read(con, "towns_jvs") %>%
+towns = st_read(con, "towns") %>%
   st_transform(crs = st_crs(ndvi_rasters)) %>%
   st_crop(st_bbox(ndvi_rasters))
 # get center of towns-polygon

### 2020-08-12 code/15_modeling.R

@@ -47,7 +47,7 @@ env_data$year = gsub(".*_", "", env_data$pnr)
 # attaching spatial data
 con = dbConnect(SQLite(), "data/tables.gpkg")
 rivers = st_read(con, "rivers")
-towns = st_read(con, "towns_jvs")
+towns = st_read(con, "towns")
 dbDisconnect(con)
 
 #**********************************************************

### 2020-08-12 R/download_zenodo.R
add download_zenodo()

@@ -1,0 +1,16 @@
+#' @title Download data from Zenodo
+#' @description
+#' `download_zenodo()`: Download tables.gpkg from Zenodo.
+#' @importFrom curl curl_download
+#' @importFrom sf st_read
+#' @importFrom glue glue
+#'
+#' @template url
+#' @name download
+download_zenodo = function(url) {
+  if (!file.exists("data/tables.gpkg")) {
+    curl_download(url,
+                  destfile = "data/tables.gpkg", 
+                  quiet = FALSE)
+  }
+}

### 2020-08-29 docs/enso_comp.Rmd
refine caption of table 1

@@
 cap = paste("Different biodiversity measures across the studied years.",
-            "min = minimum observed number of species per plot.",
-            "max = maximum observed number of species per plot.")
+            "min = minimum observed number of species over all plots.",
+            "max = maximum observed number of species over all plots.")
 kable(div, "latex", booktabs = TRUE, escape = TRUE, 

### 2020-08-29 docs/enso_comp.Rmd
delete superfluous the

@@
 Plant cover, and thus biomass production, stopped increasing or even decreased in the fertilized Super-El Niño plots after 8 March 2013, which corresponded to 1487 mm of irrigation and 200 kg ha^-1^ of NH~4~NO~3~.
-Species richness followed a similar pattern -- highest numbers in species were recorded in the the fertilized Super-El Niño plots, though with just seven observed species the overall number remained low (Fig.
+Species richness followed a similar pattern -- highest numbers in species were recorded in the fertilized Super-El Niño plots, though with just seven observed species the overall number remained low (Fig.
 \@ref(fig:spri-exp)).

### 2020-09-24 README.Rmd
Rmd README and using here() in first script

# ENSO

This repository contains the code and data for the article

> Muenchow, J., Dieker, P., Böttcher, T., Brock, J., Didenko, G., Fremout, T., Jakubka, D., Jentsch, A., Nüst, D., Richter, M., Rodríguez, E.F., Rodríguez, R.A., Rollenbeck, R., Salazar Zarsosa, P., Schratz, P. and Brenning, A. (2020), **Monitoring and predictive mapping of floristic biodiversity along a climatic gradient in ENSO's terrestrial core region, NW Peru**. Ecography. doi: [10.1111/ecog.05091](https://doi.org/10.1111/ecog.05091)

## Reproduce locally

To install all required packages for this analysis, please execute

```{r environment, eval = FALSE}
renv::init()
renv::install()
```

from the repository root.
The _renv_ package is the successor of the _packrat_ package and can be [installed from CRAN](https://cran.r-project.org/package=renv).
This will create a project-based R library in a subdirectory `renv` based on the lockfile `renv.lock`, with R packages being kept at a specific version for better reproducibility.

## Contents

This repository contains the data, code, and text for the article and realises a [research compendium](https://research-compendium.science/).

### Text

The main body of the article can be found in `docs/enso_comp.Rmd`.
This [R Markdown](https://rmarkdown.rstudio.com/) source be rendered to a PDF document for easier reading, printing, etc.
The file `docs/supplementary_information.Rmd` contains Appendix 1-4 of the article.
The file `docs/title_page.Rmd` contains the authors, article metadata, and abstract.

### Data

Zenodo deposit

[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.3981436.svg)](https://doi.org/10.5281/zenodo.3981436)

### Code

The directory `code` contains script files for all analysis steps.

The directory `R` contains helper functions for the analysis and the visualisations.

### 2020-09-24 code/00_modis_download.R

@@ -16,7 +16,7 @@
 # 1 ATTACH PACKAGES AND (INSTALLATION) SET UP--------------
 #**********************************************************
 
-# create an Earthdata account:
+# create an Earthdata account and store username and password as described in the file README.Rmd:
 # https://urs.earthdata.nasa.gov/home
 
 # optional: download MODIS Reprojection Tool (MRT) from
@@ -27,6 +27,8 @@ library("dplyr")
 library("sf")
 library("raster")
 library("RSQLite")
+library("here")
+devtools::load_all()
 
 # Note: Download GDAL from http://www.gisinternals.com/release.php
 # specify path to GDAL and paths
@@ -37,14 +39,15 @@ MODISoptions(
   # gdalPath = "C:/Program Files/GDAL",
   # gdalPath = "C:/OSGeo4W64/bin",
   # where to save downloaded MODIS data
-  localArcPath = "data/raw_data/modis",
+  localArcPath = here::here("data/raw_data/modis"),
   # where to save processed (projected, cropped and resampled) data
-  outDirPath = "data/raw_data/modis/processed",
+  outDirPath = here::here("data/processed/modis"),
   checkTools = TRUE,
-  MODISserverOrder = c("LPDAAC", "LAADS"),
-  dlmethod = "auto")
-# save options for later on
-opts = MODISoptions()
+  MODISserverOrder = c("LAADS"), #c("LPDAAC", "LAADS"),
+  dlmethod = "auto",
+  save = FALSE,
+  ask = FALSE
+)
 
 # optional:
 # assuming the MRT tools are installed in C:/modis_mrt
@@ -58,31 +61,34 @@ opts = MODISoptions()
 # MODIS:::checkTools('MRT')
 
 # attach data
-con = dbConnect(SQLite(), "data/tables.gpkg")
+download_zenodo()
+con = dbConnect(SQLite(), here("data/tables.gpkg"))
 dbListTables(con)
 sa = sf::read_sf(con, "study_area") %>%
   st_transform(crs = 4326)
+dbDisconnect(con)
 
 #**********************************************************
 # 2 DOWNLOAD AND PREPROCESS MODIS--------------------------
 #**********************************************************
 
 # create file with Earthdata login credentials
-lpdaacLogin(server = "LPDAAC") # -u enso_earthdata -p asdfASDF1234
+file.remove(path.expand("~/.netrc")) # see https://github.com/MatMatt/MODIS/issues/75
[…]
